---
title: "ACE Project discrete"
author: "Caleb"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

Things to note
- 85% binary, 15% 3 state
- 100% fossil preservation too
- only compare nodes present in all trees - keep root, but in living make t1 at time = 0 (effectively becomes outgroup)
- use z-score scaling or whatever it is with scale() function.
- explore with extinction.
- use median rather than mean.
- then look at disparity metrics


## Generate trees

```{r}

library(treats)

bd_params <- make.bd.params(speciation = 1, extinction = 0.7)

stop_rule <- list(max.living = 50)
set.seed(123)
trees <- treats(stop.rule = stop_rule, bd.params = bd_params, null.error = 100, replicates = 100)

## write the saved trees
write.tree(trees, "../Data/trees/simulated.tre")
# write.tree(tree, file = "../Data/processed/trees/simulated.tre")
# tree <- read.tree("../Data/processed/trees/simulated.tre")

# tree <- read.nexus("../Data/raw/trees/Beck2014.tre")
# tree <- makeNodeLabel(tree)
# tree <- set.root.time(tree)
trees <- read.tree("../Data/trees/simulated.tre")

```


## Discrete trait simulation

Firstly lets simulate completely random matrix.

```{r}

tree_labels <- lapply(trees, function(tree) {
  c(tree$tip.label, tree$node.label)
})

# Create a matrix with random numbers between 0 and 3
# how do the number of character states influence ancestral states???? say if there are 5 different states, does this influence disparity analyses. could do another matrix where the number of states changes halfway through the tree to see how ace compares??? 
random_discrete <- lapply(tree_labels, function(tree){
                        matrix <- matrix(sample(0:2, length(tree) * 100, replace = TRUE), 
                        nrow = length(tree), 
                        ncol = 100)
                        rownames(matrix) <- tree
                        return(matrix)})





```


Now we can simulate trait data at varying levels of transition rate.

```{r}

slow_binary_transitions <- matrix(c(
    0.95, 0.05,
    0.05, 0.95
), nrow = 2, byrow = TRUE)

slow_multi_transitions <- matrix(c(
    0.95, 0.025, 0.025,
    0.025, 0.95, 0.025,
    0.025, 0.025, 0.95
), nrow = 3, byrow = TRUE)

slow_binary <- treats::make.traits(process = discrete.process, n = 85, process.args = list(transitions = slow_binary_transitions))
slow_multi <- treats::make.traits(process = discrete.process, n = 15, process.args = list(transitions = slow_multi_transitions))


med_binary_transitions <- matrix(c(
  0.7, 0.3,
  0.3, 0.7
), nrow = 2, byrow = TRUE)


med_multi_transitions <- matrix(c(
  0.6, 0.2, 0.2,
  0.3, 0.4, 0.3,
  0.2, 0.3, 0.5
), nrow = 3, byrow = TRUE)

med_binary <- treats::make.traits(process = discrete.process, n = 85, process.args = list(transitions = med_binary_transitions))
med_multi <- treats::make.traits(process = discrete.process, n = 15, process.args = list(transitions = med_multi_transitions))

fast_binary_transitions <- matrix(c(
    1/2, 1/2,
    1/2, 1/2 
), nrow = 2, byrow = TRUE)

fast_multi_transitions <- matrix(c(
    1/3, 1/3, 1/3, 
    1/3, 1/3, 1/3,
    1/3, 1/3, 1/3 
), nrow = 3, byrow = TRUE)

fast_binary <- treats::make.traits(process = discrete.process, n = 85, process.args = list(transitions = fast_binary_transitions))
fast_multi <- treats::make.traits(process = discrete.process, n = 15, process.args = list(transitions = fast_multi_transitions))



## map the traits

slow_matrices <- lapply(trees, function(tree) {
    mapped_binary <- (map.traits(slow_binary, tree))$data
    mapped_multi <- (map.traits(slow_multi, tree))$data
    matrix <- cbind(mapped_binary, mapped_multi)
    return(matrix)
})

med_matrices <- lapply(trees, function(tree) {
    mapped_binary <- (map.traits(med_binary, tree))$data
    mapped_multi <- (map.traits(med_multi, tree))$data
    matrix <- cbind(mapped_binary, mapped_multi)
    return(matrix)
})

fast_matrices <- lapply(trees, function(tree) {
    mapped_binary <- (map.traits(fast_binary, tree))$data
    mapped_multi <- (map.traits(fast_multi, tree))$data
    matrix <- cbind(mapped_binary, mapped_multi)
    return(matrix)
})


discrete_matrices <- list(slow = slow_matrices, med = med_matrices, fast = fast_matrices)
saveRDS(discrete_matrices, "../Data/matrices/discretematrices.rds")
discrete_matrices <- readRDS("../Data/matrices/discretematrices.rds")

```

## Different tree coverages

```{r}

## fossil degredation as well

source("../Functions/fossil.pres.R")

living <- lapply(discrete_matrices, remove.fossil, trees = trees, type = "discrete")
fossilised_high <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.5, type = "discrete")
all_fossil <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 1.0, type = "discrete")
fossilised_med <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.15, type = "discrete")
fossilised_low <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.05, type = "discrete")

# plot(fossilised_low$slow[[13]]$tree)

fossil_levels <- list(all = all_fossil, fossil_high = fossilised_high, fossil_med = fossilised_med, fossil_low = fossilised_low, living = living)


fossil_matrices <- lapply(names(discrete_matrices), function(level) {
  lapply(names(fossil_levels), function(fossil_type) {
    fossil_levels[[fossil_type]][[level]]
  })
})

names(fossil_matrices) <- names(discrete_matrices)
for (level in names(fossil_matrices)) {
  names(fossil_matrices[[level]]) <- names(fossil_levels)
}

str(fossil_matrices$slow$all[[56]])

```


## discrete pipeline

```{R}
# source("../Functions/apply.dist.R")
# source("../Functions/apply.ordination.R")

# distances <- lapply(discrete_matrices, dist.func)
# traitspaces <- ord.func(distances)

# disp_obj <- lapply(traitspaces, make.dispRity, tree = tree)

# do.time.slice <- function(matrices, chrono_models, FADLAD = NULL, time_interval) {
#   time_slices <- lapply(chrono_models, function(model) {
#     chrono.subsets(matrices, method = "continuous", FADLAD = FADLAD, model = model, time = time_interval)
#   })
#   names(time_slices) <- chrono_models
#   return(time_slices)
# }

# chrono_models <- c("acctran", "deltran", "random", "proximity", "equal.split", "gradual.split")


# time_sliced_discrete <- lapply(disp_obj, do.time.slice, chrono_models = chrono_models, time_interval = 10)

# bootstrapped_discrete <- lapply(time_sliced_discrete, lapply, boot.matrix, bootstraps = 100)

# disparity_discrete <- lapply(bootstrapped_discrete, lapply,  dispRity, metric = c(sum, variances), na.rm = TRUE, verbose = TRUE)

# saveRDS(disparity_discrete, "../Data/processed/disparity/aceproject/beckdiscrete.Rds")
# saveRDS(disparity_discrete, "../Data/processed/disparity/aceproject/simdiscrete.Rds")


# plot(disparity_discrete$low$gradual.split)

```


## Ace discrete

```{r}

anc.states <- function(fossil_matrices) {
  lapply(fossil_matrices, function(level) {
    lapply(level, function(x) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = x$matrix, 
                          tree = x$tree, 
                          models = "ER", 
                          output = "combined",
                          verbose = TRUE)
  return(anc_states)})})
  }

fossil_anc <- lapply(fossil_matrices, anc.states)

saveRDS(fossil_anc, "../Data/matrices/discreteanc.Rds")
fossil_anc <- readRDS("../Data/matrices/discreteanc.Rds")



```




## Compare true vs ace nodes by distance

```{r}
## this function returns matrices of ace and true node states
get.living.nodes <- function(true, ace) {
  node_v <- rownames(ace)[grepl("^n", rownames(ace))]
  ace_filtered <- ace[node_v, ]
  true_filtered <- true[node_v, ]
  return(list(ace = ace_filtered, true = true_filtered))
}

filter.nodes <- function(true, ace, node_v) {
  ace_filtered <- ace[node_v, ]
  true_filtered <- true[node_v, ]
  return(list(ace = ace_filtered, true = true_filtered))
}


# Iterate over each level (slow, med, fast) and extract nodes for $living
nodes_living <- lapply(names(discrete_matrices), function(level) {
  lapply(seq_along(discrete_matrices[[level]]), function(i) {
    # Step 1: Get node names from "living" ACE
    living_node <- get.living.nodes(
      true = discrete_matrices[[level]][[i]],
      ace = fossil_anc[[level]]$living[[i]]
    )
    node_v <- rownames(living_node$ace)

    # Step 2: For the current level, filter all preservation states
    filtered_preservations <- lapply(names(fossil_anc[[level]]), function(pres) {
      filter.nodes(
        true = discrete_matrices[[level]][[i]],
        ace = fossil_anc[[level]][[pres]][[i]],
        node_v = node_v
      )
    })
    names(filtered_preservations) <- names(fossil_anc[[level]])
    return(filtered_preservations)
  })
})
names(nodes_living) <- names(discrete_matrices)

source("../Functions/node.dist.R")


dist_lenient <- lapply(nodes_living, function(preservation) {
lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    distances <- node.dist(matrices, distance = "lenient")
  })
})
})

dist_strict <- lapply(nodes_living, function(preservation) {
lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    distances <- node.dist(matrices, distance = "strict")
  })
})
})


strict_means <- lapply(dist_strict, function(preservation) {
  lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    means <- mean(matrices)
  })
})
})

## do i need to divide by number of rows?

lenient_means <- lapply(dist_lenient, function(preservation) {
  lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    means <- mean(matrices)
  })
})
})

saveRDS(uncertain_means, "../Data/matrices/uncertainmeans.rds")
saveRDS(strict_means, "../Data/matrices/strictmeans.rds")


```



## Comparison of morphospace

```{R}

# true_nodes <- lapply(nodes_living, function(rate){
#   lapply(rate, function(matrix){
#     lapply(matrix, function(preservation){
#       node_labels <- rownames(preservation$ace)
#     })
#   })
# })


# true_nodes <- lapply(true_nodes, function(rate){
#   lapply(rate, function(matrix){
#     matrix$fossil_high
#   })
# })
# true_nodes <- true_nodes$fast




# ## get the true nodes
# bound_true <- 
#   lapply(discrete_matrices, function(rate){
#   Map(function(nodes, matrix){
#   node_characters <- matrix[nodes,]
# }, true_nodes, rate)
# })

# ## bind to true tips
# true_matrices <- lapply(names(living), function(rate) {
#   lapply(seq_along(living[[rate]]), function(i) {
#     rbind(living[[rate]][[i]]$matrix, bound_true[[rate]][[i]])
#   })
# })
# names(true_matrices) <- names(bound_true)

# # get the tips for each matrix
# filtered_fossil_anc <- lapply(names(fossil_anc), function(level) {
#   lapply(names(fossil_anc[[level]]), function(preservation) {
#     lapply(seq_along(fossil_anc[[level]][[preservation]]), function(i) {
#       # Get the current matrix from fossil_anc
#       fossil_matrix <- fossil_anc[[level]][[preservation]][[i]]
      
#       # Get the corresponding true nodes
#       nodes_to_keep <- true_nodes[[i]]
      
#       # Filter the fossil_matrix to only keep rows with matching nodes
#       fossil_matrix[rownames(fossil_matrix) %in% nodes_to_keep, , drop = FALSE]
#     })
#   })
# })

# # Preserve the names of the structure
# names(filtered_fossil_anc) <- names(fossil_anc)
# for (level in names(fossil_anc)) {
#   names(filtered_fossil_anc[[level]]) <- names(fossil_anc[[level]])
# }


distances <- lapply(nodes_living, lapply, lapply, dist.func)


## these are anc estimated matrices with relevant tips but only with relevant nodes.
anc_matrices_living


source("../Functions/apply.dist.R")
source("../Functions/apply.ordination.R")


ord <- lapply(distances, lapply, lapply, ord.func)

scaled_ord <- lapply(ord, function(preservation){
  lapply(preservation, function(rate){
    lapply(rate, function(treatment){
      lapply(treatment, function(matrix){
        scaled_pc <- scale(matrix)
        return(scaled_pc)
      })
    })
  })
})


test <- dispRity::dispRity(ord$slow[[1]]$living$true, metric = c(pairwise.dist))

## Relative distance in trait space

morphospace.dist <- function(matrix, ...) {
    ## Get the pairwise distances between all points
    distances <- pairwise.dist(matrix)
    ## Calculate the longest distance from the center
    longest_distance_from_center <- max(as.numeric(centroids(matrix, centroid = 0)), na.rm = TRUE)
    ## Make all these distances relative to the longest distance from the center of the traitspace
    return(distances/longest_distance_from_center)
}



morpho_diffs <- lapply(scaled_ord, lapply, lapply, function(matrices){
  diffs <- abs((morphospace.dist(matrices$ace)) - (morphospace.dist(matrices$true)))
  averages <- mean(diffs)
})


sum_var <- lapply(ord, lapply, lapply, lapply, function(matrix){
    disparity <- (dispRity(matrix, metric = c(sum, variances)))$disparity})

sum_var <- lapply(ord, lapply, lapply, lapply, function(matrix){
    disparity <- (dispRity(matrix, metric = c(sum, variances)))$disparity})

disparity_diffs <- lapply(sum_var, lapply, lapply, function(matrices){
  diffs <- abs(matrices$ace[[1]]$elements[1] - matrices$true[[1]]$elements[1])
})


## Effect of removing fossils




plot(scaled_ace_ord$slow$fossil_high[[15]])
text(scaled_ace_ord$slow$fossil_high[[15]], labels = rownames(scaled_ace_ord$slow$fossil_high[[15]]), pos = 4, cex = 0.6)


# morphospace.dist <- function(ace_ord) {
#   ace_scores <- ace_ord[grepl("^ace", rownames(ace_ord)), ]
#   true_scores <- ace_ord[!grepl("^ace", rownames(ace_ord)), ]
#   distances <- sqrt(rowSums((ace_scores[,1:15] - true_scores[,1:15])^2))
#   return(distances)
# }

# ace_ord_scores <- ace_ord$fossil_high$low[[1]][grepl("^ace", rownames(ace_ord$fossil_high$low[[1]])), ]
# true_ord_scores <- ace_ord$fossil_high$low[[1]][!grepl("^ace", rownames(ace_ord$fossil_high$low[[1]])), ]
# test <- ace_ord_scores[94,1:10]
# test2 <- true_ord_scores[94,1:10]
# morpho_distances <- sqrt(rowSums((ace_ord_scores[,1:10] - true_ord_scores[,1:10])^2))
# ## tends to converge on exactly the same value
# # sqrt(sum((test - test2)^2))

morpho_dists <- lapply(scaled_ace_ord, lapply, lapply, morphospace.dist)
# morpho_dists_med <- unlist(morpho_dists$slow)
saveRDS(morpho_dists, "../Data/matrices/orddists.Rds")

unlisted_morpho_dists <- lapply(morpho_dists, function(rate) {
    unlist(rate, recursive = TRUE)
})

split_dists <- lapply(unlisted_morpho_dists, function(rate){
  split(rate, gsub("\\d+", "", names(rate)))
})

categories <- c("living", "fossil_high", "fossil_med", "fossil_low")
boxplot(split_dists$med, 
        main = "", 
        ylab = "", 
        xlab = "Categories", 
        las = 2)

## lowering number of pc axes increases high fossil recovery rate. Try this by lowering the number of pc axes included in distance function

mean(morpho_dists$fossil_low$low[[15]])
mean(morpho_dists$fossil_high$low[[15]])

## larger distances in fossil high cos range is larger - perhaps need to scale.
## using first 10 pc axes finds considerably more correct values.


```

## map the corresponding trees and ancestral states

```{R}
disp_obj_anc <- lapply(names(traitspaces_anc), function(level) {
  Map(function(traitspace, fossil_level) {
    make.dispRity(traitspace, fossil_level$tree)
  }, traitspaces_anc[[level]], low_fossil_levels[[level]])
})
names(disp_obj_anc) <- names(traitspaces_anc)



time_sliced_discrete_anc <- lapply(disp_obj_anc, lapply, do.time.slice, chrono_models = "gradual.split", time_interval = 10)

bootstrapped_discrete_anc <- lapply(time_sliced_discrete_anc, lapply, lapply, boot.matrix, bootstraps = 100)

disparity_discrete_anc <- lapply(bootstrapped_discrete_anc, lapply, lapply,  dispRity, metric = c(sum, variances), na.rm = TRUE, verbose = TRUE)

plot(disparity_discrete_anc$fossil_low[[8]]$gradual.split)


##but when running with beck tree, i find that ace recovers the correct disparity signal. this is likely because of the distribution of tips through time
plot(disparity_discrete_anc$wide_random$gradual.split)

saveRDS(disparity_discrete_anc, "../Data/processed/disparity/aceproject/beckdiscreteanc.Rds")

saveRDS(disparity_discrete_anc, "../Data/processed/disparity/aceproject/simdiscreteanc.Rds")


## this is super important, as this means we have likely been inflating disparity of earlier nodes due to ancestral state estimation. and because the fossil record of these earlier nodes is usually very incomplete anyway, it adds a smoothing effect.


```

## ace continuous

```{r}



cont_anc_states <- lapply(tip_matrices_cont, function(matrix) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = matrix, 
                          tree = tree, 
                          models = "ML", 
                          output = "combined")
  
  return(anc_states)  
})

traitspace_cont_anc <- pca.func(cont_anc_states)

disp_obj_cont_anc <- lapply(traitspace_cont_anc, make.dispRity, tree = tree)


time_sliced_cont_anc <- lapply(disp_obj_cont_anc, do.time.slice, chrono_models = chrono_models, time_interval = 10)

bootstrapped_contanc <- lapply(time_sliced_cont_anc, lapply, boot.matrix, bootstraps = 100)

disparity_cont_anc <- lapply(bootstrapped_contanc, lapply,  dispRity, metric = c(sum, variances), na.rm = TRUE, verbose = TRUE)

saveRDS(disparity_cont_anc, "../Data/processed/disparity/aceproject/beckcontanc.Rds")
saveRDS(disparity_cont_anc, "../Data/processed/disparity/aceproject/simcontanc.Rds")



plot(disparity_cont_anc$ou$gradual.split)

```

## Discrete with ace (post-ord)

```{r}

tip_matrices <- lapply(discrete_matrices, function(matrix){
    tip_matrix <- matrix[!grepl("^n", rownames(matrix)), ]
    tip_matrix_characters <- apply(tip_matrix, c(1, 2), as.character)
    return(tip_matrix_characters)
})


distances <- dist.func(tip_matrices)
traitspaces <- ord.func(distances)

post_ord_anc <- lapply(traitspaces, function(matrix) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = matrix, 
                          tree = tree, 
                          models = "ML", 
                          output = "combined")
  
  return(anc_states)  
})


disp_obj_post_anc <- lapply(post_ord_anc, make.dispRity, tree = tree)


time_sliced_discrete_post_anc <- lapply(disp_obj_post_anc, do.time.slice, chrono_models = chrono_models, time_interval = 10)

bootstrapped_discrete_post_anc <- lapply(time_sliced_discrete_post_anc, lapply, boot.matrix, bootstraps = 100)

disparity_discrete_post_anc <- lapply(bootstrapped_discrete_post_anc, lapply,  dispRity, metric = c(sum, variances), na.rm = TRUE, verbose = TRUE)

plot(disparity_discrete_post_anc$low$gradual.split)

saveRDS(disparity_discrete_post_anc, "../Data/processed/disparity/aceproject/simpostancdiscrete.Rds")

## this is super important, as this means we have likely been inflating disparity of earlier nodes due to ancestral state estimation. and because the fossil record of these earlier nodes is usually very incomplete anyway, it adds a smoothing effect.


```


## Discrete with ace (pre-ord & sample)

```{r}

tip_matrices <- lapply(discrete_matrices, function(matrix){
    tip_matrix <- matrix[!grepl("^n", rownames(matrix)), ]
    tip_matrix_characters <- apply(tip_matrix, c(1, 2), as.character)
    return(tip_matrix_characters)
})

normal_sampling <- list(fun = rnorm, param = list(mean = mean, sd = function(x)return(diff(range(x))/4)))


discrete_anc_states_sample <- lapply(tip_matrices, function(matrix) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = matrix, 
                          tree = tree, 
                          models = "ER", 
                          output = "combined",
                          sample = 1000,
                          sample.fun = normal_sampling)
  
  return(anc_states)  
})

distances_anc_sample <- lapply(discrete_anc_states_sample, dist.func)
traitspaces_anc_sample <- lapply(distances_anc_sample, ord.func)

disp_obj_anc_sample <- lapply(traitspaces_anc_sample, lapply, make.dispRity, tree = tree)


time_sliced_discrete_anc_sample <- lapply(disp_obj_anc_sample, lapply, chrono.subsets, method = "continuous", model = "gradual.split", time = 10)

# bootstrapped_discrete_anc_sample <- lapply(time_sliced_discrete_anc_sample, lapply, boot.matrix, bootstraps = 100)

disparity_discrete_anc_sample <- lapply(time_sliced_discrete_anc_sample, lapply,  dispRity, metric = c(sum, variances), na.rm = TRUE, verbose = TRUE)

saveRDS(disparity_discrete_anc_sample, "../Data/processed/disparity/aceproject/simancdiscretesample.Rds")

plot(disparity_discrete_anc_sample$low[[234]])

low <- lapply(disparity_discrete_anc_sample$low, get.disparity)
low <- do.call(cbind, low)



subset_names <- names(disparity_discrete_anc_sample$low[[1]]$disparity)

# Group the 1000 values into 10 subsets
grouped_subsets <- lapply(subset_names, function(subset_name) {
  # Extract the values for the current subset across all 1000 lists
  values <- sapply(low, function(x) x[[subset_name]])
  
  # Calculate the median of the values
  median_value <- median(values, na.rm = TRUE)
  
  # Reshape the values into a single row with 1000 columns
  values_matrix <- matrix(values, nrow = 1, byrow = TRUE)
  
  # Create the desired structure
  list(
    elements = matrix(median_value, nrow = 1),  # Median as a single-row matrix
     values_matrix               # All values as a matrix
  )
})
names(grouped_subsets) <- subset_names


disparity_discrete_anc_sample$low[[1]]$disparity <- grouped_subsets
disparity_discrete_anc_sample$low[[1]]$call$bootstrap <- list(1000, "full", NULL)

plot(disparity_discrete_anc_sample$low[[1]])

print.dispRity(disparity_discrete_anc_sample$low[[1]])


    
## this is super important, as this means we have likely been inflating disparity of earlier nodes due to ancestral state estimation. and because the fossil record of these earlier nodes is usually very incomplete anyway, it adds a smoothing effect.


```


## Testing against claddis

```{r echo = FALSE}
# write.nexus.data(discrete_matrices$low, file = "../Data/raw/matrices/random.nex") 

# claddis_matrix <- read_nexus_matrix("../Data/raw/matrices/Brusatte2014.nex")
# bru_tree <- read.nexus("../Data/raw/trees/Brusatte2014.nex")


# ancestral_states <- estimate_ancestral_states(claddis_matrix,  bru_tree)
```