---
title: "ACE Project discrete"
author: "Caleb"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

Things to note
- 85% binary, 15% 3 state
- 100% fossil preservation too
- only compare nodes present in all trees - keep root, but in living make t1 at time = 0 (effectively becomes outgroup)
- use z-score scaling or whatever it is with scale() function.
- explore with extinction.
- use median rather than mean.
- then look at disparity metrics


## Generate trees

```{r}

library(treats)

bd_params <- make.bd.params(speciation = 1, extinction = 0.7)

stop_rule <- list(max.living = 50)
set.seed(123)
trees <- treats(stop.rule = stop_rule, bd.params = bd_params, null.error = 100, replicates = 100)

## write the saved trees
write.tree(trees, "../Data/trees/simulated.tre")
# write.tree(tree, file = "../Data/processed/trees/simulated.tre")
# tree <- read.tree("../Data/processed/trees/simulated.tre")

# tree <- read.nexus("../Data/raw/trees/Beck2014.tre")
# tree <- makeNodeLabel(tree)
# tree <- set.root.time(tree)
trees <- read.tree("../Data/trees/simulated.tre")

```


## Discrete trait simulation

Firstly lets simulate completely random matrix.

```{r}

tree_labels <- lapply(trees, function(tree) {
  c(tree$tip.label, tree$node.label)
})

# Create a matrix with random numbers between 0 and 3
# how do the number of character states influence ancestral states???? say if there are 5 different states, does this influence disparity analyses. could do another matrix where the number of states changes halfway through the tree to see how ace compares??? 
random_discrete <- lapply(tree_labels, function(tree){
                        matrix <- matrix(sample(0:2, length(tree) * 100, replace = TRUE), 
                        nrow = length(tree), 
                        ncol = 100)
                        rownames(matrix) <- tree
                        return(matrix)})





```


Now we can simulate trait data at varying levels of transition rate.

```{r}

slow_binary_transitions <- matrix(c(
    0.99, 0.01,
    0.01, 0.99
), nrow = 2, byrow = TRUE)

slow_multi_transitions <- matrix(c(
    0.99, 0.025, 0.025,
    0.025, 0.99, 0.025,
    0.025, 0.025, 0.99
), nrow = 3, byrow = TRUE)

slow_binary <- treats::make.traits(process = discrete.process, n = 85, process.args = list(transitions = slow_binary_transitions))
slow_multi <- treats::make.traits(process = discrete.process, n = 15, process.args = list(transitions = slow_multi_transitions))


med_binary_transitions <- matrix(c(
  0.90, 0.1,
  0.1, 0.9
), nrow = 2, byrow = TRUE)


med_multi_transitions <- matrix(c(
  0.9, 0.05, 0.05,
  0.05, 0.9, 0.05,
  0.05, 0.05, 0.9
), nrow = 3, byrow = TRUE)

med_binary <- treats::make.traits(process = discrete.process, n = 85, process.args = list(transitions = med_binary_transitions))
med_multi <- treats::make.traits(process = discrete.process, n = 15, process.args = list(transitions = med_multi_transitions))

fast_binary_transitions <- matrix(c(
    0.1, 0.9,
    0.9, 0.1
), nrow = 2, byrow = TRUE)

fast_multi_transitions <- matrix(c(
    0.1, 0.45, 0.45, 
    0.45, 0.1, 0.45,
    0.45, 0.45, 0.1 
), nrow = 3, byrow = TRUE)

fast_binary <- treats::make.traits(process = discrete.process, n = 85, process.args = list(transitions = fast_binary_transitions))
fast_multi <- treats::make.traits(process = discrete.process, n = 15, process.args = list(transitions = fast_multi_transitions))



## map the traits

slow_matrices <- lapply(trees, function(tree) {
    mapped_binary <- (map.traits(slow_binary, tree))$data
    mapped_multi <- (map.traits(slow_multi, tree))$data
    matrix <- cbind(mapped_binary, mapped_multi)
    return(matrix)
})

med_matrices <- lapply(trees, function(tree) {
    mapped_binary <- (map.traits(med_binary, tree))$data
    mapped_multi <- (map.traits(med_multi, tree))$data
    matrix <- cbind(mapped_binary, mapped_multi)
    return(matrix)
})

fast_matrices <- lapply(trees, function(tree) {
    mapped_binary <- (map.traits(fast_binary, tree))$data
    mapped_multi <- (map.traits(fast_multi, tree))$data
    matrix <- cbind(mapped_binary, mapped_multi)
    return(matrix)
})


discrete_matrices <- list(slow = slow_matrices, med = med_matrices, fast = fast_matrices)
saveRDS(discrete_matrices, "../Data/matrices/discretematrices.rds")
discrete_matrices <- readRDS("../Data/matrices/discretematrices.rds")

```

## Different tree coverages

```{r}

## fossil degredation as well

source("../Functions/fossil.pres.R")

living <- lapply(discrete_matrices, remove.fossil, trees = trees, type = "discrete")
fossilised_high <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.5, type = "discrete")
all_fossil <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 1.0, type = "discrete")
fossilised_med <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.15, type = "discrete")
fossilised_low <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.05, type = "discrete")



# plot(fossilised_low$slow[[13]]$tree)

fossil_matrices <- lapply(names(discrete_matrices), function(level) {
  lapply(seq_along(discrete_matrices[[level]]), function(i) {
    list(
      all = all_fossil[[level]][[i]],
      fossil_high = fossilised_high[[level]][[i]],
      fossil_med = fossilised_med[[level]][[i]],
      fossil_low = fossilised_low[[level]][[i]],
      living = living[[level]][[i]]
    )
  })
})

# Assign names to the outer list
names(fossil_matrices) <- names(discrete_matrices)


```


## discrete pipeline

```{R}
# source("../Functions/apply.dist.R")
# source("../Functions/apply.ordination.R")

# distances <- lapply(discrete_matrices, dist.func)
# traitspaces <- ord.func(distances)

# disp_obj <- lapply(traitspaces, make.dispRity, tree = tree)

# do.time.slice <- function(matrices, chrono_models, FADLAD = NULL, time_interval) {
#   time_slices <- lapply(chrono_models, function(model) {
#     chrono.subsets(matrices, method = "continuous", FADLAD = FADLAD, model = model, time = time_interval)
#   })
#   names(time_slices) <- chrono_models
#   return(time_slices)
# }

# chrono_models <- c("acctran", "deltran", "random", "proximity", "equal.split", "gradual.split")


# time_sliced_discrete <- lapply(disp_obj, do.time.slice, chrono_models = chrono_models, time_interval = 10)

# bootstrapped_discrete <- lapply(time_sliced_discrete, lapply, boot.matrix, bootstraps = 100)

# disparity_discrete <- lapply(bootstrapped_discrete, lapply,  dispRity, metric = c(sum, variances), na.rm = TRUE, verbose = TRUE)

# saveRDS(disparity_discrete, "../Data/processed/disparity/aceproject/beckdiscrete.Rds")
# saveRDS(disparity_discrete, "../Data/processed/disparity/aceproject/simdiscrete.Rds")


# plot(disparity_discrete$low$gradual.split)

```


## Ace discrete

```{r}

anc.states <- function(fossil_matrices) {
  lapply(fossil_matrices, function(level) {
    lapply(level, function(x) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = x$matrix, 
                          tree = x$tree, 
                          models = "SYM", 
                          output = "multi.ace",
                          verbose = TRUE)
  return(anc_states)})})
  }

point.anc.states <- function(fossil_matrices) {
  lapply(fossil_matrices, function(level) {
    lapply(level, function(x) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = x$matrix, 
                          tree = x$tree, 
                          models = "SYM", 
                          output = "combined",
                          verbose = TRUE)
  return(anc_states)})})
  }
normal_sampling <- list(fun = rnorm, param = list(mean = mean, sd = function(x)return(diff(range(x))/4)))

sample.anc.states <- function(fossil_matrices) {
  lapply(fossil_matrices, function(level) {
    lapply(level, function(x) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = x,
                          models = "SYM",
                          sample = 5,
                          sample.fun = normal_sampling,
                          output = "combined",
                          verbose = TRUE)
  return(anc_states)})})
  }

fossil_anc <- lapply(fossil_matrices, anc.states)

# sampled_fossil_anc <- lapply(fossil_anc, sample.anc.states)

sampled_fossil_anc <- lapply(fossil_anc, lapply, lapply, multi.ace, sample = 10, sample.fun = normal_sampling,  verbose = TRUE)




point_fossil_anc <- lapply(fossil_anc, lapply, lapply, multi.ace, output = "combined.matrix", verbose = TRUE)

fossil_anc <- point_fossil_anc
saveRDS(fossil_anc, "../Data/matrices/discreteanc.rds")
saveRDS(sampled_fossil_anc, "../Data/matrices/sampleddiscreteanc.Rds")
saveRDS(point_fossil_anc, "../Data/matrices/pointddiscreteanc.Rds")


fossil_anc <- readRDS("../Data/matrices/discreteanc.rds")

sampled_fossil_anc <- readRDS("../Data/matrices/sampleddiscreteanc.Rds")

slow <- sampled_fossil_anc$slow

```




## Compare true vs ace nodes by distance

```{r}
## this function returns matrices of ace and true node states
get.living.nodes <- function(living) {
  if (inherits(living$living, "list")) {
  matrix <- living$living[[1]]
  } else {
  matrix <- living$living}
  node_v <- rownames(matrix)[grepl("^n", rownames(matrix))]
  return(node_v)
}


node_labels <- lapply(sampled_fossil_anc$slow, get.living.nodes)

filter.nodes <- function(true, ace, node_v) {
  ace_filtered <- ace[node_v, , drop = FALSE]
  true_filtered <- true[node_v, , drop = FALSE]
  return(list(ace = ace_filtered, true = true_filtered))
}

# nodes_living <- function(true_matrices, aced_matrices) {
#   output <- lapply(names(true_matrices), function(i) {

#     # Step 1: Get node names from "living" ACE
#     living_node <- get.living.nodes(
#       living = aced_matrices$living[[i]]
#     )

#     # Step 2: For the current level, filter all preservation states
#     filtered_preservations <- lapply(names(aced_matrices), function(pres) {
#       lapply(seq_along(aced_matrices[[pres]][[i]]), function(j) {
#         filter.nodes(
#           true = true_matrices[[i]],
#           ace = aced_matrices[[pres]][[i]][[j]],
#           node_v = living_node
#         )
#       })
#     })
#     names(filtered_preservations) <- names(aced_matrices)
#     return(filtered_preservations)
#   })

#   names(output) <- names(true_matrices)
#   return(output)
# }


sample_living_nodes <- lapply(seq_along(sampled_fossil_anc), function(rate) {
  lapply(seq_along(sampled_fossil_anc[[rate]]), function(i) {
    node_v <- node_labels[[i]]
    levels <- names(sampled_fossil_anc[[rate]][[i]])
    
    level_out <- lapply(seq_along(levels), function(level_idx) {
      level_name <- levels[[level_idx]]
      reps <- sampled_fossil_anc[[rate]][[i]][[level_name]]
      
      lapply(seq_along(reps), function(rep) {
        filter.nodes(
          true = discrete_matrices[[rate]][[i]],
          ace = reps[[rep]],
          node_v = node_v
        )
      })
    })
    names(level_out) <- levels
    return(level_out)
  })
})
names(sample_living_nodes) <- names(discrete_matrices)


point_living_nodes <- lapply(seq_along(point_fossil_anc), function(rate) {
  lapply(seq_along(point_fossil_anc[[rate]]), function(i) {
    node_v <- node_labels[[i]]
    levels <- names(point_fossil_anc[[rate]][[i]])
    
    level_out <- lapply(seq_along(levels), function(level_idx) {
      filter.nodes(
        true = discrete_matrices[[rate]][[i]],
        ace = point_fossil_anc[[rate]][[i]][[level_idx]],
        node_v = node_v
      )
    })
    
    names(level_out) <- levels
    return(level_out)
  })
})


names(point_living_nodes) <- names(discrete_matrices)








# living_nodes <- lapply(point$slow, get.living.nodes)


# nodes_living <- lapply(names(discrete_matrices), function(level) {
#   lapply(seq_along(discrete_matrices[[level]]), function(i) {
#     # Step 1: Get node names from "living" ACE
#     living_node <- get.living.nodes(
#       living = point_fossil_anc[[level]]$living[[i]]
#     )

#     # Step 2: For the current level, filter all preservation states
#     filtered_preservations <- lapply(names(fossil_anc[[level]]), function(pres) {
#       filter.nodes(
#         true = discrete_matrices[[level]][[i]],
#         ace = point_fossil_anc[[level]][[pres]][[i]],
#         node_v = living_node
#       )
#     })
#     names(filtered_preservations) <- names(point_fossil_anc[[level]])
#     return(filtered_preservations)
#   })
# })
# names(nodes_living) <- names(discrete_matrices)






# slow <- sampled_fossil_anc$slow

# med_filt <- lapply(sampled_fossil_anc$med, function(i) {
#   lapply(seq_along(i), function(preservation) {
#     # Step 1: Filter the true matrix once for the current replicate (i)
#     true_filtered <- discrete_matrices$med[[i]][living_nodes[[i]], ]

#     # Step 2: Filter all ACE matrices (j levels) for the current replicate (i)
#     ace_filtered_list <- lapply(seq_along(i[[preservation]]), function(j) {
#       ace_filtered <- i[[preservation]][[j]][living_nodes[[i]], ]
#       return(ace_filtered)
#     })

#     # Combine the true filtered matrix and the list of ACE filtered matrices
#     return(list(true = true_filtered, ace = ace_filtered_list))
#   })
# })

# med_filt <- lapply(sampled_fossil_anc$med, function(preservation_level) {
#   lapply(seq_along(preservation_level), function(i) {
#     # Step 1: Filter the true matrix once for the current replicate (i)
#     true_filtered <- discrete_matrices$med[[i]][living_nodes[[i]], ]

#     # Step 2: Filter all ACE matrices (j levels) for the current replicate (i)
#     ace_filtered_list <- lapply(seq_along(preservation_level[[i]]), function(j) {
#       ace_filtered <- preservation_level[[i]][[j]][living_nodes[[i]], ]
#       return(ace_filtered)
#     })

#     # Combine the true filtered matrix and the list of ACE filtered matrices
#     return(list(true = true_filtered, ace = ace_filtered_list))
#   })
# })

# fast_filt <- lapply(sampled_fossil_anc$fast, function(preservation_level) {
#   lapply(seq_along(preservation_level), function(i) {
#     # Step 1: Filter the true matrix once for the current replicate (i)
#     true_filtered <- discrete_matrices$fast[[i]][living_nodes[[i]], ]

#     # Step 2: Filter all ACE matrices (j levels) for the current replicate (i)
#     ace_filtered_list <- lapply(seq_along(preservation_level[[i]]), function(j) {
#       ace_filtered <- preservation_level[[i]][[j]][living_nodes[[i]], ]
#       return(ace_filtered)
#     })

#     # Combine the true filtered matrix and the list of ACE filtered matrices
#     return(list(true = true_filtered, ace = ace_filtered_list))
#   })
# })



# nodes_living_sample <- list(slow = slow_filt, med = med_filt, fast = fast_filt)


source("../Functions/node.dist.R")


dist_lenient <- lapply(nodes_living, function(preservation) {
lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    distances <- node.dist(matrices, distance = "lenient")
  })
})
})

dist_strict <- lapply(point_living_nodes, function(preservation) {
lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    distances <- node.dist(matrices, distance = "strict")
  })
})
})


dist_strict_sample <- lapply(sample_living_nodes, function(rate) {
lapply(rate, function(matrix) {
  lapply(matrix, function(pres) {
    distances <- node.dist(pres, distance = "strict")
  })
})
})

strict_means_sample <- lapply(dist_strict_sample, function(preservation) {
  lapply(preservation, function(level) {
    lapply(level, function(matrices) {
      # Compute the mean of each replicate
      replicate_means <- sapply(matrices, mean)
      
      # Compute the mean of the replicate means
      overall_mean <- mean(replicate_means)
      
      return(overall_mean)
    })
  })
})


strict_means <- lapply(dist_strict, function(preservation) {
  lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    means <- mean(matrices)
  })
})
})

## do i need to divide by number of rows?

lenient_means <- lapply(dist_lenient, function(preservation) {
  lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    means <- mean(matrices)
  })
})
})

saveRDS(lenient_means, "../Data/matrices/lenientmeans.rds")
saveRDS(strict_means, "../Data/matrices/strictmeans.rds")
saveRDS(strict_means_sample, "../Data/matrices/samplestrictmeans.rds")


```



## Comparison of morphospace

```{R}

# true_nodes <- lapply(nodes_living, function(rate){
#   lapply(rate, function(matrix){
#     lapply(matrix, function(preservation){
#       node_labels <- rownames(preservation$ace)
#     })
#   })
# })


# true_nodes <- lapply(true_nodes, function(rate){
#   lapply(rate, function(matrix){
#     matrix$fossil_high
#   })
# })
# true_nodes <- true_nodes$fast




# ## get the true nodes
# bound_true <- 
#   lapply(discrete_matrices, function(rate){
#   Map(function(nodes, matrix){
#   node_characters <- matrix[nodes,]
# }, true_nodes, rate)
# })

# ## bind to true tips
# true_matrices <- lapply(names(living), function(rate) {
#   lapply(seq_along(living[[rate]]), function(i) {
#     rbind(living[[rate]][[i]]$matrix, bound_true[[rate]][[i]])
#   })
# })
# names(true_matrices) <- names(bound_true)

# # get the tips for each matrix
# filtered_fossil_anc <- lapply(names(fossil_anc), function(level) {
#   lapply(names(fossil_anc[[level]]), function(preservation) {
#     lapply(seq_along(fossil_anc[[level]][[preservation]]), function(i) {
#       # Get the current matrix from fossil_anc
#       fossil_matrix <- fossil_anc[[level]][[preservation]][[i]]
      
#       # Get the corresponding true nodes
#       nodes_to_keep <- true_nodes[[i]]
      
#       # Filter the fossil_matrix to only keep rows with matching nodes
#       fossil_matrix[rownames(fossil_matrix) %in% nodes_to_keep, , drop = FALSE]
#     })
#   })
# })

# # Preserve the names of the structure
# names(filtered_fossil_anc) <- names(fossil_anc)
# for (level in names(fossil_anc)) {
#   names(filtered_fossil_anc[[level]]) <- names(fossil_anc[[level]])
# }




## these are anc estimated matrices with relevant tips but only with relevant nodes.
# anc_matrices_living


source("../Functions/apply.dist.R")
source("../Functions/apply.ordination.R")

distances <- lapply(nodes_living, lapply, lapply, dist.func)

ord <- lapply(distances, lapply, lapply, function(matrices){
  ordinated_true <- cmdscale(matrices$true, k = ncol(matrices$true) - 2, add = TRUE)
    traitspace <- ordinated_true$points})

true <- ord$slow[[1]]$living

# perform principal components analysis

# project new data onto the PCA space
scale(newdata, v$center, true$scale) %*% true$rotation 

## To project things on a already estimated PCA (fixed eigenvector)
## 1- get your pca
my_pca <- prcomp(...)
## 2- get your new data centered
new_data <- scale(my_new_data, centre = TRUE, scale = FALSE) # not sure about scale FALSE
## 3- project it
new_pca <- stats::predict(my_pca, newdata = new_data)

# scaled_ord <- lapply(ord, function(preservation){
#   lapply(preservation, function(rate){
#     lapply(rate, function(treatment){
#       lapply(treatment, function(matrix){
#         scaled_pc <- scale(matrix)
#         return(scaled_pc)
#       })
#     })
#   })
# })


## Relative distance in trait space

morphospace.dist <- function(matrix, ...) {
    ## Get the pairwise distances between all points
    distances <- pairwise.dist(matrix)
    ## Calculate the longest distance from the center
    longest_distance_from_center <- max(as.numeric(centroids(matrix, centroid = 0)), na.rm = TRUE)
    ## Make all these distances relative to the longest distance from the center of the traitspace
    return(distances/longest_distance_from_center)
}



morpho_diffs <- lapply(ord, lapply, lapply, function(matrices){
  diffs <- abs((morphospace.dist(matrices$ace)) - (morphospace.dist(matrices$true)))
  averages <- median(diffs)
})

saveRDS(morpho_diffs, "../Data/matrices/morphospacediffs.Rds")

## can we make this relative disparity, so that we know what is good vs bad metrics for ancestral state estimation?
sum_var <- lapply(ord, lapply, lapply, lapply, function(matrix){
    disparity <- (dispRity(matrix, metric = c(sum, variances)))$disparity})

sum_var <- lapply(ord, lapply, lapply, lapply, function(matrix){
    disparity <- (dispRity(matrix, metric = c(sum, variances)))$disparity})


sum_var_low_dim <- lapply(ord, lapply, lapply, lapply, function(matrix){
    disparity <- (dispRity(matrix[,1:10], metric = c(sum, variances)))$disparity})

sum_ranges <- lapply(ord, lapply, lapply, lapply, function(matrix){
    disparity <- (dispRity(matrix, metric = c(sum, ranges)))$disparity})

av_displ <- lapply(ord, lapply, lapply, lapply, function(matrix){
    disparity <- (dispRity(matrix, metric = c(displacements)))$disparity})

ranges_diffs <- lapply(sum_ranges, lapply, lapply, function(matrices){
  diffs <- abs(matrices$ace[[1]]$elements[1] - matrices$true[[1]]$elements[1])
})

variance_diffs_low_dim <- lapply(sum_var_low_dim, lapply, lapply, function(matrices){
  diffs <- abs(matrices$ace[[1]]$elements[1] - matrices$true[[1]]$elements[1])
})

displ_diffs <- lapply(av_displ, lapply, lapply, function(matrices){
  diffs <- abs(matrices$ace[[1]]$elements[1] - matrices$true[[1]]$elements[1])
})




variance_diffs <- lapply(sum_var, lapply, lapply, function(matrices){
  diffs <- abs(matrices$ace[[1]]$elements[1] - matrices$true[[1]]$elements[1]) / matrices$true[[1]]$elements[1]
})
saveRDS(variance_diffs, "../Data/matrices/vardiffs.rds")
saveRDS(ranges_diffs, "../Data/matrices/rangediffs.rds")
saveRDS(variance_diffs_low_dim, "../Data/matrices/vardiffs_lowdim.rds")

saveRDS(displ_diffs, "../Data/matrices/dispdiffs.rds")

## Effect of removing fossils




plot(scaled_ace_ord$slow$fossil_high[[15]])
text(scaled_ace_ord$slow$fossil_high[[15]], labels = rownames(scaled_ace_ord$slow$fossil_high[[15]]), pos = 4, cex = 0.6)


# morphospace.dist <- function(ace_ord) {
#   ace_scores <- ace_ord[grepl("^ace", rownames(ace_ord)), ]
#   true_scores <- ace_ord[!grepl("^ace", rownames(ace_ord)), ]
#   distances <- sqrt(rowSums((ace_scores[,1:15] - true_scores[,1:15])^2))
#   return(distances)
# }

# ace_ord_scores <- ace_ord$fossil_high$low[[1]][grepl("^ace", rownames(ace_ord$fossil_high$low[[1]])), ]
# true_ord_scores <- ace_ord$fossil_high$low[[1]][!grepl("^ace", rownames(ace_ord$fossil_high$low[[1]])), ]
# test <- ace_ord_scores[94,1:10]
# test2 <- true_ord_scores[94,1:10]
# morpho_distances <- sqrt(rowSums((ace_ord_scores[,1:10] - true_ord_scores[,1:10])^2))
# ## tends to converge on exactly the same value
# # sqrt(sum((test - test2)^2))

morpho_dists <- lapply(scaled_ace_ord, lapply, lapply, morphospace.dist)
# morpho_dists_med <- unlist(morpho_dists$slow)
saveRDS(morpho_dists, "../Data/matrices/orddists.Rds")

unlisted_morpho_dists <- lapply(morpho_dists, function(rate) {
    unlist(rate, recursive = TRUE)
})

split_dists <- lapply(unlisted_morpho_dists, function(rate){
  split(rate, gsub("\\d+", "", names(rate)))
})

categories <- c("living", "fossil_high", "fossil_med", "fossil_low")
boxplot(split_dists$med, 
        main = "", 
        ylab = "", 
        xlab = "Categories", 
        las = 2)

## lowering number of pc axes increases high fossil recovery rate. Try this by lowering the number of pc axes included in distance function

mean(morpho_dists$fossil_low$low[[15]])
mean(morpho_dists$fossil_high$low[[15]])

## larger distances in fossil high cos range is larger - perhaps need to scale.
## using first 10 pc axes finds considerably more correct values.


```