---
title: "ACE Project discrete"
author: "Caleb"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

Things to note
- 85% binary, 15% 3 state
- 100% fossil preservation too
- only compare nodes present in all trees - keep root, but in living make t1 at time = 0 (effectively becomes outgroup)
- use z-score scaling or whatever it is with scale() function.
- explore with extinction.
- use median rather than mean.
- then look at disparity metrics
- the threshold? how do we know what threshold to set? can we set this by looking at our fossil coverage? because not setting a threshold makes ace too permissive, which creates artifacts in downstream analyses.
- Something to consider also - so these traits have been simulated directly proportional to the branch lengths... so this makes ace suitable. but when people perform ace on chronograms, is this comparable?
- How can we look at threshold for multi-state characters, not just binary?


## Generate trees

```{r}

library(treats)

bd_params <- make.bd.params(speciation = 1, extinction = 0.7)

stop_rule <- list(max.living = 50)
set.seed(123)
trees <- treats(stop.rule = stop_rule, bd.params = bd_params, null.error = 100, replicates = 100)

## write the saved trees
write.tree(trees, "../Data/trees/simulated.tre")
# write.tree(tree, file = "../Data/processed/trees/simulated.tre")
# tree <- read.tree("../Data/processed/trees/simulated.tre")

# tree <- read.nexus("../Data/raw/trees/Beck2014.tre")
# tree <- makeNodeLabel(tree)
# tree <- set.root.time(tree)
trees <- read.tree("../Data/trees/simulated.tre")

```


## Discrete trait simulation

Firstly lets simulate completely random matrix.

```{r}

tree_labels <- lapply(trees, function(tree) {
  c(tree$tip.label, tree$node.label)
})

# Create a matrix with random numbers between 0 and 3
# how do the number of character states influence ancestral states???? say if there are 5 different states, does this influence disparity analyses. could do another matrix where the number of states changes halfway through the tree to see how ace compares??? 
random_discrete <- lapply(tree_labels, function(tree){
                        matrix <- matrix(sample(0:2, length(tree) * 100, replace = TRUE), 
                        nrow = length(tree), 
                        ncol = 100)
                        rownames(matrix) <- tree
                        return(matrix)})





```


Now we can simulate trait data at varying levels of transition rate.

```{r}

slow_binary_transitions <- matrix(c(
    0.99, 0.01,
    0.01, 0.99
), nrow = 2, byrow = TRUE)

slow_multi_transitions <- matrix(c(
    0.99, 0.025, 0.025,
    0.025, 0.99, 0.025,
    0.025, 0.025, 0.99
), nrow = 3, byrow = TRUE)

slow_binary <- treats::make.traits(process = discrete.process, n = 85, process.args = list(transitions = slow_binary_transitions))
slow_multi <- treats::make.traits(process = discrete.process, n = 15, process.args = list(transitions = slow_multi_transitions))


med_binary_transitions <- matrix(c(
  0.90, 0.1,
  0.1, 0.9
), nrow = 2, byrow = TRUE)


med_multi_transitions <- matrix(c(
  0.9, 0.05, 0.05,
  0.05, 0.9, 0.05,
  0.05, 0.05, 0.9
), nrow = 3, byrow = TRUE)

med_binary <- treats::make.traits(process = discrete.process, n = 85, process.args = list(transitions = med_binary_transitions))
med_multi <- treats::make.traits(process = discrete.process, n = 15, process.args = list(transitions = med_multi_transitions))

fast_binary_transitions <- matrix(c(
    0.1, 0.9,
    0.9, 0.1
), nrow = 2, byrow = TRUE)

fast_multi_transitions <- matrix(c(
    0.1, 0.45, 0.45, 
    0.45, 0.1, 0.45,
    0.45, 0.45, 0.1 
), nrow = 3, byrow = TRUE)

fast_binary <- treats::make.traits(process = discrete.process, n = 85, process.args = list(transitions = fast_binary_transitions))
fast_multi <- treats::make.traits(process = discrete.process, n = 15, process.args = list(transitions = fast_multi_transitions))



## map the traits

slow_matrices <- lapply(trees, function(tree) {
    mapped_binary <- (map.traits(slow_binary, tree))$data
    mapped_multi <- (map.traits(slow_multi, tree))$data
    matrix <- cbind(mapped_binary, mapped_multi)
    return(matrix)
})

med_matrices <- lapply(trees, function(tree) {
    mapped_binary <- (map.traits(med_binary, tree))$data
    mapped_multi <- (map.traits(med_multi, tree))$data
    matrix <- cbind(mapped_binary, mapped_multi)
    return(matrix)
})

fast_matrices <- lapply(trees, function(tree) {
    mapped_binary <- (map.traits(fast_binary, tree))$data
    mapped_multi <- (map.traits(fast_multi, tree))$data
    matrix <- cbind(mapped_binary, mapped_multi)
    return(matrix)
})


discrete_matrices <- list(slow = slow_matrices, med = med_matrices, fast = fast_matrices)
saveRDS(discrete_matrices, "../Data/matrices/discretematrices.rds")
discrete_matrices <- readRDS("../Data/matrices/discretematrices.rds")

```

## Different tree coverages

```{r}

## fossil degredation as well

source("../Functions/fossil.pres.R")

living <- lapply(discrete_matrices, remove.fossil, trees = trees, type = "discrete")
fossilised_high <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.5, type = "discrete")
all_fossil <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 1.0, type = "discrete")
fossilised_med <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.15, type = "discrete")
fossilised_low <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.05, type = "discrete")



# plot(fossilised_low$slow[[13]]$tree)

fossil_matrices <- lapply(names(discrete_matrices), function(level) {
  lapply(seq_along(discrete_matrices[[level]]), function(i) {
    list(
      all = all_fossil[[level]][[i]],
      fossil_high = fossilised_high[[level]][[i]],
      fossil_med = fossilised_med[[level]][[i]],
      fossil_low = fossilised_low[[level]][[i]],
      living = living[[level]][[i]]
    )
  })
})

# Assign names to the outer list
names(fossil_matrices) <- names(discrete_matrices)

saveRDS(fossil_matrices, "../Data/matrices/discrete/fossilmatrices.Rds")



```


## discrete pipeline

```{R}
# source("../Functions/apply.dist.R")
# source("../Functions/apply.ordination.R")

# distances <- lapply(discrete_matrices, dist.func)
# traitspaces <- ord.func(distances)

# disp_obj <- lapply(traitspaces, make.dispRity, tree = tree)

# do.time.slice <- function(matrices, chrono_models, FADLAD = NULL, time_interval) {
#   time_slices <- lapply(chrono_models, function(model) {
#     chrono.subsets(matrices, method = "continuous", FADLAD = FADLAD, model = model, time = time_interval)
#   })
#   names(time_slices) <- chrono_models
#   return(time_slices)
# }

# chrono_models <- c("acctran", "deltran", "random", "proximity", "equal.split", "gradual.split")


# time_sliced_discrete <- lapply(disp_obj, do.time.slice, chrono_models = chrono_models, time_interval = 10)

# bootstrapped_discrete <- lapply(time_sliced_discrete, lapply, boot.matrix, bootstraps = 100)

# disparity_discrete <- lapply(bootstrapped_discrete, lapply,  dispRity, metric = c(sum, variances), na.rm = TRUE, verbose = TRUE)

# saveRDS(disparity_discrete, "../Data/processed/disparity/aceproject/beckdiscrete.Rds")
# saveRDS(disparity_discrete, "../Data/processed/disparity/aceproject/simdiscrete.Rds")


# plot(disparity_discrete$low$gradual.split)

```


## Ace discrete

```{r}
n_cores  <- 11
anc.states <- function(fossil_matrices) {
  lapply(fossil_matrices, function(level) {
    lapply(level, function(x) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = x$matrix, 
                          tree = x$tree, 
                          models = "SYM", 
                          output = "multi.ace",
                          verbose = TRUE,
                          parallel = n_cores)
  return(anc_states)})})
}

point.anc.states <- function(fossil_matrices) {
  lapply(fossil_matrices, function(level) {
    lapply(level, function(x) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = x$matrix, 
                          tree = x$tree, 
                          models = "SYM", 
                          output = "combined",
                          verbose = TRUE)
  return(anc_states)})})
  }
normal_sampling <- list(fun = rnorm, param = list(mean = mean, sd = function(x)return(diff(range(x))/4)))

sample.anc.states <- function(fossil_matrices) {
  lapply(fossil_matrices, function(level) {
    lapply(level, function(x) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = x,
                          models = "SYM",
                          sample = 5,
                          sample.fun = normal_sampling,
                          output = "combined",
                          verbose = TRUE)
  return(anc_states)})})
  }

fossil_anc <- lapply(fossil_matrices, anc.states)

# sampled_fossil_anc <- lapply(fossil_anc, sample.anc.states)

sample_fossil_anc <- lapply(fossil_anc, lapply, lapply, multi.ace, sample = 30, sample.fun = normal_sampling,  verbose = TRUE)


strict_fossil_anc <- lapply(fossil_anc, lapply, lapply, multi.ace, threshold = FALSE, output = "combined.matrix", verbose = TRUE)

relative_fossil_anc <- lapply(fossil_anc, lapply, lapply, multi.ace, output = "combined.matrix", verbose = TRUE)

fossil_anc <- point_fossil_anc
saveRDS(fossil_anc, "../Data/matrices/discrete/discreteanc.Rds")
saveRDS(sample_fossil_anc, "../Data/matrices/discrete/sampleddiscreteanc.Rds")
saveRDS(relative_fossil_anc, "../Data/matrices/discrete/relanc.Rds")
saveRDS(strict_fossil_anc, "../Data/matrices/discrete/threshanc.Rds")




```

## Make simmap

```{R}

# Function to run SIMMAP on discrete matrices


# test_matrix <- fossil_matrices$slow[[1]]$living
# what <- test_matrix$matrix[,1]

# test_simmap <- lapply(1:ncol(test_matrix$matrix), function(y){
#   characters <- character_matrix[,y]
#   simmap_trees <- make.simmap(test_matrix$tree, characters, model = "ER", nsim = 1)
# })


# simmap_trees <- make.simmap(test_matrix$tree, what , model = "ER", nsim = 100)
# ace <- describe.simmap(simmap_trees)$ace


# test <- fossil_matrices$slow[[1]]$living[[1]]

# test_handle <- handle.missing.data(test)

# test_simmap <- make.simmap(tree = fossil_matrices$slow[[1]]$living$tree, test_handle[[1]], model = "SYM", nsim = 10)
# describe <- describe.simmap(test_simmap)
# describe$ace


# test <- SCM.matrix(tree = test_matrix$tree, matrix = test_matrix$matrix, model.set = "ER", n.sim = 2, threshold = 0.5, name.input = "test")

# # Apply SIMMAP to all discrete matrices
# simmap_results <- lapply(names(discrete_matrices), function(rate) {
#   lapply(seq_along(discrete_matrices[[rate]]), function(i) {
#     # Get the tree and character matrix for the current replicate
#     tree <- trees[[i]]
#     char_matrix <- discrete_matrices[[rate]][[i]]
    
#     # Define the model (e.g., "SYM" for symmetric transitions)
#     model <- "SYM"
    
#     # Run SIMMAP
#     run_simmap(tree = tree, char_matrix = char_matrix, model = model, n_sim = 100, threshold = 0.5)
#   })
# })


# Assign names to the results
# names(simmap_results) <- names(discrete_matrices)






```


## Compare true vs ace nodes by distance

```{r}
## this function returns matrices of ace and true node states
get.living.nodes <- function(living) {
  if (inherits(living$living, "list")) {
  matrix <- living$living[[1]]
  } else {
  matrix <- living$living}
  node_v <- rownames(matrix)[grepl("^n", rownames(matrix))]
  return(node_v)
}


node_labels <- lapply(sampled_fossil_anc$slow, get.living.nodes)

# filter.nodes <- function(true, ace, node_v) {
#   ace_filtered <- ace[node_v, , drop = FALSE]
#   return(list(ace = ace_filtered))
# }

# nodes_living <- function(true_matrices, aced_matrices) {
#   output <- lapply(names(true_matrices), function(i) {

#     # Step 1: Get node names from "living" ACE
#     living_node <- get.living.nodes(
#       living = aced_matrices$living[[i]]
#     )

#     # Step 2: For the current level, filter all preservation states
#     filtered_preservations <- lapply(names(aced_matrices), function(pres) {
#       lapply(seq_along(aced_matrices[[pres]][[i]]), function(j) {
#         filter.nodes(
#           true = true_matrices[[i]],
#           ace = aced_matrices[[pres]][[i]][[j]],
#           node_v = living_node
#         )
#       })
#     })
#     names(filtered_preservations) <- names(aced_matrices)
#     return(filtered_preservations)
#   })

#   names(output) <- names(true_matrices)
#   return(output)
# }


sample_living_nodes <- lapply(seq_along(sampled_fossil_anc), function(rate) {
  lapply(seq_along(sampled_fossil_anc[[rate]]), function(i) {
    # Step 1: Compute the true matrix once for the current replicate (i)
    node_v <- node_labels[[i]]
    true <- discrete_matrices[[rate]][[i]][node_v, ]

    # Step 2: Iterate over preservation levels
    levels <- names(sampled_fossil_anc[[rate]][[i]])
    ace_out <- lapply(seq_along(levels), function(level_idx) {
      level_name <- levels[[level_idx]]
      reps <- sampled_fossil_anc[[rate]][[i]][[level_name]]

      # Filter ACE matrices for each replicate
      lapply(seq_along(reps), function(rep) {
        reps[[rep]][node_v, , drop = FALSE]
      })
    })

    # Assign names to the ACE output
    names(ace_out) <- levels

    # Return the ACE matrices and the single true matrix
    return(list(ace = ace_out, true = true))
  })
})

# Assign names to the outer list
names(sample_living_nodes) <- names(discrete_matrices)


point_living_nodes <- lapply(seq_along(point_fossil_anc), function(rate) {
  lapply(seq_along(point_fossil_anc[[rate]]), function(i) {
    # Step 1: Get the node labels
    node_v <- node_labels[[i]]

    # Step 2: Get the true matrix for the current replicate
    true <- discrete_matrices[[rate]][[i]][node_v, ]

    # Step 3: Iterate over preservation levels
    levels <- names(point_fossil_anc[[rate]][[i]])
    ace_out <- lapply(levels, function(level_name) {
      # Filter ACE matrix for the current preservation level
      point_fossil_anc[[rate]][[i]][[level_name]][node_v, , drop = FALSE]
    })

    # Assign names to the ACE output
    names(ace_out) <- levels

    # Return the ACE matrices and the single true matrix
    return(list(ace = ace_out, true = true))
  })
})

# Assign names to the outer list
names(point_living_nodes) <- names(discrete_matrices)


thresh_living_nodes <- lapply(seq_along(thresh_fossil_anc), function(rate) {
  lapply(seq_along(thresh_fossil_anc[[rate]]), function(i) {
    # Step 1: Get the node labels
    node_v <- node_labels[[i]]

    # Step 2: Get the true matrix for the current replicate
    true <- discrete_matrices[[rate]][[i]][node_v, ]

    # Step 3: Iterate over preservation levels
    levels <- names(thresh_fossil_anc[[rate]][[i]])
    ace_out <- lapply(levels, function(level_name) {
      # Filter ACE matrix for the current preservation level
      thresh_fossil_anc[[rate]][[i]][[level_name]][node_v, , drop = FALSE]
    })

    # Assign names to the ACE output
    names(ace_out) <- levels

    # Return the ACE matrices and the single true matrix
    return(list(ace = ace_out, true = true))
  })
})

# Assign names to the outer list
names(thresh_living_nodes) <- names(discrete_matrices)




source("../Functions/node.dist.R")

point_dist_strict <- lapply(point_living_nodes, lapply, node.dist, distance = "strict")

thresh_dist <- lapply(thresh_living_nodes, lapply, node.dist, distance = "strict")

root_dist_point <- lapply(point_dist_strict, lapply, lapply, function(matrix) {
  v <- matrix[1, ]
  mean(v)})

root_dist_thresh <- lapply(thresh_dist, lapply, lapply, function(matrix) {
  v <- matrix[1, ]
  mean(v)})


sample_dist_strict <- lapply(sample_living_nodes, lapply, node.dist, distance = "strict")

# root_dist_sample <- lapply(sample_dist_strict, function(rate){
#   lapply(rate, function(matrix){
#     roots <- lapply(matrix, function(level){
#       # Calculate the mean of all values in the level
#       all_values <- unlist(level, recursive = TRUE)
#       mean(all_values)
#     })
#     return(roots)
#   })
# })

root_dist_sample <- lapply(sample_dist_strict, function(rate){
  lapply(rate, function(matrix){
    roots <- lapply(matrix, function(level){
      # For each replicate in the level, get the first row and calculate mean
      root_values <- sapply(level, function(mat) {
        v <- mat[1, ]  # Get the first row (root)
        mean(v)        # Calculate mean of that row
      })
      # Return the mean of all root means for this level
      mean(root_values)
    })
    return(roots)
  })
})

saveRDS(root_dist_sample, "../Data/matrices/rootdistsamp.rds")
saveRDS(root_dist_point, "../Data/matrices/rootdistpoint.rds")
saveRDS(root_dist_thresh, "../Data/matrices/rootdistthresh.rds")





# dist_strict_sample <- lapply(sample_living_nodes, function(rate) {
# lapply(rate, function(matrix) {
#   lapply(matrix, function(pres) {
#     distances <- node.dist(pres, distance = "strict")
#   })
# })
# })

strict_means_sample <- lapply(sample_dist_strict, function(preservation) {
  lapply(preservation, function(level) {
    lapply(level, function(matrices) {
      # Compute the mean of each replicate
      replicate_means <- sapply(matrices, mean)
      
      # Compute the mean of the replicate means
      # overall_mean <- mean(replicate_means)
      
      return(replicate_means)
    })
  })
})


strict_means_point <- lapply(point_dist_strict, function(preservation) {
  lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    means <- mean(matrices)
  })
})
})

strict_means_thresh <- lapply(thresh_dist, function(preservation) {
  lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    means <- mean(matrices)
  })
})
})

## do i need to divide by number of rows?

lenient_means <- lapply(dist_lenient, function(preservation) {
  lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    means <- mean(matrices)
  })
})
})

saveRDS(lenient_means, "../Data/matrices/lenientmeans.rds")
saveRDS(strict_means_point, "../Data/matrices/pointstrictmeans.rds")
saveRDS(strict_means_sample, "../Data/matrices/samplestrictmeans.rds")
saveRDS(strict_means_thresh, "../Data/matrices/threshstrictmeans.rds")



```



## Comparison of morphospace

```{R}


## these are anc estimated matrices with relevant tips but only with relevant nodes.
# anc_matrices_living


source("../Functions/apply.dist.R")
source("../Functions/apply.ordination.R")

distances_point <- lapply(point_living_nodes, lapply, function(matrices) {
  true_dist <- char.diff(matrices$true, method = "mord", by.col = FALSE)
  ace <- matrices$ace
  ace_dist <- lapply(ace, function(pres) {
    dist <- char.diff(pres, method = "mord", by.col = FALSE)
  })
  return(list(true = true_dist, ace = ace_dist ))
})

distances_thresh <- lapply(thresh_living_nodes, lapply, function(matrices) {
  true_dist <- char.diff(matrices$true, method = "mord", by.col = FALSE)
  ace <- matrices$ace
  ace_dist <- lapply(ace, function(pres) {
    dist <- char.diff(pres, method = "mord", by.col = FALSE)
  })
  return(list(true = true_dist, ace = ace_dist ))
})

distances_sample <- lapply(sample_living_nodes, lapply, function(matrices) {
  true_dist <- char.diff(matrices$true, method = "mord", by.col = FALSE)
  ace <- matrices$ace
  ace_dist <- lapply(ace, function(pres) {
    lapply(pres, function(reps) {
      dist <- char.diff(reps, method = "mord", by.col = FALSE)
    })
  })
  return(list(true = true_dist, ace = ace_dist ))
})

ord_point <- lapply(distances_point, lapply, function(matrices) {
  true_ord <- (cmdscale(matrices$true, k = ncol(matrices$true) - 2, add = TRUE))$points
  ace <- matrices$ace
  ace_ord <- lapply(ace, function(pres) {
    ord <- (cmdscale(pres, k = ncol(pres) - 2, add = TRUE))$points
  })
  return(list(true = true_ord, ace = ace_ord ))
})

ord_thresh <- lapply(distances_thresh, lapply, function(matrices) {
  true_ord <- (cmdscale(matrices$true, k = ncol(matrices$true) - 2, add = TRUE))$points
  ace <- matrices$ace
  ace_ord <- lapply(ace, function(pres) {
    ord <- (cmdscale(pres, k = ncol(pres) - 2, add = TRUE))$points
  })
  return(list(true = true_ord, ace = ace_ord ))
})

ord_sample <- lapply(distances_sample, lapply, function(matrices) {
  true_ord <- (cmdscale(matrices$true, k = ncol(matrices$true) - 2, add = TRUE))$points
  ace <- matrices$ace
  ace_ord <- lapply(ace, function(pres) {
    lapply(pres, function(reps) {
      ord <- (cmdscale(reps, k = ncol(reps) - 2, add = TRUE))$points
    })
  })
  return(list(true = true_ord, ace = ace_ord ))
})




# perform principal components analysis

# project new data onto the PCA space
scale(newdata, v$center, true$scale) %*% true$rotation 

## To project things on a already estimated PCA (fixed eigenvector)
## 1- get your pca
my_pca <- prcomp(...)
## 2- get your new data centered
new_data <- scale(my_new_data, centre = TRUE, scale = FALSE) # not sure about scale FALSE
## 3- project it
new_pca <- stats::predict(my_pca, newdata = new_data)

# scaled_ord <- lapply(ord, function(preservation){
#   lapply(preservation, function(rate){
#     lapply(rate, function(treatment){
#       lapply(treatment, function(matrix){
#         scaled_pc <- scale(matrix)
#         return(scaled_pc)
#       })
#     })
#   })
# })


## Relative distance in trait space

morphospace.dist <- function(matrix, ...) {
    ## Get the pairwise distances between all points
    distances <- pairwise.dist(matrix)
    ## Calculate the longest distance from the center
    longest_distance_from_center <- max(as.numeric(centroids(matrix, centroid = 0)), na.rm = TRUE)
    ## Make all these distances relative to the longest distance from the center of the traitspace
    return(distances/longest_distance_from_center)
}



morpho_diffs <- lapply(ord, lapply, lapply, function(matrices){
  diffs <- abs((morphospace.dist(matrices$ace)) - (morphospace.dist(matrices$true)))
  averages <- median(diffs)
})

saveRDS(morpho_diffs, "../Data/matrices/morphospacediffs.Rds")

## can we make this relative disparity, so that we know what is good vs bad metrics for ancestral state estimation?

## i think this should be the whole morphospace, rather than just identical nodes.
sum_var_point <- lapply(ord_point, lapply, function(matrices) {
  true_disp <- (dispRity(matrices$true, metric = c(sum, variances)))$disparity
  ace <- matrices$ace
  ace_disp <- lapply(ace, function(pres) {
    disp <- (dispRity(pres, metric = c(sum, variances)))$disparity
  })
  return(list(true = true_disp, ace = ace_disp ))
})


sum_var_thresh <- lapply(ord_thresh, lapply, function(matrices) {
  true_disp <- (dispRity(matrices$true, metric = c(sum, variances)))$disparity
  ace <- matrices$ace
  ace_disp <- lapply(ace, function(pres) {
    disp <- (dispRity(pres, metric = c(sum, variances)))$disparity
  })
  return(list(true = true_disp, ace = ace_disp ))
})

sum_var_sample <- lapply(ord_sample, lapply, function(matrices) {
  true_disp <- (dispRity(matrices$true, metric = c(sum, variances)))$disparity
  ace <- matrices$ace
  ace_disp <- lapply(ace, function(pres) {
    lapply(pres, function(reps) {
      disp <- (dispRity(reps, metric = c(sum, variances)))$disparity
    })
  })
  return(list(true = true_disp, ace = ace_disp ))
})


var_point_diffs <- lapply(sum_var_point, lapply, function(disparity){
  true <- disparity$true[[1]]$elements[1]
  ace <- disparity$ace
  diffs <- lapply(ace, function(pres){
    diff <- (pres[[1]]$elements[1] - true) / true
  })
  return(diffs)
})

var_thresh_diffs <- lapply(sum_var_thresh, lapply, function(disparity){
  true <- disparity$true[[1]]$elements[1]
  ace <- disparity$ace
  diffs <- lapply(ace, function(pres){
    diff <- (pres[[1]]$elements[1] - true) / true
  })
  return(diffs)
})

var_sample_diffs <- lapply(sum_var_sample, lapply, function(disparity) {
  true <- disparity$true[[1]]$elements[1]
  ace <- disparity$ace
  diffs <- lapply(ace, function(pres) {
    lapply(pres, function(rep) {
      # Compute the difference
      diff <- (rep[[1]]$elements[1] - true) / true
      return(diff)
    })
  })
  
  # Apply unlist to the lowest level
  diffs <- lapply(diffs, function(pres) {
    unlist(pres, recursive = TRUE)
  })
  
  return(diffs)
})


sum_range_point <- lapply(ord_point, lapply, function(matrices) {
  true_disp <- (dispRity(matrices$true, metric = c(sum, ranges)))$disparity
  ace <- matrices$ace
  ace_disp <- lapply(ace, function(pres) {
    disp <- (dispRity(pres, metric = c(sum, ranges)))$disparity
  })
  return(list(true = true_disp, ace = ace_disp ))
})

sum_range_sample <- lapply(ord_sample, lapply, function(matrices) {
  true_disp <- (dispRity(matrices$true, metric = c(sum, ranges)))$disparity
  ace <- matrices$ace
  ace_disp <- lapply(ace, function(pres) {
    lapply(pres, function(reps) {
      disp <- (dispRity(reps, metric = c(sum, ranges)))$disparity
    })
  })
  return(list(true = true_disp, ace = ace_disp ))
})


range_point_diffs <- lapply(sum_range_point, lapply, function(disparity){
  true <- disparity$true[[1]]$elements[1]
  ace <- disparity$ace
  diffs <- lapply(ace, function(pres){
    diff <- abs(pres[[1]]$elements[1] - true) / true
  })
  return(diffs)
})

range_sample_diffs <- lapply(sum_range_sample, lapply, function(disparity) {
  true <- disparity$true[[1]]$elements[1]
  ace <- disparity$ace
  diffs <- lapply(ace, function(pres) {
    lapply(pres, function(rep) {
      # Compute the difference
      diff <- abs(rep[[1]]$elements[1] - true) / true
      return(diff)
    })
  })
  
  # Apply unlist to the lowest level
  diffs <- lapply(diffs, function(pres) {
    unlist(pres, recursive = TRUE)
  })
  
  return(diffs)
})





neighbours_point <- lapply(ord_point, lapply, function(matrices) {
  true_disp <- (dispRity(matrices$true, metric = c(mean, neighbours)))$disparity
  ace <- matrices$ace
  ace_disp <- lapply(ace, function(pres) {
    disp <- (dispRity(pres, metric = c(mean, neighbours)))$disparity
  })
  return(list(true = true_disp, ace = ace_disp ))
})

neighbours_sample <- lapply(ord_sample, lapply, function(matrices) {
  true_disp <- (dispRity(matrices$true, metric = c(mean, neighbours)))$disparity
  ace <- matrices$ace
  ace_disp <- lapply(ace, function(pres) {
    lapply(pres, function(reps) {
      disp <- (dispRity(reps, metric = c(mean,neighbours)))$disparity
    })
  })
  return(list(true = true_disp, ace = ace_disp ))
})


neighbours_point_diffs <- lapply(neighbours_point, lapply, function(disparity){
  true <- disparity$true[[1]]$elements[1]
  ace <- disparity$ace
  diffs <- lapply(ace, function(pres){
    diff <- abs(pres[[1]]$elements[1] - true) / true
  })
  return(diffs)
})

neighbours_sample_diffs <- lapply(neighbours_sample, lapply, function(disparity) {
  true <- disparity$true[[1]]$elements[1]
  ace <- disparity$ace
  diffs <- lapply(ace, function(pres) {
    lapply(pres, function(rep) {
      # Compute the difference
      diff <- abs(rep[[1]]$elements[1] - true) / true
      return(diff)
    })
  })
  
  # Apply unlist to the lowest level
  diffs <- lapply(diffs, function(pres) {
    unlist(pres, recursive = TRUE)
  })
  
  return(diffs)
})



saveRDS(var_point_diffs, "../Data/matrices/discrete/relvardiffs.rds")
saveRDS(var_sample_diffs, "../Data/matrices/discrete/samplevardiffs.rds")
saveRDS(var_thresh_diffs, "../Data/matrices/discrete/strictvardiffs.rds")

saveRDS(range_point_diffs, "../Data/matrices/pointrangediffs.rds")
saveRDS(range_sample_diffs, "../Data/matrices/samplerangediffs.rds")
## Effect of removing fossils

saveRDS(neighbours_point_diffs, "../Data/matrices/pointneighboursdiffs.rds")
saveRDS(neighbours_sample_diffs, "../Data/matrices/sampleneighoursdiffs.rds")



plot(scaled_ace_ord$slow$fossil_high[[15]])
text(scaled_ace_ord$slow$fossil_high[[15]], labels = rownames(scaled_ace_ord$slow$fossil_high[[15]]), pos = 4, cex = 0.6)


# morphospace.dist <- function(ace_ord) {
#   ace_scores <- ace_ord[grepl("^ace", rownames(ace_ord)), ]
#   true_scores <- ace_ord[!grepl("^ace", rownames(ace_ord)), ]
#   distances <- sqrt(rowSums((ace_scores[,1:15] - true_scores[,1:15])^2))
#   return(distances)
# }

# ace_ord_scores <- ace_ord$fossil_high$low[[1]][grepl("^ace", rownames(ace_ord$fossil_high$low[[1]])), ]
# true_ord_scores <- ace_ord$fossil_high$low[[1]][!grepl("^ace", rownames(ace_ord$fossil_high$low[[1]])), ]
# test <- ace_ord_scores[94,1:10]
# test2 <- true_ord_scores[94,1:10]
# morpho_distances <- sqrt(rowSums((ace_ord_scores[,1:10] - true_ord_scores[,1:10])^2))
# ## tends to converge on exactly the same value
# # sqrt(sum((test - test2)^2))

morpho_dists <- lapply(scaled_ace_ord, lapply, lapply, morphospace.dist)
# morpho_dists_med <- unlist(morpho_dists$slow)
saveRDS(morpho_dists, "../Data/matrices/orddists.Rds")

unlisted_morpho_dists <- lapply(morpho_dists, function(rate) {
    unlist(rate, recursive = TRUE)
})

split_dists <- lapply(unlisted_morpho_dists, function(rate){
  split(rate, gsub("\\d+", "", names(rate)))
})

categories <- c("living", "fossil_high", "fossil_med", "fossil_low")
boxplot(split_dists$med, 
        main = "", 
        ylab = "", 
        xlab = "Categories", 
        las = 2)

## lowering number of pc axes increases high fossil recovery rate. Try this by lowering the number of pc axes included in distance function

mean(morpho_dists$fossil_low$low[[15]])
mean(morpho_dists$fossil_high$low[[15]])

## larger distances in fossil high cos range is larger - perhaps need to scale.
## using first 10 pc axes finds considerably more correct values.


```


## All nodes and raw distance

This is going to be run including tips

```{r}

rel_all_nodes <- lapply(seq_along(point_fossil_anc), function(rate) {
  lapply(seq_along(point_fossil_anc[[rate]]), function(i) {
    # Step 1: Get the node labels
    # node_v <- node_labels[[i]]

    # Step 2: Get the true matrix for the current replicate
    # true <- discrete_matrices[[rate]][[i]][rownames(i), ]

    # Step 3: Iterate over preservation levels
    levels <- names(point_fossil_anc[[rate]][[i]])
    ace_out <- lapply(levels, function(level_name) {
      # Filter ACE matrix for the current preservation level
      ace <- point_fossil_anc[[rate]][[i]][[level_name]]
      ace_filtered <- ace[grepl("^n", rownames(ace)), ]
      true <- discrete_matrices[[rate]][[i]][rownames(ace_filtered), ]
      return(list(true = true, ace = ace_filtered))
    })

    # Assign names to the ACE output
    names(ace_out) <- levels

    # Return the ACE matrices and the single true matrix
    return(ace_out)
  })
})

# Assign names to the outer list
names(rel_all_nodes) <- names(discrete_matrices)

thresh_all_nodes <- lapply(seq_along(thresh_fossil_anc), function(rate) {
  lapply(seq_along(thresh_fossil_anc[[rate]]), function(i) {
    # Step 1: Get the node labels
    # node_v <- node_labels[[i]]

    # Step 2: Get the true matrix for the current replicate
    # true <- discrete_matrices[[rate]][[i]][rownames(i), ]

    # Step 3: Iterate over preservation levels
    levels <- names(thresh_fossil_anc[[rate]][[i]])
    ace_out <- lapply(levels, function(level_name) {
      # Filter ACE matrix for the current preservation level
      ace <- thresh_fossil_anc[[rate]][[i]][[level_name]]
      ace_filtered <- ace[grepl("^n", rownames(ace)), ]
      true <- discrete_matrices[[rate]][[i]][rownames(ace_filtered), ]
      return(list(true = true, ace = ace_filtered))
    })

    # Assign names to the ACE output
    names(ace_out) <- levels

    # Return the ACE matrices and the single true matrix
    return(ace_out)
  })
})

# Assign names to the outer list
names(thresh_all_nodes) <- names(discrete_matrices)



sample_all_nodes <- lapply(seq_along(sampled_fossil_anc), function(rate) {
  lapply(seq_along(sampled_fossil_anc[[rate]]), function(i) {
    
    # Step 1: Iterate over preservation levels
    levels <- names(sampled_fossil_anc[[rate]][[i]]) # e.g., c("fossil", "living")
    
    ace_out <- lapply(levels, function(level_name) {
      # Get all replicates for this level
      replicates <- sampled_fossil_anc[[rate]][[i]][[level_name]]
      
      # Step 2: Filter each replicate to keep only rows starting with "n"
      filtered <- lapply(seq_along(replicates), function(rep_idx) {
        ace <- replicates[[rep_idx]]
        ace_filtered <- ace[grepl("^n", rownames(ace)), ]
        return(ace_filtered)
      })
      
      # Step 3: Get the true matrix using the same rows as the first filtered replicate
      if (length(filtered) > 0 && nrow(filtered[[1]]) > 0) {
        true <- discrete_matrices[[rate]][[i]][rownames(filtered[[1]]), ]
      } else {
        true <- NULL
      }
      
      return(list(true = true, ace = filtered))
    })
    
    # Assign names to the ACE output
    names(ace_out) <- levels
    
    # Return the ACE matrices and true matrix for each level
    return(ace_out)
  })
})

# Assign names to the outer list
names(sample_all_nodes) <- names(discrete_matrices)







rel_dist <- lapply(rel_all_nodes, lapply, lapply, all.node.dist, distance = "strict")

thresh_dist <- lapply(thresh_all_nodes, lapply, lapply, all.node.dist, distance = "strict")

sample_dist <- lapply(sample_all_nodes, lapply, lapply, all.node.dist, distance = "strict")

allnode_means_rel <- lapply(rel_dist, function(preservation) {
  lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    means <- mean(matrices)
  })
})
})

allnode_means_thresh <- lapply(thresh_dist, function(preservation) {
  lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    means <- mean(matrices)
  })
})
})

allnode_means_sample <- lapply(sample_dist, function(rate_level) {
  lapply(rate_level, function(replicate) {
    lapply(replicate, function(level) {
     q <-  lapply(level, function(x) {
        m <- mean(x)
      })
      unlist <- unlist(q)
      return(unlist)
    })
  })
})


saveRDS(allnode_means_rel, "../Data/matrices/discrete/rel_raw_dist.Rds")
saveRDS(allnode_means_thresh, "../Data/matrices/discrete/thresh_raw_dist.Rds")
saveRDS(allnode_means_sample, "../Data/matrices/discrete/sample_raw_dist.Rds")


```


## Disparity for all nodes and tips



```{r}


## add to this - > just tips disparity? so it is true (tips + nodes all) , aced (tips + estimated node states) and just tips (varies by fossil sampling like aced).

sample_all <- lapply(seq_along(sample_fossil_anc), function(rate) {
  lapply(seq_along(sample_fossil_anc[[rate]]), function(i) {
    # Step 1: Compute the true matrix once for the current replicate (i)
    # node_v <- node_labels[[i]]
    true <- discrete_matrices[[rate]][[i]]
    true <- apply(true, c(1,2), as.character)

    # Step 2: Iterate over preservation levels
    levels <- names(sample_fossil_anc[[rate]][[i]])
    ace_out <- lapply(seq_along(levels), function(level_idx) {
      level_name <- levels[[level_idx]]
      reps <- sample_fossil_anc[[rate]][[i]][[level_name]]

      # Filter ACE matrices for each replicate
      lapply(seq_along(reps), function(rep) {
        reps[[rep]]
      })
    })

    # Assign names to the ACE output
    names(ace_out) <- levels

    # Return the ACE matrices and the single true matrix
    return(list(ace = ace_out, true = true))
  })
})

# Assign names to the outer list
names(sample_all) <- names(discrete_matrices)


rel_all <- lapply(seq_along(relative_fossil_anc), function(rate) {
  lapply(seq_along(relative_fossil_anc[[rate]]), function(i) {
    # Step 1: Get the node labels
    # node_v <- node_labels[[i]]

    # Step 2: Get the true matrix for the current replicate
    true <- discrete_matrices[[rate]][[i]]
    true <- apply(true, c(1,2), as.character)

    # Step 3: Iterate over preservation levels
    levels <- names(relative_fossil_anc[[rate]][[i]])
    ace_out <- lapply(levels, function(level_name) {
      # Filter ACE matrix for the current preservation level
      relative_fossil_anc[[rate]][[i]][[level_name]]
    })

    # Assign names to the ACE output
    names(ace_out) <- levels

    # Return the ACE matrices and the single true matrix
    return(list(ace = ace_out, true = true))
  })
})

# Assign names to the outer list
names(rel_all) <- names(discrete_matrices)




strict_all <- lapply(seq_along(strict_fossil_anc), function(rate) {
  lapply(seq_along(strict_fossil_anc[[rate]]), function(i) {
    # Step 1: Get the node labels
    # node_v <- node_labels[[i]]

    # Step 2: Get the true matrix for the current replicate
    true <- discrete_matrices[[rate]][[i]]
    true <- apply(true, c(1,2), as.character)

    # Step 3: Iterate over preservation levels
    levels <- names(strict_fossil_anc[[rate]][[i]])
    ace_out <- lapply(levels, function(level_name) {
      # Filter ACE matrix for the current preservation level
      strict_fossil_anc[[rate]][[i]][[level_name]]
    })

    # Assign names to the ACE output
    names(ace_out) <- levels

    # Return the ACE matrices and the single true matrix
    return(list(ace = ace_out, true = true))
  })
})

# Assign names to the outer list
names(strict_all) <- names(discrete_matrices)


source("../Functions/apply.dist.R")
source("../Functions/apply.ordination.R")

distances_strict <- lapply(strict_all, lapply, function(matrices) {
  true_dist <- char.diff(matrices$true, method = "mord", by.col = FALSE)
  ace <- matrices$ace
  ace_dist <- lapply(ace, function(pres) {
    dist <- char.diff(pres, method = "mord", by.col = FALSE)
  })
  return(list(true = true_dist, ace = ace_dist ))
})

distances_rel <- lapply(rel_all, lapply, function(matrices) {
  true_dist <- char.diff(matrices$true, method = "mord", by.col = FALSE)
  ace <- matrices$ace
  ace_dist <- lapply(ace, function(pres) {
    dist <- char.diff(pres, method = "mord", by.col = FALSE)
  })
  return(list(true = true_dist, ace = ace_dist ))
})

distances_sample <- lapply(sample_all, lapply, function(matrices) {
  true_dist <- char.diff(matrices$true, method = "mord", by.col = FALSE)
  ace <- matrices$ace
  ace_dist <- lapply(ace, function(pres) {
    lapply(pres, function(reps) {
      dist <- char.diff(reps, method = "mord", by.col = FALSE)
    })
  })
  return(list(true = true_dist, ace = ace_dist ))
})

ord_rel <- lapply(distances_rel, lapply, function(matrices) {
  true_ord <- (cmdscale(matrices$true, k = ncol(matrices$true) - 2, add = TRUE))$points
  ace <- matrices$ace
  ace_ord <- lapply(ace, function(pres) {
    ord <- (cmdscale(pres, k = ncol(pres) - 2, add = TRUE))$points
  })
  return(list(true = true_ord, ace = ace_ord ))
})

ord_strict <- lapply(distances_strict, lapply, function(matrices) {
  true_ord <- (cmdscale(matrices$true, k = ncol(matrices$true) - 2, add = TRUE))$points
  ace <- matrices$ace
  ace_ord <- lapply(ace, function(pres) {
    ord <- (cmdscale(pres, k = ncol(pres) - 2, add = TRUE))$points
  })
  return(list(true = true_ord, ace = ace_ord ))
})

ord_sample <- lapply(distances_sample, lapply, function(matrices) {
  true_ord <- (cmdscale(matrices$true, k = ncol(matrices$true) - 2, add = TRUE))$points
  ace <- matrices$ace
  ace_ord <- lapply(ace, function(pres) {
    lapply(pres, function(reps) {
      ord <- (cmdscale(reps, k = ncol(reps) - 2, add = TRUE))$points
    })
  })
  return(list(true = true_ord, ace = ace_ord ))
})

### is it to do with the fact that they are ordinated at different dimensions??? should they all be ordinated at a set number of dimensions

## i've decided to keep the first 50?


## Why then 


sum_var_rel <- lapply(ord_rel, lapply, function(matrices) {
  true_disp <- (dispRity(matrices$true[, 1:48], metric = c(sum, variances)))$disparity
  ace <- matrices$ace
  ace_disp <- lapply(ace, function(pres) {
    disp <- (dispRity(pres[ ,1:48], metric = c(sum, variances)))$disparity
  })
  return(list(true = true_disp, ace = ace_disp ))
})


sum_var_strict <- lapply(ord_strict, lapply, function(matrices) {
  true_disp <- (dispRity(matrices$true[ ,1:48], metric = c(sum, variances)))$disparity
  ace <- matrices$ace
  ace_disp <- lapply(ace, function(pres) {
    disp <- (dispRity(pres[ ,1:48], metric = c(sum, variances)))$disparity
  })
  return(list(true = true_disp, ace = ace_disp ))
})

sum_var_sample <- lapply(ord_sample, lapply, function(matrices) {
  true_disp <- (dispRity(matrices$true[ ,1:48], metric = c(sum, variances)))$disparity
  ace <- matrices$ace
  ace_disp <- lapply(ace, function(pres) {
    lapply(pres, function(reps) {
      disp <- (dispRity(reps[ ,1:48], metric = c(sum, variances)))$disparity
    })
  })
  return(list(true = true_disp, ace = ace_disp ))
})


var_rel_diffs <- lapply(sum_var_rel, lapply, function(disparity){
  true <- disparity$true[[1]]$elements[1]
  ace <- disparity$ace
  diffs <- lapply(ace, function(pres){
    diff <- (pres[[1]]$elements[1] - true) / true
  })
  return(diffs)
})

var_strict_diffs <- lapply(sum_var_strict, lapply, function(disparity){
  true <- disparity$true[[1]]$elements[1]
  ace <- disparity$ace
  diffs <- lapply(ace, function(pres){
    diff <- (pres[[1]]$elements[1] - true) / true
  })
  return(diffs)
})

var_sample_diffs <- lapply(sum_var_sample, lapply, function(disparity) {
  true <- disparity$true[[1]]$elements[1]
  ace <- disparity$ace
  diffs <- lapply(ace, function(pres) {
    lapply(pres, function(rep) {
      # Compute the difference
      diff <- (rep[[1]]$elements[1] - true) / true
      return(diff)
    })
  })
  
  # Apply unlist to the lowest level
  diffs <- lapply(diffs, function(pres) {
    unlist(pres, recursive = TRUE)
  })
  
  return(diffs)
})
```

## Post-ord ace

```{r}
# load the discrete matrices
discrete_matrices <- readRDS("../Data/matrices/discretematrices.rds")

source("../Functions/fossil.pres.R")

living <- lapply(discrete_matrices, remove.fossil, trees = trees, type = "discrete")
fossilised_high <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.5, type = "discrete")
all_fossil <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 1.0, type = "discrete")
fossilised_med <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.15, type = "discrete")
fossilised_low <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.05, type = "discrete")





# plot(fossilised_low$slow[[13]]$tree)

fossil_matrices <- lapply(names(discrete_matrices), function(level) {
  lapply(seq_along(discrete_matrices[[level]]), function(i) {
    list(
      all = all_fossil[[level]][[i]],
      fossil_high = fossilised_high[[level]][[i]],
      fossil_med = fossilised_med[[level]][[i]],
      fossil_low = fossilised_low[[level]][[i]],
      living = living[[level]][[i]]
    )
  })
})

# Assign names to the outer list
names(fossil_matrices) <- names(discrete_matrices)
## Firstly, find the overall disparity




distances <- lapply(fossil_matrices, lapply, lapply, function(x) {
  matrix <- x$matrix
  tree <- x$tree
  distance_matrix <- char.diff(matrix, method = "mord", by.col = FALSE)
  return(list(tree = tree, matrix = distance_matrix))
})


ordinations <- lapply(distances, lapply, lapply, function(x) {
  matrix <- x$matrix
  tree <- x$tree
  ord_matrix <- (cmdscale(matrix, k = ncol(matrix) - 2, add = TRUE))$points
  return(list(tree = tree, matrix = ord_matrix))
})

n_cores <- 11
anc.states <- function(x){ 
  anc_states <- multi.ace(data = x$matrix[,1:48], 
                          tree = x$tree, 
                          models = "BM", 
                          output = "multi.ace",
                          parallel = n_cores,
                          verbose = TRUE)
  return(anc_states)
}



post_ord_ace <- lapply(ordinations, lapply, lapply, anc.states)

```