---
title: "ACE Project discrete"
author: "Caleb"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

Things to note
- 85% binary, 15% 3 state
- 100% fossil preservation too
- only compare nodes present in all trees - keep root, but in living make t1 at time = 0 (effectively becomes outgroup)
- use z-score scaling or whatever it is with scale() function.
- explore with extinction.
- use median rather than mean.
- then look at disparity metrics
- the threshold? how do we know what threshold to set? can we set this by looking at our fossil coverage? because not setting a threshold makes ace too permissive, which creates artifacts in downstream analyses.
- Something to consider also - so these traits have been simulated directly proportional to the branch lengths... so this makes ace suitable. but when people perform ace on chronograms, is this comparable?
- How can we look at threshold for multi-state characters, not just binary?


## Generate trees

```{r}

library(treats)

bd_params <- make.bd.params(speciation = 1, extinction = 0.7)

stop_rule <- list(max.living = 50)
set.seed(123)
trees <- treats(stop.rule = stop_rule, bd.params = bd_params, null.error = 100, replicates = 10)

## write the saved trees
write.tree(trees, "../Data/trees/simulated.tre")
# write.tree(tree, file = "../Data/processed/trees/simulated.tre")
# tree <- read.tree("../Data/processed/trees/simulated.tre")

# tree <- read.nexus("../Data/raw/trees/Beck2014.tre")
# tree <- makeNodeLabel(tree)
# tree <- set.root.time(tree)
trees <- read.tree("../Data/trees/simulated.tre")

```


## Discrete trait simulation

Now we can simulate trait data at varying levels of transition rate.

```{r}

slow_binary_transitions <- matrix(c(
    0.99, 0.01,
    0.01, 0.99
), nrow = 2, byrow = TRUE)

slow_multi_transitions <- matrix(c(
    0.99, 0.025, 0.025,
    0.025, 0.99, 0.025,
    0.025, 0.025, 0.99
), nrow = 3, byrow = TRUE)

slow_binary <- treats::make.traits(process = discrete.process, n = 85, process.args = list(transitions = slow_binary_transitions))
slow_multi <- treats::make.traits(process = discrete.process, n = 15, process.args = list(transitions = slow_multi_transitions))


med_binary_transitions <- matrix(c(
  0.90, 0.1,
  0.1, 0.9
), nrow = 2, byrow = TRUE)


med_multi_transitions <- matrix(c(
  0.9, 0.05, 0.05,
  0.05, 0.9, 0.05,
  0.05, 0.05, 0.9
), nrow = 3, byrow = TRUE)

med_binary <- treats::make.traits(process = discrete.process, n = 85, process.args = list(transitions = med_binary_transitions))
med_multi <- treats::make.traits(process = discrete.process, n = 15, process.args = list(transitions = med_multi_transitions))

fast_binary_transitions <- matrix(c(
    0.1, 0.9,
    0.9, 0.1
), nrow = 2, byrow = TRUE)

fast_multi_transitions <- matrix(c(
    0.1, 0.45, 0.45, 
    0.45, 0.1, 0.45,
    0.45, 0.45, 0.1 
), nrow = 3, byrow = TRUE)

fast_binary <- treats::make.traits(process = discrete.process, n = 85, process.args = list(transitions = fast_binary_transitions))
fast_multi <- treats::make.traits(process = discrete.process, n = 15, process.args = list(transitions = fast_multi_transitions))



## map the traits

slow_matrices <- lapply(trees, function(tree) {
    mapped_binary <- (map.traits(slow_binary, tree))$data
    mapped_multi <- (map.traits(slow_multi, tree))$data
    colnames(mapped_multi) <- c(paste0("B", 1:15))
    matrix <- cbind(mapped_binary, mapped_multi)
    return(matrix)
})

med_matrices <- lapply(trees, function(tree) {
    mapped_binary <- (map.traits(med_binary, tree))$data
    mapped_multi <- (map.traits(med_multi, tree))$data
    colnames(mapped_multi) <- c(paste0("B", 1:15))
    matrix <- cbind(mapped_binary, mapped_multi)
    return(matrix)
})

fast_matrices <- lapply(trees, function(tree) {
    mapped_binary <- (map.traits(fast_binary, tree))$data
    mapped_multi <- (map.traits(fast_multi, tree))$data
    colnames(mapped_multi) <- c(paste0("B", 1:15))
    matrix <- cbind(mapped_binary, mapped_multi)
    return(matrix)
})


discrete_matrices <- list(slow = slow_matrices, med = med_matrices, fast = fast_matrices)
saveRDS(discrete_matrices, "../Data/matrices/discrete/discretematrices.rds")
discrete_matrices <- readRDS("../Data/matrices/discrete/discretematrices.rds")





```

## Different tree coverages

```{r}

## fossil degredation as well

source("../Functions/fossil.pres.R")

living <- lapply(discrete_matrices, remove.fossil, trees = trees, type = "discrete")
fossilised_high <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.5, type = "discrete")
all_fossil <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 1.0, type = "discrete")
fossilised_med <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.15, type = "discrete")
fossilised_low <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.05, type = "discrete")



# plot(fossilised_low$slow[[13]]$tree)

fossil_matrices <- lapply(names(discrete_matrices), function(level) {
  lapply(seq_along(discrete_matrices[[level]]), function(i) {
    list(
      all = all_fossil[[level]][[i]],
      fossil_high = fossilised_high[[level]][[i]],
      fossil_med = fossilised_med[[level]][[i]],
      fossil_low = fossilised_low[[level]][[i]],
      living = living[[level]][[i]]
    )
  })
})

# Assign names to the outer list
names(fossil_matrices) <- names(discrete_matrices)

saveRDS(fossil_matrices, "../Data/matrices/discrete/fossilmatrices.Rds")

fossil_matrices <- readRDS("../Data/matrices/discrete/fossilmatrices.Rds")


```

## Ace discrete

```{r}
n_cores  <- 19
anc.states <- function(x) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = x$matrix, 
                          tree = x$tree, 
                          models = "SYM", 
                          output = "multi.ace",
                          verbose = TRUE,
                          parallel = n_cores)
return(anc_states)}


fossil_anc <- lapply(fossil_matrices, lapply, lapply, anc.states)
names(fossil_anc) <- names(fossil_matrices)

# sampled_fossil_anc <- lapply(fossil_anc, sample.anc.states)

sample_fossil_anc <- lapply(fossil_anc, lapply, lapply, multi.ace, sample = 16)

strict_fossil_anc <- lapply(fossil_anc, lapply, lapply, multi.ace, threshold = FALSE, output = "combined.matrix", verbose = TRUE)

relative_fossil_anc <- lapply(fossil_anc, lapply, lapply, multi.ace, output = "combined.matrix", verbose = TRUE)

saveRDS(fossil_anc, "../Data/matrices/discrete/discreteanc.Rds")
saveRDS(sample_fossil_anc, "../Data/matrices/discrete/sampleddiscreteanc.Rds")
saveRDS(relative_fossil_anc, "../Data/matrices/discrete/relanc.Rds")
saveRDS(strict_fossil_anc, "../Data/matrices/discrete/strictanc.Rds")

fossil_anc <- readRDS("../Data/matrices/discrete/discreteanc.Rds")

relative_fossil_anc <- readRDS("../Data/matrices/discrete/relanc.Rds")
strict_fossil_anc <- readRDS("../Data/matrices/discrete/strictanc.Rds")
sample_fossil_anc <- readRDS("../Data/matrices/discrete/sampleddiscreteanc.Rds")

```

## Uncertainty in estimates

```{R}
fossil_anc$slow[[1]]$living$discrete$estimates[[1]]

test <- lapply(fossil_anc$slow[[1]]$living$discrete$estimates[[1]], function(characters){
  characters$results
})



##### check the number of traits - must be an issue with naming
uncertainty <- function(likelihoods){
  unlist(lapply(likelihoods, function(character){
  sums <- matrix()
  for(row in 1:nrow(character)) {
    sums[[row]] <- 1- max(character[row, ]) 
  }
  return(sum(sums) / nrow(character))
}))
}

uncertainties <- lapply(fossil_anc, lapply, lapply,  function(ace){
  x <- ace$discrete$estimates[[1]]
  likelihoods <- lapply(x, function(characters){
    characters$results
  })
  output <- uncertainty(likelihoods)
  return(output)
})


test <- lapply(fossil_anc$slow[[1]]$living$discrete$estimates[[1]], function(characters){
  characters$results
})

discrete_matrices$slow[[1]]





error <- function(likelihoods, true_matrix){
  unlist(lapply(likelihoods, function(character){
  sums <- matrix()
  for(row in 1:nrow(character)) {
  }
  return(sum(sums) / nrow(character))
}))
}


discrete_matrices$slow[[1]][,1]


likelihoods <- fossil_anc$slow[[1]]$living$discrete$estimates[[1]][[1]]$results


rownames(fossil_anc$slow[[1]]$living$discrete$estimates[[1]][[1]]$results)


rownames(fossil_anc$slow[[1]]$living$discrete$estimates[[1]][[1]]$results)



true_states <- discrete_matrices$slow[[1]][rownames(fossil_anc$slow[[1]]$living$discrete$estimates[[1]][[1]]$results),1]

errors <- c()

for(node in 1:length(true_states)){
  true <- true_states[node]
  estimated <- likelihoods[node, ]
  
  # Get likelihood assigned to true state
  true_likelihood <- estimated[as.character(true)]
  
  # Calculate error (1 - probability of correct state)
  error <- 1 - true_likelihood
  errors <- c(errors, error)
}

living_nodes <- lapply(labels, lapply, function(x){
  x$living[grepl("^n", x$living)]
})



error <- function(anc, true, node_labels){
  level <- Map(function(rate_anc, rate_true, rate_nodes){
    Map(function(i_anc, i_true, i_nodes){
      lapply(i_anc, function(fossil){
        traits <- fossil$discrete$estimates[[1]]
        all_errors <- c()  # Single vector of all errors
        
        for(trait in 1:length(traits)){
          likelihoods <- traits[[trait]]$results
          true_trait <- i_true[, trait]
          true_living_nodes <- true_trait[i_nodes]
          
          for(node in 1:length(true_living_nodes)){
            true_state <- true_living_nodes[node]
            estimated <- likelihoods[node, , drop = FALSE]
            
            if(as.character(true_state) %in% colnames(estimated)) {
              true_likelihood <- estimated[1, as.character(true_state)]
              error <- 1 - true_likelihood
            } else {
              error <- NA
            }
            
            all_errors <- c(all_errors, error)
          }
        }
        
        # Return single unlisted vector
        return(all_errors)
      })
    }, rate_anc, rate_true, rate_nodes)
  }, anc, true, node_labels)
  
  return(level)
}


errors <- error(anc = fossil_anc, true = discrete_matrices, node_labels = living_nodes)




discrete_matrices$slow[[1]][living_nodes$slow[[1]],]


saveRDS(uncertainties, "../Data/matrices/discrete/uncertainties.Rds")



```


## Compare true vs ace nodes by distance

```{r}
## this function returns matrices of ace and true node states
get.living.nodes <- function(living) {
  if (inherits(living$living, "list")) {
  matrix <- living$living[[1]]
  } else {
  matrix <- living$living}
  node_v <- rownames(matrix)[grepl("^n", rownames(matrix))]
  return(node_v)
}
## what i need this function to do is get the mrca of the living tree, then keep all the taxa in the other fossil preservation levels that descend from that tip. so that it keeps the fossil sampling of tips descending from the most basal node of the living tree. thus keeps a fair comparison of tips. so n of nodes stays the same.
get.living <- function(living) {
  if (inherits(living$living, "list")) {
  matrix <- living$living[[1]]
  } else {
  matrix <- living$living}
  node_v <- rownames(matrix)
  return(node_v)
}



test_living <- fossil_matrices$slow[[1]]$living$tree
test_living_node <- test_living$node.label[1]
descendents <- lapply(fossil_matrices$slow[[1]], function(x){
  tree <- x$tree
  descendents <- extract.clade(tree, node = test_living_node)
  tip_labels <-  tree$tip.label
  return(tip_labels)
})


descendents <- (extract.clade(fossil_matrices$slow[[1]]$fossil_low$tree, node = test_living_node))$tip.label



extract.living <- function(fossils) {
  basal_node <- fossils$living$tree$node.label[1]
  living_nodes <- fossils$living$tree$node.label
  labels <- lapply(fossils, function(level){
    tree <- level$tree
    descendents <- (extract.clade(tree, node = basal_node))$tip.label
    add_nodes <- c(descendents, living_nodes)
  })
  return(labels)
}

labels <- lapply(fossil_matrices, lapply, extract.living)


strict_living <- Map(function(rate_anc, rate_labels) {
  Map(function(i_anc, i_labels) {
    Map(function(fossil_anc, label_anc) {
      fossil_anc[label_anc, , drop = FALSE]
    }, i_anc, i_labels)
  }, rate_anc, rate_labels)
}, strict_fossil_anc, labels)

names(strict_living) <- names(strict_fossil_anc)

rel_living <- Map(function(rate_anc, rate_labels) {
  Map(function(i_anc, i_labels) {
    Map(function(fossil_anc, label_anc) {
      fossil_anc[label_anc, , drop = FALSE]
    }, i_anc, i_labels)
  }, rate_anc, rate_labels)
}, relative_fossil_anc, labels)

names(rel_living) <- names(relative_fossil_anc)

sample_living <- Map(function(rate_anc, rate_labels) {
  Map(function(i_anc, i_labels) {
    Map(function(fossil_anc, label_anc) {
      lapply(fossil_anc, function(rep) {
        rep[label_anc, , drop = FALSE]
      })
    }, i_anc, i_labels)
  }, rate_anc, rate_labels)
}, sample_fossil_anc, labels)

names(sample_living) <- names(sample_fossil_anc)

no_ace_living <- lapply(strict_living, lapply, lapply, function(matrix){
  no_node <- matrix[!grepl("^n", rownames(matrix)),]
})

true_living <-  lapply(seq_along(discrete_matrices), function(rate) {
  lapply(seq_along(discrete_matrices[[rate]]), function(i) {
    labs <- labels[[rate]][[i]]$all
    true_living <- discrete_matrices[[rate]][[i]][labs, , drop = FALSE]
    })
})
names(true_living) <- names(discrete_matrices)



source("../Functions/node.dist.R")

point_dist_strict <- lapply(point_living_nodes, lapply, node.dist, distance = "strict")

thresh_dist <- lapply(thresh_living_nodes, lapply, node.dist, distance = "strict")

root_dist_point <- lapply(point_dist_strict, lapply, lapply, function(matrix) {
  v <- matrix[1, ]
  mean(v)})

root_dist_thresh <- lapply(thresh_dist, lapply, lapply, function(matrix) {
  v <- matrix[1, ]
  mean(v)})


sample_dist_strict <- lapply(sample_living_nodes, lapply, node.dist, distance = "strict")

# root_dist_sample <- lapply(sample_dist_strict, function(rate){
#   lapply(rate, function(matrix){
#     roots <- lapply(matrix, function(level){
#       # Calculate the mean of all values in the level
#       all_values <- unlist(level, recursive = TRUE)
#       mean(all_values)
#     })
#     return(roots)
#   })
# })

root_dist_sample <- lapply(sample_dist_strict, function(rate){
  lapply(rate, function(matrix){
    roots <- lapply(matrix, function(level){
      # For each replicate in the level, get the first row and calculate mean
      root_values <- sapply(level, function(mat) {
        v <- mat[1, ]  # Get the first row (root)
        mean(v)        # Calculate mean of that row
      })
      # Return the mean of all root means for this level
      mean(root_values)
    })
    return(roots)
  })
})

saveRDS(root_dist_sample, "../Data/matrices/rootdistsamp.rds")
saveRDS(root_dist_point, "../Data/matrices/rootdistpoint.rds")
saveRDS(root_dist_thresh, "../Data/matrices/rootdistthresh.rds")





# dist_strict_sample <- lapply(sample_living_nodes, function(rate) {
# lapply(rate, function(matrix) {
#   lapply(matrix, function(pres) {
#     distances <- node.dist(pres, distance = "strict")
#   })
# })
# })

strict_means_sample <- lapply(sample_dist_strict, function(preservation) {
  lapply(preservation, function(level) {
    lapply(level, function(matrices) {
      # Compute the mean of each replicate
      replicate_means <- sapply(matrices, mean)
      
      # Compute the mean of the replicate means
      # overall_mean <- mean(replicate_means)
      
      return(replicate_means)
    })
  })
})


strict_means_point <- lapply(point_dist_strict, function(preservation) {
  lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    means <- mean(matrices)
  })
})
})

strict_means_thresh <- lapply(thresh_dist, function(preservation) {
  lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    means <- mean(matrices)
  })
})
})

## do i need to divide by number of rows?

lenient_means <- lapply(dist_lenient, function(preservation) {
  lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    means <- mean(matrices)
  })
})
})

saveRDS(lenient_means, "../Data/matrices/lenientmeans.rds")
saveRDS(strict_means_point, "../Data/matrices/pointstrictmeans.rds")
saveRDS(strict_means_sample, "../Data/matrices/samplestrictmeans.rds")
saveRDS(strict_means_thresh, "../Data/matrices/threshstrictmeans.rds")



```



## Disparity diffs of just living tips/nodes

```{R}
source("../Functions/utility.R")

## these are anc estimated matrices with relevant tips but only with relevant nodes.
# anc_matrices_living
## sum of variances
disp_sample <- lapply(sample_living, lapply, lapply, lapply, function(rep){
  dist <- char.diff(rep, method = "mord", by.col = FALSE)
  ord <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp <- dispRity(ord, metric = c(sum, variances))$disparity
})

disp_rel <- lapply(rel_living, lapply, lapply, function(rep){
  dist <- char.diff(rep, method = "mord", by.col = FALSE)
  ord <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp <- dispRity(ord, metric = c(sum, variances))$disparity
})

disp_strict <- lapply(strict_living, lapply, lapply, function(rep){
  dist <- char.diff(rep, method = "mord", by.col = FALSE)
  ord <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp <- dispRity(ord, metric = c(sum, variances))$disparity
})

disp_true <- lapply(true_living, lapply, function(rep){
  dist <- char.diff(rep, method = "mord", by.col = FALSE)
  ord <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp <- dispRity(ord, metric = c(sum, variances))$disparity
})

disp_no_ace <- lapply(no_ace_living, lapply, lapply, function(rep){
  dist <- char.diff(rep, method = "mord", by.col = FALSE)
  ord <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp <- dispRity(ord, metric = c(sum, variances))$disparity
})


disp_rel_diff <- Map(function(rate_data, true_data) {
  Map(function(rep_data, true_rep) {
    Map(function(fossil_data) {
      disp.diff(fossil_data, true_rep)
    }, rep_data)
  }, rate_data, true_data)
}, disp_rel, disp_true)



disp_no_ace_diff <- Map(function(rate_data, true_data) {
  Map(function(rep_data, true_rep) {
    Map(function(fossil_data) {
      disp.diff(fossil_data, true_rep)
    }, rep_data)
  }, rate_data, true_data)
}, disp_no_ace, disp_true)

disp_sample_diff <- Map(function(rate_data, true_data) {
  Map(function(rep_data, true_rep) {
    Map(function(fossil_data) {
      diffs <- Map(function(sample) {
     (disp.diff(sample, true_rep))
      }, fossil_data)
      return(unlist(diffs))
    }, rep_data)
  }, rate_data, true_data)
}, disp_sample, disp_true)

disp_strict_diff <- Map(function(rate_data, true_data) {
  Map(function(rep_data, true_rep) {
    Map(function(fossil_data) {
      disp.diff(fossil_data, true_rep)
    }, rep_data)
  }, rate_data, true_data)
}, disp_strict, disp_true)

saveRDS(disp_strict_diff, "../Data/matrices/discrete/strictvardiffs.Rds")
saveRDS(disp_sample_diff, "../Data/matrices/discrete/samplevardiffs.Rds")
saveRDS(disp_rel_diff, "../Data/matrices/discrete/relvardiffs.Rds")
saveRDS(disp_no_ace_diff, "../Data/matrices/discrete/noacevardiffs.Rds")





########################################################################################################################

neighbours_sample <- lapply(sample_living, lapply, lapply, lapply, function(rep){
  dist <- char.diff(rep, method = "mord", by.col = FALSE)
  ord <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp <- dispRity(ord, metric = c(mean, neighbours))$disparity
})

neighbours_rel <- lapply(rel_living, lapply, lapply, function(rep){
  dist <- char.diff(rep, method = "mord", by.col = FALSE)
  ord <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp <- dispRity(ord, metric = c(mean, neighbours))$disparity
})

neighbours_strict <- lapply(strict_living, lapply, lapply, function(rep){
  dist <- char.diff(rep, method = "mord", by.col = FALSE)
  ord <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp <- dispRity(ord, metric = c(mean, neighbours))$disparity
})

neighbours_true <- lapply(true_living, lapply, function(rep){
  dist <- char.diff(rep, method = "mord", by.col = FALSE)
  ord <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp <- dispRity(ord, metric = c(mean, neighbours))$disparity
})

neighbours_no_ace <- lapply(no_ace_living, lapply, lapply, function(rep){
  dist <- char.diff(rep, method = "mord", by.col = FALSE)
  ord <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp <- dispRity(ord, metric = c(mean, neighbours))$disparity
})

## Get diffs in aced vs true

neighbours_rel_diff <- Map(function(rate_data, true_data) {
  Map(function(rep_data, true_rep) {
    Map(function(fossil_data) {
      disp.diff(fossil_data, true_rep)
    }, rep_data)
  }, rate_data, true_data)
}, neighbours_rel, neighbours_true)


neighbours_sample_diff <- Map(function(rate_data, true_data) {
  Map(function(rep_data, true_rep) {
    Map(function(fossil_data) {
      diffs <- Map(function(sample) {
     (disp.diff(sample, true_rep))
      }, fossil_data)
      return(unlist(diffs))
    }, rep_data)
  }, rate_data, true_data)
}, neighbours_sample, neighbours_true)

neighbours_strict_diff <- Map(function(rate_data, true_data) {
  Map(function(rep_data, true_rep) {
    Map(function(fossil_data) {
      disp.diff(fossil_data, true_rep)
    }, rep_data)
  }, rate_data, true_data)
}, neighbours_strict, neighbours_true)

neighbours_no_ace_diff <- Map(function(rate_data, true_data) {
  Map(function(rep_data, true_rep) {
    Map(function(fossil_data) {
      disp.diff(fossil_data, true_rep)
    }, rep_data)
  }, rate_data, true_data)
}, neighbours_no_ace, neighbours_true)


saveRDS(neighbours_rel_diff, "../Data/matrices/discrete/neighbours.rel.diff.Rds")
saveRDS(neighbours_sample_diff, "../Data/matrices/discrete/neighbours.sample.diff.Rds")
saveRDS(neighbours_strict_diff, "../Data/matrices/discrete/neighbours.strict.diff.Rds")
saveRDS(neighbours_no_ace_diff, "../Data/matrices/discrete/neighbours.noace.diff.Rds")




########################################################################################################################

position_sample <- lapply(sample_living, lapply, lapply, lapply, function(rep){
  dist <- char.diff(rep, method = "mord", by.col = FALSE)
  ord <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp <- dispRity(ord, metric = c(mean, displacements))$disparity
})

position_rel <- lapply(rel_living, lapply, lapply, function(rep){
  dist <- char.diff(rep, method = "mord", by.col = FALSE)
  ord <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp <- dispRity(ord, metric = c(mean, displacements))$disparity
})

position_strict <- lapply(strict_living, lapply, lapply, function(rep){
  dist <- char.diff(rep, method = "mord", by.col = FALSE)
  ord <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp <- dispRity(ord, metric = c(mean, displacements))$disparity
})

position_true <- lapply(true_living, lapply, function(rep){
  dist <- char.diff(rep, method = "mord", by.col = FALSE)
  ord <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp <- dispRity(ord, metric = c(mean, displacements))$disparity
})

## Get diffs in aced vs true

position_rel_diff <- Map(function(rate_data, true_data) {
  Map(function(rep_data, true_rep) {
    Map(function(fossil_data) {
      disp.diff(fossil_data, true_rep)
    }, rep_data)
  }, rate_data, true_data)
}, position_rel, position_true)


position_sample_diff <- Map(function(rate_data, true_data) {
  Map(function(rep_data, true_rep) {
    Map(function(fossil_data) {
      diffs <- Map(function(sample) {
     (disp.diff(sample, true_rep))
      }, fossil_data)
      return(unlist(diffs))
    }, rep_data)
  }, rate_data, true_data)
}, position_sample, position_true)

position_strict_diff <- Map(function(rate_data, true_data) {
  Map(function(rep_data, true_rep) {
    Map(function(fossil_data) {
      disp.diff(fossil_data, true_rep)
    }, rep_data)
  }, rate_data, true_data)
}, position_strict, position_true)


saveRDS(position_rel_diff, "../Data/matrices/discrete/position.rel.diff.Rds")
saveRDS(position_sample_diff, "../Data/matrices/discrete/position.sample.diff.Rds")
saveRDS(position_strict_diff, "../Data/matrices/discrete/position.strict.diff.Rds")



####################################################################################################
disp_post_ord_ace <- lapply(post_ord_ace_living_nodes, lapply, function(x){
  true <- x$true
  aced <- x$ace
  dist <- char.diff(true, method = "mord", by.col = FALSE)
  ord_true <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp_true <- dispRity(ord_true, metric = c(sum, variances))$disparity
  disp_diff <- lapply(aced, function(fossil){
  disp <- dispRity(fossil, metric = c(sum, variances))$disparity
  diff <- (disp[[1]]$elements[1] - disp_true[[1]]$elements[1]) / disp_true[[1]]$elements[1]
  })
  return(disp_diff)
})

saveRDS(disp_post_ord_ace, "../Data/matrices/discrete/post_ord_ace.Rds")



```


## All nodes and raw distance

This is going to be run including tips

```{r}

rel_all_nodes <- lapply(seq_along(point_fossil_anc), function(rate) {
  lapply(seq_along(point_fossil_anc[[rate]]), function(i) {
    # Step 1: Get the node labels
    # node_v <- node_labels[[i]]

    # Step 2: Get the true matrix for the current replicate
    # true <- discrete_matrices[[rate]][[i]][rownames(i), ]

    # Step 3: Iterate over preservation levels
    levels <- names(point_fossil_anc[[rate]][[i]])
    ace_out <- lapply(levels, function(level_name) {
      # Filter ACE matrix for the current preservation level
      ace <- point_fossil_anc[[rate]][[i]][[level_name]]
      ace_filtered <- ace[grepl("^n", rownames(ace)), ]
      true <- discrete_matrices[[rate]][[i]][rownames(ace_filtered), ]
      return(list(true = true, ace = ace_filtered))
    })

    # Assign names to the ACE output
    names(ace_out) <- levels

    # Return the ACE matrices and the single true matrix
    return(ace_out)
  })
})

# Assign names to the outer list
names(rel_all_nodes) <- names(discrete_matrices)

thresh_all_nodes <- lapply(seq_along(thresh_fossil_anc), function(rate) {
  lapply(seq_along(thresh_fossil_anc[[rate]]), function(i) {
    # Step 1: Get the node labels
    # node_v <- node_labels[[i]]

    # Step 2: Get the true matrix for the current replicate
    # true <- discrete_matrices[[rate]][[i]][rownames(i), ]

    # Step 3: Iterate over preservation levels
    levels <- names(thresh_fossil_anc[[rate]][[i]])
    ace_out <- lapply(levels, function(level_name) {
      # Filter ACE matrix for the current preservation level
      ace <- thresh_fossil_anc[[rate]][[i]][[level_name]]
      ace_filtered <- ace[grepl("^n", rownames(ace)), ]
      true <- discrete_matrices[[rate]][[i]][rownames(ace_filtered), ]
      return(list(true = true, ace = ace_filtered))
    })

    # Assign names to the ACE output
    names(ace_out) <- levels

    # Return the ACE matrices and the single true matrix
    return(ace_out)
  })
})

# Assign names to the outer list
names(thresh_all_nodes) <- names(discrete_matrices)



sample_all_nodes <- lapply(seq_along(sampled_fossil_anc), function(rate) {
  lapply(seq_along(sampled_fossil_anc[[rate]]), function(i) {
    
    # Step 1: Iterate over preservation levels
    levels <- names(sampled_fossil_anc[[rate]][[i]]) # e.g., c("fossil", "living")
    
    ace_out <- lapply(levels, function(level_name) {
      # Get all replicates for this level
      replicates <- sampled_fossil_anc[[rate]][[i]][[level_name]]
      
      # Step 2: Filter each replicate to keep only rows starting with "n"
      filtered <- lapply(seq_along(replicates), function(rep_idx) {
        ace <- replicates[[rep_idx]]
        ace_filtered <- ace[grepl("^n", rownames(ace)), ]
        return(ace_filtered)
      })
      
      # Step 3: Get the true matrix using the same rows as the first filtered replicate
      if (length(filtered) > 0 && nrow(filtered[[1]]) > 0) {
        true <- discrete_matrices[[rate]][[i]][rownames(filtered[[1]]), ]
      } else {
        true <- NULL
      }
      
      return(list(true = true, ace = filtered))
    })
    
    # Assign names to the ACE output
    names(ace_out) <- levels
    
    # Return the ACE matrices and true matrix for each level
    return(ace_out)
  })
})

# Assign names to the outer list
names(sample_all_nodes) <- names(discrete_matrices)







rel_dist <- lapply(rel_all_nodes, lapply, lapply, all.node.dist, distance = "strict")

thresh_dist <- lapply(thresh_all_nodes, lapply, lapply, all.node.dist, distance = "strict")

sample_dist <- lapply(sample_all_nodes, lapply, lapply, all.node.dist, distance = "strict")

allnode_means_rel <- lapply(rel_dist, function(preservation) {
  lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    means <- mean(matrices)
  })
})
})

allnode_means_thresh <- lapply(thresh_dist, function(preservation) {
  lapply(preservation, function(level) {
  lapply(level, function(matrices) {
    means <- mean(matrices)
  })
})
})

allnode_means_sample <- lapply(sample_dist, function(rate_level) {
  lapply(rate_level, function(replicate) {
    lapply(replicate, function(level) {
     q <-  lapply(level, function(x) {
        m <- mean(x)
      })
      unlist <- unlist(q)
      return(unlist)
    })
  })
})


saveRDS(allnode_means_rel, "../Data/matrices/discrete/rel_raw_dist.Rds")
saveRDS(allnode_means_thresh, "../Data/matrices/discrete/thresh_raw_dist.Rds")
saveRDS(allnode_means_sample, "../Data/matrices/discrete/sample_raw_dist.Rds")


```


## Disparity for all nodes and tips



```{r}
discrete_matrices <- readRDS("../Data/matrices/discrete/discretematrices.rds")
relative_fossil_anc <- readRDS("../Data/matrices/discrete/relanc.Rds")
strict_fossil_anc <- readRDS("../Data/matrices/discrete/strictanc.Rds")
sample_fossil_anc <- readRDS("../Data/matrices/discrete/sampleddiscreteanc.Rds")
fossil_matrices <- readRDS("../Data/matrices/discrete/fossilmatrices.Rds")

## add to this - > just tips disparity? so it is true (tips + nodes all) , aced (tips + estimated node states) and just tips (varies by fossil sampling like aced).

# source("../Functions/apply.dist.R")
# source("../Functions/apply.ordination.R")

distances_true <- lapply(discrete_matrices, lapply, function(x){
  matrix <-  apply(x, c(1,2), as.character)
  dist <- char.diff(matrix, method = "mord", by.col = FALSE)
  return(dist)
})

distances_no_ace <- lapply(fossil_matrices, lapply, lapply, function(x){
  matrix <- x$matrix
    dist <- char.diff(matrix, method = "mord", by.col = FALSE)
    return(dist)
})

distances_strict <- lapply(strict_fossil_anc, lapply, lapply, function(matrices) {
  dist <- char.diff(matrices, method = "mord", by.col = FALSE)
  return(dist)
})

distances_rel <- lapply(relative_fossil_anc, lapply, lapply, function(matrices) {
  dist <- char.diff(matrices, method = "mord", by.col = FALSE)
  return(dist)
})




distances_sample <- lapply(sample_fossil_anc, function(rate) {
  mclapply(rate, function(matrix) {
    lapply(matrix, function(fossil){
      lapply(fossil, function(sample) {
        dist <- char.diff(sample, method = "mord", by.col = FALSE)
      })
    })
  }, mc.cores = 10)
})


ord_no_ace <- lapply(distances_no_ace, lapply, lapply, function(matrix){
    ord <-  (cmdscale(matrix, k = ncol(matrix) - 2, add = TRUE))$points
    return(ord)
})

ord_true <- lapply(distances_true, lapply, function(matrix){
    ord <-  (cmdscale(matrix, k = ncol(matrix) - 2, add = TRUE))$points
    return(ord)
})


# ord_sample <- lapply(distances_sample, function(rate_data) {
#   mclapply(rate_data, function(matrices) {
#     true_ord <- (cmdscale(matrices$true, k = ncol(matrices$true) - 2, add = TRUE))$points
#     ace <- matrices$ace
#     ace_ord <- lapply(ace, function(pres) {
#       lapply(pres, function(reps) {
#         ord <- (cmdscale(reps, k = ncol(reps) - 2, add = TRUE))$points
#       })
#     })
#   return(list(true = true_ord, ace = ace_ord ))
#   }, mc.cores = 10)
# })


# Limit to first 20 dimensions to reduce memory usage
# Process and save each rate separately to prevent memory buildup
for(rate_name in names(distances_sample)) {
  cat("Processing rate:", rate_name, "\n")
  
  rate_ord <- list()
  
  for(i in seq_along(distances_sample[[rate_name]])) {
    cat("  Replicate", i, "of", length(distances_sample[[rate_name]]), "\n")
    
    matrices <- distances_sample[[rate_name]][[i]]
    
    # Process this single replicate
    ord_result <- lapply(matrices, function(fossil) {
      lapply(fossil, function(sample) {
        ord <- (cmdscale(sample, k = ncol(sample) - 2, add = TRUE))$points
      })
    })
    
    rate_ord[[i]] <- ord_result
    
    # Cleanup every 10 replicates
    if(i %% 10 == 0) {
      gc()
      cat("    Memory cleanup at replicate", i, "\n")
    }
  }
  
  # Save this rate immediately
  saveRDS(rate_ord, paste0("../Data/matrices/discrete/ord_sample_", rate_name, ".rds"))
  
  # Clean up
  rm(rate_ord)
  gc()
}

ord_sample <- list(fast = readRDS("../Data/matrices/discrete/ord_sample_fast.rds"), med = readRDS("../Data/matrices/discrete/ord_sample_med.rds"), slow = readRDS("../Data/matrices/discrete/ord_sample_slow.rds"))
### is it to do with the fact that they are ordinated at different dimensions??? should they all be ordinated at a set number of dimensions

## i've decided to keep the first 50?


## Why then 


sum_var_no_ace <- lapply(ord_no_ace, lapply, lapply, function(matrix){
    var <-  (dispRity(matrix[ ,1:48], metric = c(sum, variances)))$disparity
    return(var)
})


sum_var_true <- lapply(ord_true, lapply, function(matrix){
    var <-  (dispRity(matrix[ ,1:48], metric = c(sum, variances)))$disparity
    return(var)
})

sum_var_rel <- lapply(ord_rel, lapply, function(matrices) {
  true_disp <- (dispRity(matrices$true[, 1:48], metric = c(sum, variances)))$disparity
  ace <- matrices$ace
  ace_disp <- lapply(ace, function(pres) {
    disp <- (dispRity(pres[ ,1:48], metric = c(sum, variances)))$disparity
  })
  return(list(true = true_disp, ace = ace_disp ))
})


sum_var_strict <- lapply(ord_strict, lapply, function(matrices) {
  true_disp <- (dispRity(matrices$true[ ,1:48], metric = c(sum, variances)))$disparity
  ace <- matrices$ace
  ace_disp <- lapply(ace, function(pres) {
    disp <- (dispRity(pres[ ,1:48], metric = c(sum, variances)))$disparity
  })
  return(list(true = true_disp, ace = ace_disp ))
})

sum_var_sample <- lapply(ord_sample, function(matrices) {
  lapply(matrices, function(pres) {
    lapply(pres, function(reps) {
        # lapply(reps, function(matrix){
        (dispRity(reps, metric = c(sum, variances)))$disparity
      })
    })
})


sum_var_sample <- lapply(ord_sample, lapply, lapply, dispRity, metric = c(sum, variances))

var_rel_diffs <- lapply(sum_var_rel, lapply, function(disparity){
  true <- disparity$true[[1]]$elements[1]
  ace <- disparity$ace
  diffs <- lapply(ace, function(pres){
    diff <- (pres[[1]]$elements[1] - true) / true
  })
  return(diffs)
})

var_strict_diffs <- lapply(sum_var_strict, lapply, function(disparity){
  true <- disparity$true[[1]]$elements[1]
  ace <- disparity$ace
  diffs <- lapply(ace, function(pres){
    diff <- (pres[[1]]$elements[1] - true) / true
  })
  return(diffs)
})

var_sample_diffs <- lapply(sum_var_sample, lapply, function(disparity) {
  true <- disparity$true[[1]]$elements[1]
  ace <- disparity$ace
  diffs <- lapply(ace, function(pres) {
    lapply(pres, function(rep) {
      # Compute the difference
      diff <- (rep[[1]]$elements[1] - true) / true
      return(diff)
    })
  })
  
  # Apply unlist to the lowest level
  diffs <- lapply(diffs, function(pres) {
    unlist(pres, recursive = TRUE)
  })
  
  return(diffs)
})




```

## Post-ord ace

```{r}
# load the discrete matrices
discrete_matrices <- readRDS("../Data/matrices/discrete/discretematrices.rds")

source("../Functions/fossil.pres.R")

living <- lapply(discrete_matrices, remove.fossil, trees = trees, type = "discrete")
fossilised_high <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.5, type = "discrete")
all_fossil <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 1.0, type = "discrete")
fossilised_med <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.15, type = "discrete")
fossilised_low <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.05, type = "discrete")





# plot(fossilised_low$slow[[13]]$tree)

fossil_matrices <- lapply(names(discrete_matrices), function(level) {
  lapply(seq_along(discrete_matrices[[level]]), function(i) {
    list(
      all = all_fossil[[level]][[i]],
      fossil_high = fossilised_high[[level]][[i]],
      fossil_med = fossilised_med[[level]][[i]],
      fossil_low = fossilised_low[[level]][[i]],
      living = living[[level]][[i]]
    )
  })
})

# Assign names to the outer list
names(fossil_matrices) <- names(discrete_matrices)
## Firstly, find the overall disparity




distances <- lapply(fossil_matrices, lapply, lapply, function(x) {
  matrix <- x$matrix
  tree <- x$tree
  distance_matrix <- char.diff(matrix, method = "mord", by.col = FALSE)
  return(list(tree = tree, matrix = distance_matrix))
})


ordinations <- lapply(distances, lapply, lapply, function(x) {
  matrix <- x$matrix
  tree <- x$tree
  ord_matrix <- (cmdscale(matrix, k = ncol(matrix) - 2, add = TRUE))$points
  return(list(tree = tree, matrix = ord_matrix))
})

n_cores <- 11
anc.states <- function(x){ 
  anc_states <- multi.ace(data = x$matrix[,1:48], 
                          tree = x$tree, 
                          models = "BM", 
                          output = "multi.ace",
                          parallel = n_cores,
                          verbose = TRUE)
  return(anc_states)
}



post_ord_ace <- lapply(ordinations, lapply, lapply, anc.states)

saveRDS(post_ord_ace, "../Data/matrices/discrete/postordace.Rds")
post_ord_ace <-  readRDS("../Data/matrices/discrete/postordace.Rds")

point_post_ord <- lapply(post_ord_ace, lapply, lapply, multi.ace, output = "combined.matrix" )
sample_post_ord <- lapply(post_ord_ace, lapply, lapply, multi.ace, output = "combined.matrix", sample = 10, sample.fun =  trait_normal)


trait_normal = list(fun = rnorm, param = list(mean = mean, sd = function(x)return(diff(range(x))/4)))

sum_var_post_ord <- lapply(point_post_ord, lapply, lapply, function(matrix){
 dispRity(matrix, metric = c(sum, variances))$disparity})


post_ord_ace_living_nodes <- lapply(seq_along(point_post_ord), function(rate) {
  lapply(seq_along(point_post_ord[[rate]]), function(i) {
    # Step 1: Get the node labels
    node_v <- node_labels[[i]]

    # Step 2: Get the true matrix for the current replicate
    true <- discrete_matrices[[rate]][[i]][node_v, ]

    # Step 3: Iterate over preservation levels
    levels <- names(point_post_ord[[rate]][[i]])
    ace_out <- lapply(levels, function(level_name) {
      # Filter ACE matrix for the current preservation level
      point_post_ord[[rate]][[i]][[level_name]][node_v, , drop = FALSE]
    })

    # Assign names to the ACE output
    names(ace_out) <- levels

    # Return the ACE matrices and the single true matrix
    return(list(ace = ace_out, true = true))
  })
})

# Assign names to the outer list
names(post_ord_ace_living_nodes) <- names(discrete_matrices)


disp_post_ord_ace <- lapply(post_ord_ace_living_nodes, lapply, function(x){
  true <- x$true
  aced <- x$ace
  dist <- char.diff(true, method = "mord", by.col = FALSE)
  ord_true <- (cmdscale(dist, k = ncol(dist) - 2, add = TRUE))$points
  disp_true <- dispRity(ord_true, metric = c(sum, variances))$disparity
  disp_diff <- lapply(aced, function(fossil){
  disp <- dispRity(fossil, metric = c(sum, variances))$disparity
  diff <- (disp[[1]]$elements[1] - disp_true[[1]]$elements[1]) / disp_true[[1]]$elements[1]
  })
  return(disp_diff)
})

saveRDS(disp_post_ord_ace, "../Data/matrices/discrete/post_ord_ace.Rds")


sum_var_post_ord <- lapply(point_post_ord, lapply, lapply, function(matrix){
 dispRity(matrix, metric = c(sum, variances))$disparity})


```



## No ACE

```{R}


source("../Functions/fossil.pres.R")

living <- lapply(discrete_matrices, remove.fossil, trees = trees, type = "discrete")
fossilised_high <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.5, type = "discrete")
all_fossil <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 1.0, type = "discrete")
fossilised_med <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.15, type = "discrete")
fossilised_low <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.05, type = "discrete")





# plot(fossilised_low$slow[[13]]$tree)

fossil_matrices <- lapply(names(discrete_matrices), function(level) {
  lapply(seq_along(discrete_matrices[[level]]), function(i) {
    list(
      all = all_fossil[[level]][[i]],
      fossil_high = fossilised_high[[level]][[i]],
      fossil_med = fossilised_med[[level]][[i]],
      fossil_low = fossilised_low[[level]][[i]],
      living = living[[level]][[i]]
    )
  })
})

# Assign names to the outer list
names(fossil_matrices) <- names(discrete_matrices)


## Firstly, find the overall disparity

no_ace_matrices <- lapply(fossil_matrices, lapply, lapply, function(x){
  x$matrix
})




```
