---
title: "Discrete with extinction"
author: "Caleb"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

## Load trees


```{r}

library(treats)
bd_params <- make.bd.params(speciation = 0.1, extinction = 0.07)

stop_rule <- list(max.living = 30)
set.seed(123)

mass_extinction <- make.events(
                      target       = "taxa",
                      condition    = taxa.condition(29),
                      modification = random.extinction(0.80))

trees <- treats(stop.rule = stop_rule, bd.params = bd_params, null.error = 100, replicates = 100, events = mass_extinction)
trees <- lapply(trees, drop.singles)

fix.zero.branches <- function(tree, min_length = 1e-8) {
  tree$edge.length[tree$edge.length <= 0] <- min_length
  return(tree)
}

trees <- lapply(trees, fix.zero.branches)

write.tree(trees, "../Data/trees/randomextinction.tre")
trees <- read.tree("../Data/trees/randomextinction.tre")

extinction_ages <- lapply(trees, function(tree){
  ages <- tree.age(tree)
  tips <- grepl("^t", ages$elements)  # Logical vector for rows with tip names
  tip_ages <- ages[tips, ]  # Subset ROWS where elements start with "t"
  return(tip_ages)
})


times <- lapply(extinction_ages, find_extinction_time)


# trees <- read.tree("../Data/trees/simulated.tre")


slow_binary_transitions <- matrix(c(
    0.99, 0.01,
    0.01, 0.99
), nrow = 2, byrow = TRUE)

slow_multi_transitions <- matrix(c(
    0.99, 0.025, 0.025,
    0.025, 0.99, 0.025,
    0.025, 0.025, 0.99
), nrow = 3, byrow = TRUE)

slow_binary <- treats::make.traits(process = discrete.process, n = 85, process.args = list(transitions = slow_binary_transitions))
slow_multi <- treats::make.traits(process = discrete.process, n = 15, process.args = list(transitions = slow_multi_transitions))


med_binary_transitions <- matrix(c(
  0.90, 0.1,
  0.1, 0.9
), nrow = 2, byrow = TRUE)


med_multi_transitions <- matrix(c(
  0.9, 0.05, 0.05,
  0.05, 0.9, 0.05,
  0.05, 0.05, 0.9
), nrow = 3, byrow = TRUE)

med_binary <- treats::make.traits(process = discrete.process, n = 85, process.args = list(transitions = med_binary_transitions))
med_multi <- treats::make.traits(process = discrete.process, n = 15, process.args = list(transitions = med_multi_transitions))

fast_binary_transitions <- matrix(c(
    0.1, 0.9,
    0.9, 0.1
), nrow = 2, byrow = TRUE)

fast_multi_transitions <- matrix(c(
    0.1, 0.45, 0.45, 
    0.45, 0.1, 0.45,
    0.45, 0.45, 0.1 
), nrow = 3, byrow = TRUE)

fast_binary <- treats::make.traits(process = discrete.process, n = 85, process.args = list(transitions = fast_binary_transitions))
fast_multi <- treats::make.traits(process = discrete.process, n = 15, process.args = list(transitions = fast_multi_transitions))

traits <- list(fast = fast_binary, med = med_binary, slow = slow_binary)



## map the traits



slow_matrices <- lapply(trees, function(tree) {
    mapped_binary <- (map.traits(slow_binary, tree))$data
    mapped_multi <- (map.traits(slow_multi, tree))$data
    matrix <- cbind(mapped_binary, mapped_multi)
    return(matrix)
})

med_matrices <- lapply(trees, function(tree) {
    mapped_binary <- (map.traits(med_binary, tree))$data
    mapped_multi <- (map.traits(med_multi, tree))$data
    matrix <- cbind(mapped_binary, mapped_multi)
    return(matrix)
})

fast_matrices <- lapply(trees, function(tree) {
    mapped_binary <- (map.traits(fast_binary, tree))$data
    mapped_multi <- (map.traits(fast_multi, tree))$data
    matrix <- cbind(mapped_binary, mapped_multi)
    return(matrix)
})


discrete_matrices <- list(slow = slow_matrices, med = med_matrices, fast = fast_matrices)
saveRDS(discrete_matrices, "../Data/matrices/randomExtinction/discretematrices.rds")

discrete_matrices <- readRDS("../Data/matrices/randomExtinction/discretematrices.rds")


```





## Different tree coverages

```{r}

## fossil degredation as well

source("../Functions/fossil.pres.R")

living <- lapply(discrete_matrices, remove.fossil, trees = trees, type = "discrete")
fossilised_high <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.5, type = "discrete")
all_fossil <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 1.0, type = "discrete")
fossilised_med <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.15, type = "discrete")
fossilised_low <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.05, type = "discrete")



# plot(fossilised_low$slow[[13]]$tree)

fossil_matrices <- lapply(names(discrete_matrices), function(level) {
  lapply(seq_along(discrete_matrices[[level]]), function(i) {
    list(
      all = all_fossil[[level]][[i]],
      fossil_high = fossilised_high[[level]][[i]],
      fossil_med = fossilised_med[[level]][[i]],
      fossil_low = fossilised_low[[level]][[i]],
      living = living[[level]][[i]]
    )
  })
})

# Assign names to the outer list
names(fossil_matrices) <- names(discrete_matrices)

saveRDS(fossil_matrices, "../Data/matrices/randomExtinction/fossilmatrices.Rds")
fossil_matrices <-  readRDS("../Data/matrices/randomExtinction/fossilmatrices.Rds")
fossil_trees <- lapply(fossil_matrices, lapply, lapply, function(level){
  tree <- level$tree
})


```


## ACE
```{r}

n_cores  <- 11
anc.states <- function(fossil_matrices) {
  lapply(fossil_matrices, function(level) {
    lapply(level, function(x) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = x$matrix, 
                          tree = x$tree, 
                          models = "SYM", 
                          output = "multi.ace",
                          verbose = TRUE, 
                          parallel = n_cores)
  return(anc_states)})})
}


fossil_anc <- lapply(fossil_matrices, anc.states)

saveRDS(fossil_anc, "../Data/matrices/randomExtinction/fossilanc.Rds")
fossil_anc <- readRDS("../Data/matrices/randomExtinction/fossilanc.Rds")


sample_fossil_anc <- lapply(fossil_anc, lapply, lapply, multi.ace, sample = 10, sample.fun = normal_sampling,  verbose = TRUE)


strict_fossil_anc <- lapply(fossil_anc, lapply, lapply, multi.ace, threshold = FALSE, output = "combined.matrix", verbose = TRUE)

relative_fossil_anc <- lapply(fossil_anc, lapply, lapply, multi.ace, output = "combined.matrix", verbose = TRUE)


saveRDS(sample_fossil_anc, "../Data/matrices/randomExtinction/sampleddiscreteanc.Rds")
saveRDS(relative_fossil_anc, "../Data/matrices/randomExtinction/relanc.Rds")
saveRDS(strict_fossil_anc, "../Data/matrices/randomExtinction/strictanc.Rds")

sample_fossil_anc <- readRDS("../Data/matrices/randomExtinction/sampleddiscreteanc.Rds")
relative_fossil_anc <- readRDS("../Data/matrices/randomExtinction/relanc.Rds")
strict_fossil_anc  <- readRDS("../Data/matrices/randomExtinction/strictanc.Rds")



```



## Disparity random extinction


```{r}






# source("../Functions/apply.dist.R")
# source("../Functions/apply.ordination.R")

distances_true <- lapply(discrete_matrices, lapply, function(x){
  matrix <-  apply(x, c(1,2), as.character)
  dist <- char.diff(matrix, method = "mord", by.col = FALSE)
  return(dist)
})

distances_no_ace <- lapply(fossil_matrices, lapply, lapply, function(x){
  matrix <- x$matrix
    dist <- char.diff(matrix, method = "mord", by.col = FALSE)
    return(dist)
})

distances_strict <- lapply(strict_fossil_anc, lapply, lapply, function(matrices) {
  dist <- char.diff(matrices, method = "mord", by.col = FALSE)
  return(dist)
})

distances_rel <- lapply(relative_fossil_anc, lapply, lapply, function(matrices) {
  dist <- char.diff(matrices, method = "mord", by.col = FALSE)
  return(dist)
})


distances_sample <- lapply(sample_fossil_anc, function(rate) {
  mclapply(rate, function(matrix) {
    lapply(matrix, function(fossil){
      lapply(fossil, function(sample) {
        dist <- char.diff(sample, method = "mord", by.col = FALSE)
      })
    })
  }, mc.cores = 10)
})


ord_no_ace <- lapply(distances_no_ace, lapply, lapply, function(matrix){
    ord <-  (cmdscale(matrix, k = ncol(matrix) - 2, add = TRUE))$points
    return(ord)
})

ord_true <- lapply(distances_true, lapply, function(matrix){
    ord <-  (cmdscale(matrix, k = ncol(matrix) - 2, add = TRUE))$points
    return(ord)
})


ord_rel <- lapply(distances_rel, lapply, lapply, function(matrix) {
 cmdscale(matrix, k = ncol(matrix) - 2, add = TRUE)$points
})


ord_strict <- lapply(distances_strict, lapply, lapply, function(matrix) {
 cmdscale(matrix, k = ncol(matrix) - 2, add = TRUE)$points
})

# ord_sample <- lapply(distances_sample, function(rate_data) {
#   mclapply(rate_data, function(matrices) {
#     true_ord <- (cmdscale(matrices$true, k = ncol(matrices$true) - 2, add = TRUE))$points
#     ace <- matrices$ace
#     ace_ord <- lapply(ace, function(pres) {
#       lapply(pres, function(reps) {
#         ord <- (cmdscale(reps, k = ncol(reps) - 2, add = TRUE))$points
#       })
#     })
#   return(list(true = true_ord, ace = ace_ord ))
#   }, mc.cores = 10)
# })


# Limit to first 20 dimensions to reduce memory usage
# Process and save each rate separately to prevent memory buildup
for(rate_name in names(distances_sample)) {
  cat("Processing rate:", rate_name, "\n")
  
  rate_ord <- list()
  
  for(i in seq_along(distances_sample[[rate_name]])) {
    cat("  Replicate", i, "of", length(distances_sample[[rate_name]]), "\n")
    
    matrices <- distances_sample[[rate_name]][[i]]
    
    # Process this single replicate
    ord_result <- lapply(matrices, function(fossil) {
      lapply(fossil, function(sample) {
        ord <- (cmdscale(sample, k = ncol(sample) - 2, add = TRUE))$points
      })
    })
    
    rate_ord[[i]] <- ord_result
    
    # Cleanup every 10 replicates
    if(i %% 10 == 0) {
      gc()
      cat("    Memory cleanup at replicate", i, "\n")
    }
  }
  
  # Save this rate immediately
  saveRDS(rate_ord, paste0("../Data/matrices/randomExtinction/ord_sample_", rate_name, ".rds"))
  
  # Clean up
  rm(rate_ord)
  gc()
}

ord_sample <- list(fast = readRDS("../Data/matrices/randomExtinction/ord_sample_fast.rds"), med = readRDS("../Data/matrices/randomExtinction/ord_sample_med.rds"), slow = readRDS("../Data/matrices/randomExtinction/ord_sample_slow.rds"))







```

#######################################################################################################################


## selective extinction load package and trait data
```{r}
setwd("/home/caleb/Documents/PhD/treats/treats")  # or wherever you cloned it
devtools::load_all()


# install_github("TGuillerme/treats", ref = "map.traits.events")

slow_binary_transitions <- matrix(c(
    0.99, 0.01,
    0.01, 0.99
), nrow = 2, byrow = TRUE)

med_binary_transitions <- matrix(c(
  0.90, 0.1,
  0.1, 0.9
), nrow = 2, byrow = TRUE)


fast_binary_transitions <- matrix(c(
    0.1, 0.9,
    0.9, 0.1
), nrow = 2, byrow = TRUE)

slow_binary <- treats::make.traits(process = discrete.process, n = 30, process.args = list(transitions = slow_binary_transitions))

med_binary <- treats::make.traits(process = discrete.process, n = 30, process.args = list(transitions = med_binary_transitions))



fast_binary <- treats::make.traits(process = discrete.process, n = 30, process.args = list(transitions = fast_binary_transitions))



slow_multi_transitions <- matrix(c(
    0.99, 0.025, 0.025,
    0.025, 0.99, 0.025,
    0.025, 0.025, 0.99
), nrow = 3, byrow = TRUE)

slow_multi <- treats::make.traits(process = discrete.process, n = 20, process.args = list(transitions = slow_multi_transitions))


med_multi_transitions <- matrix(c(
  0.9, 0.05, 0.05,
  0.05, 0.9, 0.05,
  0.05, 0.05, 0.9
), nrow = 3, byrow = TRUE)

med_multi <- treats::make.traits(process = discrete.process, n = 20, process.args = list(transitions = med_multi_transitions))


fast_multi_transitions <- matrix(c(
    0.1, 0.45, 0.45, 
    0.45, 0.1, 0.45,
    0.45, 0.45, 0.1 
), nrow = 3, byrow = TRUE)

fast_multi <- treats::make.traits(process = discrete.process, n = 20, process.args = list(transitions = fast_multi_transitions))

traits = list(slow = slow_binary, med = med_binary, fast = fast_binary)

multi_traits = list(slow = slow_multi, med = med_multi, fast = fast_multi)

```


## Making correlated discrete traits
```{R}

source("../Functions/correlate.traits.R")
# Step 1: Create the first discrete trait (e.g., 2 states)
# Create correlated traits with adjustable correlation strength

test_correlation <- correlate.traits.realistic(correlation = 0.5, n.traits = 6, transition.matrix = slow_multi_transitions)  


multi_traits <- list(fast = fast_cor_traits, med = med_cor_traits, slow = slow_cor_traits)

```

## Run selective extinction `treats`

```{r}

bd_params <- make.bd.params(speciation = 0.1, extinction = 0.07)

stop_rule <- list(max.living = 80)
set.seed(123)
## so can simulate say 40 characters that are correlated by extinction -> in this way. then on resulting tree, simulate non selective traits with map.traits. can decide what level of traits need to be selected by extinction.
selective_extinction <- make.events(
                      target       = "taxa",
                      condition    = taxa.condition(79),
                      modification = trait.extinction(x = 2, trait = 1,
                                                      condition = `<`, severity = 0.8, threshold = 0.5))

## different selectivities: directional ()

test <- treats(traits = test_correlation, stop.rule = stop_rule, bd.params = bd_params, null.error = 100, events = selective_extinction)

selective <- lapply(multi_traits, function(rate) {
  treats(traits = rate, stop.rule = stop_rule, bd.params = bd_params, null.error = 100, replicates = 10, events = selective_extinction)
})

selective <- lapply(selective, lapply, drop.singles)


# test_distance <- char.diff(trees$slow[[5]]$data, method = "mord", by.col = FALSE)

# test_ord <- cmdscale(test_distance,  k = ncol(test_distance) - 2, add = TRUE)$points

# test_chrono <- chrono.subsets(test_ord, tree = trees$slow[[5]]$tree, method = "d", time = seq(310, 0, by = -20))


# # time_slices <- get.matrix(test_chrono, subsets = 1)


# sliced_matrices <- list()
# for(i in 1:10){
#   sliced_matrices[[i]] <- get.matrix(test_chrono, subsets = i)
# }



# # options(repr.plot.width = 12, repr.plot.height = 8)

# # # Or use dev.new() with specific dimensions
# # dev.new(width = 12, height = 8)

# # Then create your plots
# par(mfrow = c(2, 5))
# for(i in 1:10) {
#   plot(sliced_matrices[[i]][, 1], sliced_matrices[[i]][, 2], 
#        main = paste("Time Slice", i),
#        xlab = "PC1", ylab = "PC2",
#        pch = 16, col = rainbow(10)[i])
# }

# test_disparity <- dispRity(test_chrono, metric = c(sum, variances))
# ## what do i need
# ## so i need a way of the selective extinction only choosing a sample of taxa to go extinct
# ## but also certain trait combinations

```



```{R}
extinction.probability <- function()

intercept <- 0
linear_predictor <- intercept + as.matrix(parent_traits[, seq_along(betas)]) %*% betas
    
    ## Logistic: extinction probability
extinction_prob <- 1 / (1 + exp(-linear_predictor))
    
    ## Draw random 0/1 for each parent node
extinct_flag <- rbinom(length(extinction_prob), size = 1, prob = extinction_prob)
    







```

## Fossil sampling

```{R}

matrices <- lapply(selective, lapply, function(x){
  x$data
})

saveRDS(matrices, "../Data/matrices/selectiveExtinction/matrices.Rds")
matrices <- readRDS("../Data/matrices/selectiveExtinction/matrices.Rds")

trees <- lapply(selective, lapply, function(x){
  x$tree
})

saveRDS(trees, "../Data/trees/selectiveExtinction.tre")
trees <- readRDS("../Data/trees/selectiveExtinction.tre")






source("../Functions/fossil.pres.R")

living <- Map(function(tree, matrix){ 
  remove.fossil(tree, matrix, type = "discrete")}, trees, matrices)


fossilised_high <- Map(function(tree, matrix){ 
  fossil.pres(tree, matrix, preservation = 0.5, type = "discrete")}, trees, matrices)
fossilised_med <- Map(function(tree, matrix){ 
  fossil.pres(tree, matrix, preservation = 0.15, type = "discrete")}, trees, matrices)

fossilised_low <- Map(function(tree, matrix){ 
  fossil.pres(tree, matrix, preservation = 0.05, type = "discrete")}, trees, matrices)

all_fossil <- Map(function(tree, matrix){ 
  fossil.pres(tree, matrix, preservation = 1.0, type = "discrete")}, trees, matrices)


fossil_matrices <- lapply(names(matrices), function(level) {
  lapply(seq_along(matrices[[level]]), function(i) {
    list(
      all = all_fossil[[level]][[i]],
      fossil_high = fossilised_high[[level]][[i]],
      fossil_med = fossilised_med[[level]][[i]],
      fossil_low = fossilised_low[[level]][[i]],
      living = living[[level]][[i]]
    )
  })
})

# Assign names to the outer list
names(fossil_matrices) <- names(matrices)

saveRDS(fossil_matrices, "../Data/matrices/selectiveExtinction/fossilmatrices.Rds")
fossil_matrices <-  readRDS("../Data/matrices/selectiveExtinction/fossilmatrices.Rds")



fossil_trees <- lapply(fossil_matrices, lapply, lapply, function(level){
  tree <- level$tree
})

```


## Estimate ancestral states

```{r}
n_cores  <- 16
anc.states <- function(fossil_matrices) {
  lapply(fossil_matrices, function(level) {
    lapply(level, function(x) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = x$matrix, 
                          tree = x$tree, 
                          models = "SYM", 
                          output = "multi.ace",
                          verbose = TRUE,
                          parallel = n_cores)
  return(anc_states)})})
}
fossil_anc <- lapply(fossil_matrices, anc.states)

saveRDS(fossil_anc, "../Data/matrices/selectiveExtinction/discreteanc.Rds")



sample_fossil_anc <- lapply(fossil_anc, lapply, lapply, multi.ace, sample = 10, sample.fun = normal_sampling,  verbose = TRUE)


strict_fossil_anc <- lapply(fossil_anc, lapply, lapply, multi.ace, threshold = FALSE, output = "combined.matrix", verbose = TRUE)

relative_fossil_anc <- lapply(fossil_anc, lapply, lapply, multi.ace, output = "combined.matrix", verbose = TRUE)

saveRDS(sample_fossil_anc, "../Data/matrices/selectiveExtinction/sampleanc.Rds")
saveRDS(relative_fossil_anc, "../Data/matrices/selectiveExtinction/relanc.Rds")
saveRDS(strict_fossil_anc, "../Data/matrices/selectiveExtinction/strictanc.Rds")


strict_fossil_anc <- readRDS("../Data/matrices/selectiveExtinction/strictanc.Rds")
relative_fossil_anc <- readRDS("../Data/matrices/selectiveExtinction/relanc.Rds")
sample_fossil_anc <- readRDS("../Data/matrices/selectiveExtinction/sampleanc.Rds")


```


```{R}
uncertainty <- function(likelihoods){
  unlist(lapply(likelihoods, function(character){
  sums <- matrix()
  for(row in 1:nrow(character)) {
    sums[[row]] <- 1- max(character[row,]) 
  }
  return(sum(sums) / nrow(character))
}))
}

uncertainties <- lapply(fossil_anc, lapply, lapply,  function(ace){
  x <- ace$discrete$estimates[[1]]
  likelihoods <- lapply(x, function(characters){
    characters$results
  })
  output <- uncertainty(likelihoods)
  return(output)
})



```

## Disparity for living nodes

```{R}

get.living.nodes <- function(living) {
  if (inherits(living$living, "list")) {
  matrix <- living$living[[1]]
  } else {
  matrix <- living$living}
  node_v <- rownames(matrix)[grepl("^n", rownames(matrix))]
  return(node_v)
}

node_v <- lapply(strict_fossil_anc, lapply, get.living.nodes)



rel_living_nodes <- lapply(seq_along(relative_fossil_anc), function(rate) {
  lapply(seq_along(relative_fossil_anc[[rate]]), function(i) {
    # Step 1: Get the node labels
    node_labels <- node_v[[rate]][[i]]

    # Step 3: Iterate over preservation levels
    levels <- names(relative_fossil_anc[[rate]][[i]])
    ace_out <- lapply(levels, function(level_name) {
      # Filter ACE matrix for the current preservation level
      relative_fossil_anc[[rate]][[i]][[level_name]][node_labels, , drop = FALSE]
    })

    # Assign names to the ACE output
    names(ace_out) <- levels

    # Return the ACE matrices and the single true matrix
    return(ace_out)
  })
})

# Assign names to the outer list
names(rel_living_nodes) <- names(matrices)



strict_living_nodes <- lapply(seq_along(strict_fossil_anc), function(rate) {
  lapply(seq_along(strict_fossil_anc[[rate]]), function(i) {
    # Step 1: Get the node labels
    node_labels <- node_v[[rate]][[i]]

    # Step 3: Iterate over preservation levels
    levels <- names(strict_fossil_anc[[rate]][[i]])
    ace_out <- lapply(levels, function(level_name) {
      # Filter ACE matrix for the current preservation level
      strict_fossil_anc[[rate]][[i]][[level_name]][node_labels, , drop = FALSE]
    })

    # Assign names to the ACE output
    names(ace_out) <- levels

    # Return the ACE matrices and the single true matrix
    return(ace_out)
  })
})

# Assign names to the outer list
names(strict_living_nodes) <- names(matrices)


sample_living_nodes <- lapply(seq_along(sample_fossil_anc), function(rate) {
  lapply(seq_along(sample_fossil_anc[[rate]]), function(i) {
    # Step 1: Compute the true matrix once for the current replicate (i)
    node_labels <- node_v[[rate]][[i]]

    # Step 2: Iterate over preservation levels
    levels <- names(sample_fossil_anc[[rate]][[i]])
    ace_out <- lapply(seq_along(levels), function(level_idx) {
      level_name <- levels[[level_idx]]
      reps <- sample_fossil_anc[[rate]][[i]][[level_name]]

      # Filter ACE matrices for each replicate
      lapply(seq_along(reps), function(rep) {
        reps[[rep]][node_labels, , drop = FALSE]
      })
    })

    # Assign names to the ACE output
    names(ace_out) <- levels

    # Return the ACE matrices and the single true matrix
    return(ace_out)
  })
})

# Assign names to the outer list
names(sample_living_nodes) <- names(matrices)



true_living_nodes <- lapply(seq_along(matrices), function(rate) {
  lapply(seq_along(matrices[[rate]]), function(i) {
    # Step 1: Get the node labels
    node_labels <- node_v[[rate]][[i]]

    true_nodes <- matrices[[rate]][[i]][node_labels, ]

    # Assign names to the ACE output

    # Return the ACE matrices and the single true matrix
    return(true_nodes)
  })
})





```


## Disparity pipeline

```{r}

distances_true <- lapply(matrices, lapply, function(x){
  matrix <-  apply(x, c(1,2), as.character)
  dist <- char.diff(matrix, method = "mord", by.col = FALSE)
  return(dist)
})

distances_strict <- lapply(strict_fossil_anc, lapply, lapply, function(matrices) {
  dist <- char.diff(matrices, method = "mord", by.col = FALSE)
  return(dist)
})

distances_rel <- lapply(relative_fossil_anc, lapply, lapply, function(matrices) {
  dist <- char.diff(matrices, method = "mord", by.col = FALSE)
  return(dist)
})

distances_sample <- lapply(sample_fossil_anc, function(rate) {
  mclapply(rate, function(matrix) {
    lapply(matrix, function(fossil){
      lapply(fossil, function(sample) {
        dist <- char.diff(sample, method = "mord", by.col = FALSE)
      })
    })
  }, mc.cores = 10)
})


ord_strict <- lapply(distances_strict, lapply, lapply, function(matrix) {
  ord <- cmdscale(matrix, k = ncol(matrix) - 2, add = TRUE)$points
  return(ord)
})

ord_rel <- lapply(distances_rel, lapply, lapply, function(matrix) {
  ord <- cmdscale(matrix, k = ncol(matrix) - 2, add = TRUE)$points
  return(ord)
})

saveRDS(ord_strict, "../Data/matrices/selectiveExtinction/ord.strict.Rds")
saveRDS(ord_rel, "../Data/matrices/selectiveExtinction/ord.rel.Rds")


# ord_sample <- lapply(distances_sample, lapply, lapply, lapply, function(matrix) {
#   ord <- cmdscale(matrix, k = ncol(matrix) - 2, add = TRUE)$points
#   return(ord)
# })


for(rate_name in names(distances_sample)) {
  cat("Processing rate:", rate_name, "\n")
  
  rate_ord <- list()
  
  for(i in seq_along(distances_sample[[rate_name]])) {
    cat("  Replicate", i, "of", length(distances_sample[[rate_name]]), "\n")
    
    matrices <- distances_sample[[rate_name]][[i]]
    
    # Process this single replicate
    ord_result <- lapply(matrices, function(fossil) {
      lapply(fossil, function(sample) {
        ord <- (cmdscale(sample, k = ncol(sample) - 2, add = TRUE))$points
      })
    })
    
    rate_ord[[i]] <- ord_result
    
    # Cleanup every 10 replicates
    if(i %% 10 == 0) {
      gc()
      cat("    Memory cleanup at replicate", i, "\n")
    }
  }
  
  # Save this rate immediately
  saveRDS(rate_ord, paste0("../Data/matrices/selectiveExtinction/ord_sample_", rate_name, ".rds"))
  
  # Clean up
  rm(rate_ord)
  gc()
}

ord_sample <- list(fast = readRDS("../Data/matrices/selectiveExtinction/ord_sample_fast.rds"), med = readRDS("../Data/matrices/selectiveExtinction/ord_sample_med.rds"), slow = readRDS("../Data/matrices/selectiveExtinction/ord_sample_slow.rds"))

ord_true <- lapply(distances_true, lapply, function(matrix){
    ord <-  (cmdscale(matrix, k = ncol(matrix) - 2, add = TRUE))$points
    return(ord)
})

saveRDS(ord_true, "../Data/matrices/selectiveExtinction/ord.true.Rds")



########################################################################################################################

source("../Functions/find.extinction.time.R")
ord_sample <- list(fast = readRDS("../Data/matrices/selectiveExtinction/ord_sample_fast.rds"), med = readRDS("../Data/matrices/selectiveExtinction/ord_sample_med.rds"), slow = readRDS("../Data/matrices/selectiveExtinction/ord_sample_slow.rds"))

ord_rel <- readRDS("../Data/matrices/selectiveExtinction/ord.rel.Rds")
ord_strict <- readRDS("../Data/matrices/selectiveExtinction/ord.strict.Rds")
ord_true <- readRDS("../Data/matrices/selectiveExtinction/ord.true.Rds")

tip_ages <- lapply(trees, lapply, tip.ages)
extinction_times <- lapply(tip_ages, lapply, find.extinction.time)


time_slices <- lapply(extinction_times, lapply, function(time){
  intervals <- c(time + 20, time,  time - 20)
  return(intervals)
})

chrono_subsets_strict <- lapply(names(ord_strict), function(rate_name){
  rate_result <- lapply(seq_along(ord_strict[[rate_name]]), function(i){
    fossil_result <- lapply(names(ord_strict[[rate_name]][[i]]), function(fossil_name){
      tryCatch({
        chrono <- chrono.subsets(ord_strict[[rate_name]][[i]][[fossil_name]], 
                                tree = fossil_trees[[rate_name]][[i]][[fossil_name]], 
                                method = "d",
                                time = time_slices[[rate_name]][[i]],
                                inc.nodes = TRUE)
        return(chrono)
      }, warning = function(w) {
        rm(w)
        })
    })
    # Preserve fossil names (innermost layer)
    names(fossil_result) <- names(ord_strict[[rate_name]][[i]])
    return(fossil_result)
  })
  return(rate_result)
})

# Preserve rate names (outermost layer)
names(chrono_subsets_strict) <- names(ord_strict)


chrono_subsets_rel <- lapply(names(ord_rel), function(rate_name){
  rate_result <- lapply(seq_along(ord_rel[[rate_name]]), function(i){
    fossil_result <- lapply(names(ord_rel[[rate_name]][[i]]), function(fossil_name){
      tryCatch({
        chrono <- chrono.subsets(ord_rel[[rate_name]][[i]][[fossil_name]], 
                                tree = fossil_trees[[rate_name]][[i]][[fossil_name]], 
                                method = "d",
                                time = time_slices[[rate_name]][[i]],
                                inc.nodes = TRUE)
        return(chrono)
      }, warning = function(w) {
        rm(w)
        })
    })
    # Preserve fossil names (innermost layer)
    names(fossil_result) <- names(ord_rel[[rate_name]][[i]])
    return(fossil_result)
  })
  return(rate_result)
})

# Preserve rate names (outermost layer)
names(chrono_subsets_rel) <- names(ord_rel)


chrono_subsets_sample <- lapply(names(ord_sample), function(rate_name){
  rate_result <- lapply(seq_along(ord_sample[[rate_name]]), function(i){
    fossil_result <- lapply(names(ord_sample[[rate_name]][[i]]), function(fossil_name){
      tryCatch({
        chrono <- chrono.subsets(ord_sample[[rate_name]][[i]][[fossil_name]], 
                                tree = fossil_trees[[rate_name]][[i]][[fossil_name]], 
                                method = "d",
                                time = time_slices[[rate_name]][[i]],
                                inc.nodes = TRUE)
        return(chrono)
      }, warning = function(w) {
        rm(w)
        })
    })
    # Preserve fossil names (innermost layer)
    names(fossil_result) <- names(ord_sample[[rate_name]][[i]])
    return(fossil_result)
  })
  return(rate_result)
})

# Preserve rate names (outermost layer)
names(chrono_subsets_sample) <- names(ord_sample)




chrono_subsets_true <- lapply(names(ord_true), function(rate_name){
  rate_result <- lapply(seq_along(ord_true[[rate_name]]), function(i){
    chrono <- chrono.subsets(ord_true[[rate_name]][[i]], 
                              tree = trees[[rate_name]][[i]], 
                              method = "d", 
                              time = time_slices[[rate_name]][[i]],
                              inc.nodes =  TRUE )
      return(chrono)
    })
  return(rate_result)
})

# Preserve rate names (outermost layer)
names(chrono_subsets_true) <- names(ord_true)

```


## Disparity across mass extinction

```{R}
disp_true <- lapply(chrono_subsets_true, lapply, dispRity, metric = c(mean, displacements))
disp_strict <- lapply(chrono_subsets_strict, lapply, lapply, function(chrono) {
  if(is.null(chrono)) {
    return(NULL)
  }
  tryCatch({
    dispRity(chrono, metric = c(mean, displacements))}, warning = function(w) {
      NULL
    })
})


disp_rel <- lapply(chrono_subsets_rel, lapply, lapply, function(chrono) {
  if(is.null(chrono)) {
    return(NULL)
  }
  tryCatch({
    dispRity(chrono, metric = c(mean, displacements))}, warning = function(w) {
      NULL
    })
})

disp_sample <- lapply(chrono_subsets_sample, lapply, lapply, function(chrono) {
  if(is.null(chrono)) {
    return(NULL)
  }
  tryCatch({
    dispRity(chrono, metric = c(mean, displacements))}, warning = function(w) {
      NULL
    })
})
disp_sample <- list(
  slow = disp_sample$slow,
  med = disp_sample$med,
  fast = disp_sample$fast
)

# disp_sample <- lapply(chrono_subsets_sample, lapply, lapply, dispRity, metric = c(sum, variances))



ace_disp <- list(strict = disp_strict, rel = disp_rel, sample = disp_sample)
### get disparity

ace_disp_change <- lapply(ace_disp, lapply, lapply, lapply, function (disp){
  if(is.null(disp)) {
    return(NULL)
  }
  values <- get.disparity(disp)
  change <- (values[[2]] - values[[1]]) / values[[1]]
})

true_change <- lapply(disp_true, lapply, function(disp){
  values <- get.disparity(disp)
  change <- (values[[2]] - values[[1]]) / values[[1]]
})


# Remove NULLs at the innermost level (fossil level)
ace_disp_change_clean <- lapply(ace_disp_change, function(method) {
  lapply(method, function(rate) {
    lapply(rate, function(replicate) {
      replicate[!sapply(replicate, is.null)]  # Keep only non-NULL elements
    })
  })
})

source("../Functions/utility.R")

matched_strict <- create.matched.data(true_change, ace_disp_change_clean["strict"])
matched_rel <- create.matched.data(true_change, ace_disp_change_clean["rel"])
matched_sample <- create.matched.data(true_change, ace_disp_change_clean["sample"])


# Extract relative changes for plotting

# Heat colors (red = high preservation, yellow = low)
library(viridis)
fossil_colors <- viridis(5, option = "plasma", direction = -1)
names(fossil_colors) <- c("all", "fossil_high", "fossil_med", "fossil_low", "living")


# # Terrain colors
# fossil_colors <- terrain.colors(5)
# names(fossil_colors) <- c("all", "fossil_high", "fossil_med", "fossil_low", "living")



png("../Manuscript/figures/selectiveExtinction/rel_disp_change_av_disp.png", width = 350, height = 250, units = "mm", res = 600)

# Create correlation plot
# Create the plot
plot(matched_rel$true, matched_rel$ace,
     xlab = "True Disparity Change",
     ylab = "ACED Disparity Change",  # Fixed label
     main = "Disparity Change: Relative ACE vs True",
     pch = 16, 
     col = fossil_colors[matched_rel$fossil_level],
     cex = 1.7)

# Add 1:1 line
abline(0, 1, col = "red", lty = 2, lwd = 2)
legend("topleft", 
       legend = names(fossil_colors),
       col = fossil_colors,
       pch = 16,
       cex = 1.2,
       title = "Fossil Level")

dev.off()

png("../Manuscript/figures/selectiveExtinction/strict_disp_change_av_disp.png", width = 350, height = 250, units = "mm", res = 600)

# Create correlation plot
# Create the plot
plot(matched_strict$true, matched_strict$ace,
     xlab = "True Disparity Change",
     ylab = "ACED Disparity Change",  # Fixed label
     main = "Disparity Change: Strict ACE vs True",
     pch = 16, 
     col = fossil_colors[matched_strict$fossil_level],
     cex = 1.7)

# Add 1:1 line
abline(0, 1, col = "red", lty = 2, lwd = 2)
legend("topleft", 
       legend = names(fossil_colors),
       col = fossil_colors,
       pch = 16,
       cex = 1.2,
       title = "Fossil Level")

dev.off()

png("../Manuscript/figures/selectiveExtinction/sample_disp_change_av_disp.png", width = 350, height = 250, units = "mm", res = 600)

# Create correlation plot
# Create the plot
plot(matched_sample$true, matched_sample$ace,
     xlab = "True Disparity Change",
     ylab = "ACED Disparity Change",  # Fixed label
     main = "Disparity Change: Sample ACE vs True",
     pch = 16, 
     col = fossil_colors[matched_sample$fossil_level],
     cex = 1.7)

# Add 1:1 line
abline(0, 1, col = "red", lty = 2, lwd = 2)
legend("topleft", 
       legend = names(fossil_colors),
       col = fossil_colors,
       pch = 16,
       cex = 1.2,
       title = "Fossil Level")

dev.off()


strict_fit <- lm( matched_strict$ace ~ matched_strict$true)
rel_fit <- lm( matched_rel$ace ~ matched_rel$true)
sample_fit <- lm( matched_sample$ace ~ matched_sample$true)

summary(sample_fit)
# Add correlation coefficient
cor_coef <- cor(matched_strict$true, matched_strict$ace, use = "complete.obs")
# text(x = min(true_changes, na.rm = TRUE), 
#      y = max(treated_changes, na.rm = TRUE),
#      labels = paste("r =", round(cor_coef, 3)),
#      adj = c(0, 1))

## perhaps do a glm or mixed effects to figure which treatment gives the best model fit.
```
1. Do the errors vary with extinction selectivity?

    Are disparity changes larger (and harder to reconstruct) when extinction is more trait-based?

    Plot: |True Disparity Change| vs |Error (True - ACE)|

```{R}


# var_strict_diffs <- lapply(sum_var_strict, lapply, function(disparity){
#   true <- disparity$true[[1]]$elements[1]
#   ace <- disparity$ace
#   diffs <- lapply(ace, function(pres){
#     diff <- (pres[[1]]$elements[1] - true) / true
#   })
#   return(diffs)
# })







```


## Testing for model switch

```{R}

tree <-  rtree(n = 100)


event <- make.events(target = "traits", condition = age.condition(5), modification = traits.update(process = BM.process))


ou <- make.traits(OU.process)
traits <- map.traits(ou, tree, event)



```