---
title: "ACE Project master"
author: "Caleb"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

## Generate trees

```{r}

library(treats)

bd_params <- make.bd.params(speciation = 1, extinction = 0.7)

stop_rule <- list(max.living = 50)
set.seed(123)
trees <- treats(stop.rule = stop_rule, bd.params = bd_params, null.error = 100, replicates = 20)
# write.tree(tree, file = "../Data/processed/trees/simulated.tre")
# tree <- read.tree("../Data/processed/trees/simulated.tre")

# tree <- read.nexus("../Data/raw/trees/Beck2014.tre")
# tree <- makeNodeLabel(tree)
# tree <- set.root.time(tree)

```


## Discrete trait simulation

Firstly lets simulate completely random matrix.

```{r}

tree_labels <- lapply(trees, function(tree) {
  c(tree$tip.label, tree$node.label)
})

# Create a matrix with random numbers between 0 and 3
# how do the number of character states influence ancestral states???? say if there are 5 different states, does this influence disparity analyses. could do another matrix where the number of states changes halfway through the tree to see how ace compares??? 
random_discrete <- lapply(tree_labels, function(tree){
                        matrix <- matrix(sample(0:3, length(tree) * 100, replace = TRUE), 
                        nrow = length(tree), 
                        ncol = 100)
                        rownames(matrix) <- tree
                        return(matrix)})





```


Now we can simulate trait data at varying levels of transition rate.

```{r}


low_transitions <- matrix(c(
    0.90, 0.05, 0.05, 
    0.05, 0.99, 0.05,
    0.05, 0.05, 0.99 
), nrow = 3, byrow = TRUE)

low_traits <- treats::make.traits(process = discrete.process, n = 100, process.args = list(transitions = low_transitions))

med_transitions <- matrix(c(
    0.5, 0.25, 0.25, 
    0.25, 0.5, 0.25,
    0.25, 0.25, 0.5 
), nrow = 3, byrow = TRUE)

med_traits <- treats::make.traits(process = discrete.process, n = 100, process.args = list(transitions = med_transitions))


high_transitions <- matrix(c(
    0.33, 0.33, 0.33, 
    0.33, 0.33, 0.33,
    0.33, 0.33, 0.33 
), nrow = 3, byrow = TRUE)

high_traits <- treats::make.traits(process = discrete.process, n = 100, process.args = list(transitions = high_transitions))


## map the traits

low_matrices <- lapply(trees, function(tree){
    mapped <- map.traits(low_traits, tree)
    matrix <- mapped$data
    return(matrix)
})

med_matrices <- lapply(trees, function(tree){
    mapped <- map.traits(low_traits, tree)
    matrix <- mapped$data
    return(matrix)
})

high_matrices <- lapply(trees, function(tree){
    mapped <- map.traits(high_traits, tree)
    matrix <- mapped$data
    return(matrix)
})

discrete_matrices <- list(random = random_discrete, low = low_matrices, med = med_matrices, high = high_matrices )



# test <- geiger::sim.char(tree, Q, nsim = 100, model = "discrete")
# test <- do.call(cbind, test)
# q <- list(rbind(c(-.5, .5), c(.5, -.5)))


```

## Different tree coverages

```{r}

## fossil degredation as well
# tree <- trees[[1]]
remove.fossil <- function(trees, matrices) {
  living_matrices <- Map(function(tree, matrix) {
    ages <- tree.age(tree) # get tip ages
    tips <- ages$element[ages$ages == 0] # find only living species
    living_matrix <- matrix[rownames(matrix) %in% tips, ]
    living_matrix <- apply(living_matrix, c(1, 2), as.character)
 # keep only living species in matrix
    living_tree <- keep.tip(tree, tips) # keep only living species in tree
    return(list(matrix = living_matrix, tree = living_tree)) # return as list
  }, trees, matrices)
  
  return(living_matrices)
}

#     tip_matrix_characters <- apply(tip_matrix, c(1, 2), as.character)


fossil.pres <- function(trees, matrices, preservation = c(0.05, 0.15, 0.5)){
  if (!preservation %in% c(0.05, 0.15, 0.5)) {
       stop("Invalid preservation value. Must be one of 0.05, 0.15, or 0.5.")
   }
   fossil_matrices <- Map(function(tree, matrix){
    ages <- tree.age(tree)
    tips <- ages$element[ages$ages == 0] # keep living species
    fossils <- ages$element[ages$ages > 0 & grepl("^t", ages$element)]  # find fossils
    if(preservation == 0.5){
        sample <- sample(fossils, size = length(fossils)/2)
        } ## sample fossils by a specified preservation level
    if(preservation == 0.15){
        sample <- sample(fossils, size = ceiling(length(fossils) * 0.15))
        }
    if(preservation == 0.05){
        sample <- sample(fossils, size = ceiling(length(fossils) * 0.05))
        }
    kept <- c(sample, tips)
    fossil_matrix <- matrix[rownames(matrix) %in% kept, ] # keep taxa that are fossil preserved
    fossil_matrix <- apply(fossil_matrix, c(1, 2), as.character) # make character for ace
    pruned  <- keep.tip(tree, kept) # prune tree
    return(list(matrix = fossil_matrix, tree = pruned))
   }, trees, matrices)
   return(fossil_matrices)
}

living <- lapply(discrete_matrices, remove.fossil, trees = trees)
fossilised_high <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.5)
fossilised_med <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.15)
fossilised_low <- lapply(discrete_matrices, fossil.pres, trees = trees, preservation = 0.05)


fossil_levels <- list(living = living, fossil_high = fossilised_high, fossil_med = fossilised_med, fossil_low = fossilised_low)


```




## Now simulating continuous characters

```{r}
## BM process

bm <- make.traits(process = BM.process, n = 100)
ou <- make.traits(process = OU.process, n = 100)
random_cont <- make.traits(process = no.process, n = 100)

bm_matrices <- lapply(trees, function(tree){
    mapped <- map.traits(bm, tree)
    matrix <- mapped$data
    return(matrix)
})
ou_matrices <- lapply(trees, function(tree){
    mapped <- map.traits(ou, tree)
    matrix <- mapped$data
    return(matrix)
})
random_cont_matrices <- lapply(trees, function(tree){
    mapped <- map.traits(random_cont, tree)
    matrix <- mapped$data
    return(matrix)
}) 
continuous_matrices <- list(bm = bm_matrices, ou = ou_matrices, random = random_cont_matrices)

living_cont <- lapply(continuous_matrices, remove.fossil, trees = trees)
fossilised_high_cont <- lapply(continuous_matrices, fossil.pres, trees = trees, preservation = 0.5)
fossilised_med_cont <- lapply(continuous_matrices, fossil.pres, trees = trees, preservation = 0.15)
fossilised_low_cont <- lapply(continuous_matrices, fossil.pres, trees = trees, preservation = 0.05)

# fossil_list <- list(living = livin)

```


## discrete pipeline

```{R}
source("../Functions/apply.dist.R")
source("../Functions/apply.ordination.R")

distances <- lapply(discrete_matrices, dist.func)
traitspaces <- ord.func(distances)

disp_obj <- lapply(traitspaces, make.dispRity, tree = tree)

do.time.slice <- function(matrices, chrono_models, FADLAD = NULL, time_interval) {
  time_slices <- lapply(chrono_models, function(model) {
    chrono.subsets(matrices, method = "continuous", FADLAD = FADLAD, model = model, time = time_interval)
  })
  names(time_slices) <- chrono_models
  return(time_slices)
}

chrono_models <- c("acctran", "deltran", "random", "proximity", "equal.split", "gradual.split")


time_sliced_discrete <- lapply(disp_obj, do.time.slice, chrono_models = chrono_models, time_interval = 10)

bootstrapped_discrete <- lapply(time_sliced_discrete, lapply, boot.matrix, bootstraps = 100)

disparity_discrete <- lapply(bootstrapped_discrete, lapply,  dispRity, metric = c(sum, variances), na.rm = TRUE, verbose = TRUE)

saveRDS(disparity_discrete, "../Data/processed/disparity/aceproject/beckdiscrete.Rds")
saveRDS(disparity_discrete, "../Data/processed/disparity/aceproject/simdiscrete.Rds")


plot(disparity_discrete$low$gradual.split)

```


## Continuous pipeline

```{r}

traitspace_cont <- pca.func(continuous_matrices)

disp_obj_cont <- lapply(traitspace_cont, make.dispRity, tree = tree)


time_sliced_cont <- lapply(disp_obj_cont, do.time.slice, chrono_models = chrono_models, time_interval = 10)

bootstrapped_cont <- lapply(time_sliced_cont, lapply, boot.matrix, bootstraps = 100)

disparity_cont <- lapply(bootstrapped_cont, lapply,  dispRity, metric = c(sum, variances), na.rm = TRUE, verbose = TRUE)

saveRDS(disparity_cont, "../Data/processed/disparity/aceproject/beckcont.Rds")
saveRDS(disparity_cont, "../Data/processed/disparity/aceproject/simcont.Rds")


plot(disparity_cont$ou$proximity)

```

## Discrete with ace (pre-ord)

```{r}

# tip_matrices <- lapply(discrete_matrices, function(matrix){
#     tip_matrix <- matrix[!grepl("^n", rownames(matrix)), ]
#     tip_matrix_characters <- apply(tip_matrix, c(1, 2), as.character)
#     return(tip_matrix_characters)
# })


# tips_noncharacter <- apply(tip_matrices$low, c(1,2), as.numeric)
# test_simmap <- make.simmap(tree, state1, nsim = 10)
# outp <- describe.simmap(test_simmap)

anc.states <- function(fossil_levels){
  lapply(fossil_levels, function(level){
    lapply(level, function(x) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = x$matrix, 
                          tree = x$tree, 
                          models = "ER", 
                          output = "combined")
  return(anc_states)})})}

fossil_anc <- lapply(fossil_levels, anc.states)

test <- fossil_anc$living$high[[1]]
true <- discrete_matrices$high[[1]]


get.nodes <- function(true, ace){
  node_v <- rownames(ace)[grepl("^n", rownames(ace))]
  ace_filtered <- ace[node_v,]
  true_filtered <- true[node_v,]
  return(list(ace = ace_filtered, true = true_filtered))
}

source("../Functions/node.dist.R")


true_vs_ace <- lapply(names(fossil_anc), function(preservation) {
  lapply(names(discrete_matrices), function(level) {
    result <- Map(function(true_matrix, ace_matrix) {
      get.nodes(true_matrix, ace_matrix)
    }, discrete_matrices[[level]], preservation[[level]])
    names(result) <- names(discrete_matrices[[level]])  # Preserve names at this level
    return(result)
  })
})

names(true_vs_ace) <- names(fossil_anc)



names(fossil_high_all) <- names(discrete_matrices)

node_dist <- lapply(true_vs_ace, function(preservation){
lapply(preservation, function(level){
  lapply(level, function(matrices){
    distances <- node.dist(matrices)
  })
})
})


distance_means <- lapply(node_dist, function(preservation){
  lapply(preservation, function(level){
  lapply(level, function(matrices){
    means <- mean(matrices)
  })
})
})




test <- lapply(fossil_high_all$high, node.dist)


# Parse ACE
n_rows <- nrow(test)
n_cols <- ncol(test)
parsed_ace <- vector("list", n_rows)
for (i in 1:n_rows) {
  parsed_ace[[i]] <- vector("list", n_cols)
  for (j in 1:n_cols) {
    cell <- test[i, j]
    if (is.na(cell) || cell == "") {
      parsed_ace[[i]][[j]] <- NA
    } else {
      parsed_ace[[i]][[j]] <- as.numeric(unlist(strsplit(cell, "/")))
    }
  }
}

# Distance function using numeric vector directly
compute_distance <- function(true_state, ace_states) {
  # If the state is the same (exact match), distance is 0
  if (true_state %in% ace_states && length(ace_states) == 1) {
    return(0)
  }
  
  # If there is uncertainty of 3 (i.e., 0/1/2), assign 0.5 distance
  if (length(ace_states) == 3) {
    return(0.5)
  }
  
  # If there is uncertainty of 2 (i.e., 0/1)
  if (length(ace_states) == 2) {
    # If true state is not in the estimated states, return 1
    if (!(true_state %in% ace_states)) {
      return(1)
    }
    # If true state is in the estimated states (0 or 1), return 0.5
    return(0.25)
  }
  
  # Otherwise, the true state and estimated state are different (i.e., distance is 1)
  return(1)
}

# Initialize distance matrix
n_rows <- length(parsed_ace)
n_cols <- length(parsed_ace[[1]])
distance_matrix <- matrix(NA, nrow = n_rows, ncol = n_cols)

# Loop through nodes and characters
for (i in 1:n_rows) {
  for (j in 1:n_cols) {
    true_state <- as.numeric(true_nodes[i, j])  # assumes true_nodes is a numeric matrix
    ace_states <- parsed_ace[[i]][[j]]          # this is already a numeric vector
    distance_matrix[i, j] <- compute_distance(true_state, ace_states)
  }
}


## mean(distance_matrix)
#[1] 0.5040306

# r$> sd(distance_matrix)
# [1] 0.244602

distances_anc <- lapply(fossil_anc,  dist.func)
traitspaces_anc <- lapply(distances_anc, ord.func)


## map the corresponding trees and ancestral states
disp_obj_anc <- lapply(names(traitspaces_anc), function(level) {
  Map(function(traitspace, fossil_level) {
    make.dispRity(traitspace, fossil_level$tree)
  }, traitspaces_anc[[level]], low_fossil_levels[[level]])
})
names(disp_obj_anc) <- names(traitspaces_anc)



time_sliced_discrete_anc <- lapply(disp_obj_anc, lapply, do.time.slice, chrono_models = "gradual.split", time_interval = 10)

bootstrapped_discrete_anc <- lapply(time_sliced_discrete_anc, lapply, lapply, boot.matrix, bootstraps = 100)

disparity_discrete_anc <- lapply(bootstrapped_discrete_anc, lapply, lapply,  dispRity, metric = c(sum, variances), na.rm = TRUE, verbose = TRUE)

plot(disparity_discrete_anc$fossil_low[[8]]$gradual.split)


##but when running with beck tree, i find that ace recovers the correct disparity signal. this is likely because of the distribution of tips through time
plot(disparity_discrete_anc$wide_random$gradual.split)

saveRDS(disparity_discrete_anc, "../Data/processed/disparity/aceproject/beckdiscreteanc.Rds")

saveRDS(disparity_discrete_anc, "../Data/processed/disparity/aceproject/simdiscreteanc.Rds")


## this is super important, as this means we have likely been inflating disparity of earlier nodes due to ancestral state estimation. and because the fossil record of these earlier nodes is usually very incomplete anyway, it adds a smoothing effect.


```

## Continuous with ace (pre-ord)

```{r}

tip_matrices_cont <- lapply(continuous_matrices, function(matrix){
    tip_matrix <- matrix[!grepl("^n", rownames(matrix)), ]
    return(tip_matrix)
})


cont_anc_states <- lapply(tip_matrices_cont, function(matrix) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = matrix, 
                          tree = tree, 
                          models = "ML", 
                          output = "combined")
  
  return(anc_states)  
})

traitspace_cont_anc <- pca.func(cont_anc_states)

disp_obj_cont_anc <- lapply(traitspace_cont_anc, make.dispRity, tree = tree)


time_sliced_cont_anc <- lapply(disp_obj_cont_anc, do.time.slice, chrono_models = chrono_models, time_interval = 10)

bootstrapped_contanc <- lapply(time_sliced_cont_anc, lapply, boot.matrix, bootstraps = 100)

disparity_cont_anc <- lapply(bootstrapped_contanc, lapply,  dispRity, metric = c(sum, variances), na.rm = TRUE, verbose = TRUE)

saveRDS(disparity_cont_anc, "../Data/processed/disparity/aceproject/beckcontanc.Rds")
saveRDS(disparity_cont_anc, "../Data/processed/disparity/aceproject/simcontanc.Rds")



plot(disparity_cont_anc$ou$gradual.split)

```

## Discrete with ace (post-ord)

```{r}

tip_matrices <- lapply(discrete_matrices, function(matrix){
    tip_matrix <- matrix[!grepl("^n", rownames(matrix)), ]
    tip_matrix_characters <- apply(tip_matrix, c(1, 2), as.character)
    return(tip_matrix_characters)
})


distances <- dist.func(tip_matrices)
traitspaces <- ord.func(distances)

post_ord_anc <- lapply(traitspaces, function(matrix) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = matrix, 
                          tree = tree, 
                          models = "ML", 
                          output = "combined")
  
  return(anc_states)  
})


disp_obj_post_anc <- lapply(post_ord_anc, make.dispRity, tree = tree)


time_sliced_discrete_post_anc <- lapply(disp_obj_post_anc, do.time.slice, chrono_models = chrono_models, time_interval = 10)

bootstrapped_discrete_post_anc <- lapply(time_sliced_discrete_post_anc, lapply, boot.matrix, bootstraps = 100)

disparity_discrete_post_anc <- lapply(bootstrapped_discrete_post_anc, lapply,  dispRity, metric = c(sum, variances), na.rm = TRUE, verbose = TRUE)

plot(disparity_discrete_post_anc$low$gradual.split)

saveRDS(disparity_discrete_post_anc, "../Data/processed/disparity/aceproject/simpostancdiscrete.Rds")

## this is super important, as this means we have likely been inflating disparity of earlier nodes due to ancestral state estimation. and because the fossil record of these earlier nodes is usually very incomplete anyway, it adds a smoothing effect.


```


## Discrete with ace (pre-ord & sample)

```{r}

tip_matrices <- lapply(discrete_matrices, function(matrix){
    tip_matrix <- matrix[!grepl("^n", rownames(matrix)), ]
    tip_matrix_characters <- apply(tip_matrix, c(1, 2), as.character)
    return(tip_matrix_characters)
})

normal_sampling <- list(fun = rnorm, param = list(mean = mean, sd = function(x)return(diff(range(x))/4)))


discrete_anc_states_sample <- lapply(tip_matrices, function(matrix) {
  # Run multi.ace for each tree
  anc_states <- multi.ace(data = matrix, 
                          tree = tree, 
                          models = "ER", 
                          output = "combined",
                          sample = 1000,
                          sample.fun = normal_sampling)
  
  return(anc_states)  
})

distances_anc_sample <- lapply(discrete_anc_states_sample, dist.func)
traitspaces_anc_sample <- lapply(distances_anc_sample, ord.func)

disp_obj_anc_sample <- lapply(traitspaces_anc_sample, lapply, make.dispRity, tree = tree)


time_sliced_discrete_anc_sample <- lapply(disp_obj_anc_sample, lapply, chrono.subsets, method = "continuous", model = "gradual.split", time = 10)

# bootstrapped_discrete_anc_sample <- lapply(time_sliced_discrete_anc_sample, lapply, boot.matrix, bootstraps = 100)

disparity_discrete_anc_sample <- lapply(time_sliced_discrete_anc_sample, lapply,  dispRity, metric = c(sum, variances), na.rm = TRUE, verbose = TRUE)

saveRDS(disparity_discrete_anc_sample, "../Data/processed/disparity/aceproject/simancdiscretesample.Rds")

plot(disparity_discrete_anc_sample$low[[234]])

low <- lapply(disparity_discrete_anc_sample$low, get.disparity)
low <- do.call(cbind, low)



subset_names <- names(disparity_discrete_anc_sample$low[[1]]$disparity)

# Group the 1000 values into 10 subsets
grouped_subsets <- lapply(subset_names, function(subset_name) {
  # Extract the values for the current subset across all 1000 lists
  values <- sapply(low, function(x) x[[subset_name]])
  
  # Calculate the median of the values
  median_value <- median(values, na.rm = TRUE)
  
  # Reshape the values into a single row with 1000 columns
  values_matrix <- matrix(values, nrow = 1, byrow = TRUE)
  
  # Create the desired structure
  list(
    elements = matrix(median_value, nrow = 1),  # Median as a single-row matrix
     values_matrix               # All values as a matrix
  )
})
names(grouped_subsets) <- subset_names


disparity_discrete_anc_sample$low[[1]]$disparity <- grouped_subsets
disparity_discrete_anc_sample$low[[1]]$call$bootstrap <- list(1000, "full", NULL)

plot(disparity_discrete_anc_sample$low[[1]])

print.dispRity(disparity_discrete_anc_sample$low[[1]])


    
## this is super important, as this means we have likely been inflating disparity of earlier nodes due to ancestral state estimation. and because the fossil record of these earlier nodes is usually very incomplete anyway, it adds a smoothing effect.


```


## Testing against claddis

```{r echo = FALSE}
# write.nexus.data(discrete_matrices$low, file = "../Data/raw/matrices/random.nex") 

# claddis_matrix <- read_nexus_matrix("../Data/raw/matrices/Brusatte2014.nex")
# bru_tree <- read.nexus("../Data/raw/trees/Brusatte2014.nex")


# ancestral_states <- estimate_ancestral_states(claddis_matrix,  bru_tree)
```