---
title: "ACE Project plotting"
author: "Caleb"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

# Aceing our understanding of ASE

Fossils are hypothesised to increase the accuracy of ASE by breaking long branches in phylogenetic trees. This project aims to investigate how fossil coverage affects the accuracy of ASE in discrete and continuous trait matrices. 

To do this, I simulated a set of 100 trees with 50 extant tips and speciation at 1 and extinction at 0.7. This leaves trees with 50 - 200 fossil tips (i.e tips > time 0).

For discrete traits, i simulated trait data across each tree (100 characters, 85% binary, 15% 3-state) using Mk model with 3 different transition rates (low - medium - high). For continuous traits, I simulated both BM and OU model trait data (100 characters again). This outputs matrices where both tips and nodes have values (these are termed 'true values').

I then pruned these matrices of the node values and tip values at 5 levels:
living - only extant tips
fossil low - 5% fossil coverage and all extant tips
fossil mid - 15% fossil coverage and all extant tips
fossil high - 50% fossil coverage and all extant tips 
all - all tips (extant and fossils)

I then ran ancestral state estimation on the pruned discrete and continuous matrices.

To allow for comparisons between different fossil coverages, only the nodes estimated in living were used in subsequent analyses. This leaves 49 nodes for each tree used.



```{r echo = FALSE}
lenient_means <- readRDS("../Data/matrices/lenientmeans.rds")
strict_means <- readRDS("../Data/matrices/strictmeans.rds")
morpho_diffs <- readRDS("../Data/matrices/morphospacediffs.Rds")
sum_var <- readRDS("../Data/matrices/vardiffs.rds")
sum_range <- readRDS("../Data/matrices/rangediffs.rds")
# displ_diffs <-  readRDS("../Data/matrices/dispdiffs.rds")

sum_var_low_dim <- readRDS("../Data/matrices/vardiffs_lowdim.rds")

cont_dists <- readRDS("../Data/matrices/contdists.Rds")
#
# unlisted_strict <- lapply(strict_means, function(mean) {
#     unlist(mean, recursive = TRUE)
# })

# unlisted_uncertain <- lapply(uncertain_means, function(mean) {
#     unlist(mean, recursive = TRUE)
# })

# split_strict <- lapply(unlisted_strict, function(mean){
#   split(mean, gsub("\\d+", "", names(mean)))
# })

# split_uncertain <- lapply(unlisted_uncertain, function(mean){
#   split(mean, gsub("\\d+", "", names(mean)))
# })#



# combined_data <- do.call(rbind, lapply(names(strict_means), function(level) {
#   # Iterate over replicates within each level
#   do.call(rbind, lapply(seq_along(strict_means[[level]]), function(replicate_index) {
#     # Extract the data for the current replicate
#     data <- strict_means[[level]][[replicate_index]]
    
#     # Create a data frame for the current replicate
#     df <- data.frame(
#       replicate = replicate_index,
#       all = data$all,
#       high = data$fossil_high,
#       mid  = data$fossil_med,
#       low  = data$fossil_low,
#       living = data$living,
#       level = level  # Add a column to indicate the level
#     )
    
#     # Convert to long format
#     df_long <- pivot_longer(df, cols = c("all", "high", "mid", "low", "living"), 
#                             names_to = "preservation_level", values_to = "distance")
#     df_long$preservation_level <- factor(df_long$preservation_level, levels = c("all", "high", "mid", "low", "living"))
    
#     return(df_long)
#   }))
# }))

# ggplot(combined_data, aes(x = preservation_level, y = distance, fill = preservation_level)) +
#   geom_boxplot() +
#   facet_wrap(~level, scales = "free_y") +  # Separate plots for each level (fast, med, slow)
#   labs(x = "Fossil Preservation Level", y = "Distance to True Value", title = "Strict Means by Preservation Level") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability


```


```{r echo = FALSE}
# boxplot(split_strict$slow, 
#         main = "Slow strict", 
#         ylab = "Distances", 
#         xlab = "Preservation levels",
#         las = 2)  # Rotate x-axis labels for better readability


# boxplot(split_strict$med, 
#         main = "Med strict",
#         ylab = "Distances",
#         xlab = "Preservation levels",
#         las = 2)  # Rotate x-axis labels for better readability

# boxplot(split_strict$fast, 
#         main = "Fast strict",
#         ylab = "Distances",
#         xlab = "Preservation levels",
#         las = 2)  # Rotate x-axis labels for better readability



```


```{r echo = FALSE}
# boxplot(split_uncertain$slow, 
#         main = "Slow uncertain", 
#         ylab = "Distances", 
#         xlab = "Preservation levels",
#         las = 2)  # Rotate x-axis labels for better readability


# boxplot(split_uncertain$med, 
#         main = "Med uncertain",
#         ylab = "Distances",
#         xlab = "Preservation levels",
#         las = 2)  # Rotate x-axis labels for better readability

# boxplot(split_uncertain$fast, 
#         main = "Fast uncertain",
#         ylab = "Distances",
#         xlab = "Preservation levels",
#         las = 2)  # Rotate x-axis labels for better readability

```


```{R echo = FALSE}
# ord_dist <- readRDS("../Data/matrices/orddists.Rds")
# unlisted_morpho_dists <- lapply(ord_dist, function(rate) {
#     unlist(rate, recursive = TRUE)
# })

# split_dists <- lapply(unlisted_morpho_dists, function(rate){
#   split(rate, gsub("\\d+", "", names(rate)))
# })

```


```{R echo = FALSE}
# boxplot(split_dists$slow, 
#         main = "Slow ", 
#         ylab = "Distances", 
#         xlab = "Preservation levels",
#         las = 2)  # Rotate x-axis labels for better readability

# boxplot(split_dists$med, 
#         main = "Med ",
#         ylab = "Distances",
#         xlab = "Preservation levels",
#         las = 2)  # Rotate x-axis labels for better readability

# boxplot(split_dists$fast, 
#         main = "Fast ",
#         ylab = "Distances",
#         xlab = "Preservation levels",
#         las = 2)  # Rotate x-axis labels for better readability



```


```{R echo = FALSE}
# library(tidyr)
# library(ggplot2)

# # List of levels to loop through
# levels <- c("fast", "med", "slow")

# # Loop through each level and generate a plot
# for (level in levels) {
#   # Extract the data for the current level
#   data <- split_strict[[level]]
#   n <- length(data$fossil_high)  # assuming all same length
  
#   # Create a data frame
#   df <- data.frame(
#     replicate = 1:n,
#     high = data$fossil_high,
#     mid  = data$fossil_med,
#     low  = data$fossil_low,
#     living = data$living
#   )
  
#   # Convert to long format
#   df_long <- pivot_longer(df, cols = c("high", "mid", "low", "living"), names_to = "preservation_level", values_to = "distance")
#   df_long$preservation_level <- factor(df_long$preservation_level, levels = c("high", "mid", "low", "living"))
  
#   # Generate the plot
#   p <- ggplot(df_long, aes(x = preservation_level, y = distance, group = replicate)) +
#     geom_line(alpha = 0.3) +  # thin, semi-transparent lines
#     geom_point(alpha = 0.3) +
#     stat_summary(aes(group = 1), fun = median, geom = "line", color = "red", size = 1.2) + 
#     labs(x = "Fossil Preservation Level", y = "Distance to True Value", title = paste("Transition rate:", level)) +
#     theme_minimal()
  
#   # Print the plot
#   print(p)
# }
```


## Strict distances

Anything over than correct single point estimates are penalised, including uncertainty. I then calculated the mean score (somewhere between 0-1) for each matrix at each preservation level.

```{R echo = FALSE}
library(tidyr)
library(ggplot2)

# List of levels to loop through
levels <- c("fast", "med", "slow")

# Loop through each level and generate a plot
library(tidyr)
library(ggplot2)


# Combine data for all levels and replicates into a single data frame

# Combine data for all levels and replicates into a single data frame
combined_data <- do.call(rbind, lapply(names(strict_means), function(level) {
  # Iterate over replicates within each level
  do.call(rbind, lapply(seq_along(strict_means[[level]]), function(replicate_index) {
    # Extract the data for the current replicate
    data <- strict_means[[level]][[replicate_index]]
    
    # Create a data frame
    df <- data.frame(
      replicate = replicate_index,
      all = data$all,
      high = data$fossil_high,
      mid  = data$fossil_med,
      low  = data$fossil_low,
      living = data$living,
      level = level  # Add a column to indicate the level
    )
    
    # Convert to long format
    df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"), 
                            names_to = "preservation_level", values_to = "distance")
    df_long$preservation_level <- factor(df_long$preservation_level, levels = c("living", "low", "mid", "high", "all"))
    
    return(df_long)
  }))
}))

# Generate the combined plot
# Generate the combined plot
ggplot(combined_data, aes(x = preservation_level, y = distance, color = level, fill = level)) +
  stat_summary(
    aes(group = level),
    fun.data = function(y) {
      data.frame(
        y = median(y),
        ymin = quantile(y, 0.25),  # 25th percentile
        ymax = quantile(y, 0.75)  # 75th percentile
      )
    },
    geom = "ribbon",
    alpha = 0.2,  # Transparency for the shaded area
    color = NA    # No border for the ribbon
  ) +
  stat_summary(
    aes(group = level),
    fun = median,
    geom = "line",
    size = 1.2
  ) +
  labs(
    x = "Fossil Preservation Level",
    y = "Distance to True State"
  ) +
  theme_minimal() +
  scale_y_reverse()    # Match fill colors to line colors

```



## Lenient distances

Lenient distances score 0 if the correct state is given, allowing uncertainty (i.e. 0/1 would be scored 0 if true state is 1). Penalties are given for wrong point estimates (i.e. ace recovers 1 when true state is 0).

```{R echo = FALSE}
# library(tidyr)
# library(ggplot2)

combined_data <- do.call(rbind, lapply(names(lenient_means), function(level) {
  # Iterate over replicates within each level
  do.call(rbind, lapply(seq_along(lenient_means[[level]]), function(replicate_index) {
    # Extract the data for the current replicate
    data <- lenient_means[[level]][[replicate_index]]
    
    # Create a data frame
    df <- data.frame(
      replicate = replicate_index,
      all = data$all,
      high = data$fossil_high,
      mid  = data$fossil_med,
      low  = data$fossil_low,
      living = data$living,
      level = level  # Add a column to indicate the level
    )
    
    # Convert to long format
    df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"), 
                            names_to = "preservation_level", values_to = "distance")
    df_long$preservation_level <- factor(df_long$preservation_level, levels = c("living", "low", "mid", "high", "all"))
    
    return(df_long)
  }))
}))

# Generate the combined plot
# Generate the combined plot
ggplot(combined_data, aes(x = preservation_level, y = distance, color = level, fill = level)) +
  stat_summary(
    aes(group = level),
    fun.data = function(y) {
      data.frame(
        y = median(y),
        ymin = quantile(y, 0.25),  # 25th percentile
        ymax = quantile(y, 0.75)  # 75th percentile
      )
    },
    geom = "ribbon",
    alpha = 0.2,  # Transparency for the shaded area
    color = NA    # No border for the ribbon
  ) +
  stat_summary(
    aes(group = level),
    fun = median,
    geom = "line",
    size = 1.2
  ) +
  labs(
    x = "Fossil Preservation Level",
    y = "Distance to True Value"
  ) +
  theme_minimal() +
  scale_y_reverse()   # Match fill colors to line colors

```

## Relative Difference in traitspace 

Here, I ordinated the true nodal morphospace vs the aced nodal morphospace. I then calcualted the pairwise distances between nodes and divided these by the maximum distance from the centre of the morphospace. I then took the median for each tree replicate for each preservation.

I'm wondering if the clustering of points near the centre of traitspace in fast is artificially decreasing this distance?

```{R echo = FALSE}
combined_data <- do.call(rbind, lapply(names(morpho_diffs), function(level) {
  # Iterate over replicates within each level
  do.call(rbind, lapply(seq_along(morpho_diffs[[level]]), function(replicate_index) {
    # Extract the data for the current replicate
    data <- morpho_diffs[[level]][[replicate_index]]
    
    # Create a data frame
    df <- data.frame(
      replicate = replicate_index,
      all = data$all,
      high = data$fossil_high,
      mid  = data$fossil_med,
      low  = data$fossil_low,
      living = data$living,
      level = level  # Add a column to indicate the level
    )
    
    # Convert to long format
    df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"), 
                            names_to = "preservation_level", values_to = "distance")
    df_long$preservation_level <- factor(df_long$preservation_level, levels = c("living", "low", "mid", "high", "all"))
    
    return(df_long)
  }))
}))

# Generate the combined plot
# Generate the combined plot
ggplot(combined_data, aes(x = preservation_level, y = distance, color = level, fill = level)) +
  stat_summary(
    aes(group = level),
    fun.data = function(y) {
      data.frame(
        y = median(y),
        ymin = quantile(y, 0.25),  # 25th percentile
        ymax = quantile(y, 0.75)  # 75th percentile
      )
    },
    geom = "ribbon",
    alpha = 0.2,  # Transparency for the shaded area
    color = NA    # No border for the ribbon
  ) +
  stat_summary(
    aes(group = level),
    fun = median,
    geom = "line",
    size = 1.2
  ) +
  labs(
    x = "Fossil Preservation Level",
    y = "Distance to True Value"
  ) +
  theme_minimal() +
  scale_y_reverse()   # Match fill colors to line colors



```

## Sum of variances

Fast and medium transition rates results in higher disparity than low transition rates (more character differences between taxa). Because there is high uncertainty in estimating fast/medium rates (resulting in 0/1/2 etc) this creates high aced disparity estimates, which are therefore more similar to the true value

Slow does not. Disparity in the true slow matrix is very low (less transitions, less differences between taxa).  With low fossil coverage, there is higher uncertainty in ace estimate which inflates disparity. With added fossils, this decreases, hence why slow estimates improve with increased fossil coverage.

```{R echo = FALSE}
combined_data <- do.call(rbind, lapply(names(sum_var), function(level) {
  # Iterate over replicates within each level
  do.call(rbind, lapply(seq_along(sum_var[[level]]), function(replicate_index) {
    # Extract the data for the current replicate
    data <- sum_var[[level]][[replicate_index]]
    
    # Create a data frame
    df <- data.frame(
      replicate = replicate_index,
      all = data$all,
      high = data$fossil_high,
      mid  = data$fossil_med,
      low  = data$fossil_low,
      living = data$living,
      level = level  # Add a column to indicate the level
    )
    
    # Convert to long format
    df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"), 
                            names_to = "preservation_level", values_to = "distance")
    df_long$preservation_level <- factor(df_long$preservation_level, levels = c("living", "low", "mid", "high", "all"))
    
    return(df_long)
  }))
}))

# Generate the combined plot
# Generate the combined plot
ggplot(combined_data, aes(x = preservation_level, y = distance, color = level, fill = level)) +
  stat_summary(
    aes(group = level),
    fun.data = function(y) {
      data.frame(
        y = median(y),
        ymin = quantile(y, 0.25),  # 25th percentile
        ymax = quantile(y, 0.75)  # 75th percentile
      )
    },
    geom = "ribbon",
    alpha = 0.2,  # Transparency for the shaded area
    color = NA    # No border for the ribbon
  ) +
  stat_summary(
    aes(group = level),
    fun = median,
    geom = "line",
    size = 1.2
  ) +
  labs(
    x = "Fossil Preservation Level",
    y = "Distance to True Value"
  ) +
  theme_minimal() +
  scale_y_reverse()   # Match fill colors to line colors



```

## Sum of variances lower dim

When use first 10 PC axes of morphospace you find that aced estimates becomes much stronger, particularly slow rates.

```{R echo = FALSE}
combined_data <- do.call(rbind, lapply(names(sum_var_low_dim), function(level) {
  # Iterate over replicates within each level
  do.call(rbind, lapply(seq_along(sum_var_low_dim[[level]]), function(replicate_index) {
    # Extract the data for the current replicate
    data <- sum_var_low_dim[[level]][[replicate_index]]
    
    # Create a data frame
    df <- data.frame(
      replicate = replicate_index,
      all = data$all,
      high = data$fossil_high,
      mid  = data$fossil_med,
      low  = data$fossil_low,
      living = data$living,
      level = level  # Add a column to indicate the level
    )
    
    # Convert to long format
    df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"), 
                            names_to = "preservation_level", values_to = "distance")
    df_long$preservation_level <- factor(df_long$preservation_level, levels = c("living", "low", "mid", "high", "all"))
    
    return(df_long)
  }))
}))

# Generate the combined plot
# Generate the combined plot
ggplot(combined_data, aes(x = preservation_level, y = distance, color = level, fill = level)) +
  stat_summary(
    aes(group = level),
    fun.data = function(y) {
      data.frame(
        y = median(y),
        ymin = quantile(y, 0.25),  # 25th percentile
        ymax = quantile(y, 0.75)  # 75th percentile
      )
    },
    geom = "ribbon",
    alpha = 0.2,  # Transparency for the shaded area
    color = NA    # No border for the ribbon
  ) +
  stat_summary(
    aes(group = level),
    fun = median,
    geom = "line",
    size = 1.2
  ) +
  labs(
    x = "Fossil Preservation Level",
    y = "Distance to True Value"
  ) +
  theme_minimal() +
  scale_y_reverse()   # Match fill colors to line colors



```

## Sum of ranges

```{R echo = FALSE}
combined_data <- do.call(rbind, lapply(names(sum_range), function(level) {
  # Iterate over replicates within each level
  do.call(rbind, lapply(seq_along(sum_range[[level]]), function(replicate_index) {
    # Extract the data for the current replicate
    data <- sum_range[[level]][[replicate_index]]
    
    # Create a data frame
    df <- data.frame(
      replicate = replicate_index,
      all = data$all,
      high = data$fossil_high,
      mid  = data$fossil_med,
      low  = data$fossil_low,
      living = data$living,
      level = level  # Add a column to indicate the level
    )
    
    # Convert to long format
    df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"), 
                            names_to = "preservation_level", values_to = "distance")
    df_long$preservation_level <- factor(df_long$preservation_level, levels = c("living", "low", "mid", "high", "all"))
    
    return(df_long)
  }))
}))

# Generate the combined plot
# Generate the combined plot
ggplot(combined_data, aes(x = preservation_level, y = distance, color = level, fill = level)) +
  stat_summary(
    aes(group = level),
    fun.data = function(y) {
      data.frame(
        y = median(y),
        ymin = quantile(y, 0.25),  # 25th percentile
        ymax = quantile(y, 0.75)  # 75th percentile
      )
    },
    geom = "ribbon",
    alpha = 0.2,  # Transparency for the shaded area
    color = NA    # No border for the ribbon
  ) +
  stat_summary(
    aes(group = level),
    fun = median,
    geom = "line",
    size = 1.2
  ) +
  labs(
    x = "Fossil Preservation Level",
    y = "Distance to True Value"
  ) +
  theme_minimal() +
  scale_y_reverse()   # Match fill colors to line colors



```



## Continuous traits (BM vs OU)

Here, i measured the euclidean distance between scaled BM and OU ace estimates and true values. BM is closer to true value than OU because there model misspecification with OU (ace estimates using BM model). Interestingly, fossil coverage appears to make no improvement to ace estimates.

```{R echo = FALSE}
combined_data <- do.call(rbind, lapply(names(cont_dists), function(level) {
  # Iterate over replicates within each level
  do.call(rbind, lapply(seq_along(cont_dists[[level]]), function(replicate_index) {
    # Extract the data for the current replicate
    data <- cont_dists[[level]][[replicate_index]]
    
    # Create a data frame
    df <- data.frame(
      replicate = replicate_index,
      all = data$all,
      high = data$fossil_high,
      mid  = data$fossil_med,
      low  = data$fossil_low,
      living = data$living,
      level = level  # Add a column to indicate the level
    )
    
    # Convert to long format
    df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"), 
                            names_to = "preservation_level", values_to = "distance")
    df_long$preservation_level <- factor(df_long$preservation_level, levels = c("living", "low", "mid", "high", "all"))
    
    return(df_long)
  }))
}))

# Generate the combined plot
# Generate the combined plot
ggplot(combined_data, aes(x = preservation_level, y = distance, color = level, fill = level)) +
  stat_summary(
    aes(group = level),
    fun.data = function(y) {
      data.frame(
        y = median(y),
        ymin = quantile(y, 0.25),  # 25th percentile
        ymax = quantile(y, 0.75)  # 75th percentile
      )
    },
    geom = "ribbon",
    alpha = 0.2,  # Transparency for the shaded area
    color = NA    # No border for the ribbon
  ) +
  stat_summary(
    aes(group = level),
    fun = median,
    geom = "line",
    size = 1.2
  ) +
  labs(
    x = "Fossil Preservation Level",
    y = "Distance to True Value"
  ) +
  theme_minimal() +
  scale_y_reverse()   # Match fill colors to line colors



```


```{R}



combined_data <- do.call(rbind, lapply(names(displacements), function(level) {
  # Iterate over replicates within each level
  do.call(rbind, lapply(seq_along(displacements[[level]]), function(replicate_index) {
    # Extract the data for the current replicate
    data <- displacements[[level]][[replicate_index]]
    
    # Create a data frame
    df <- data.frame(
      replicate = replicate_index,
      all = data$all,
      high = data$fossil_high,
      mid  = data$fossil_med,
      low  = data$fossil_low,
      living = data$living,
      level = level  # Add a column to indicate the level
    )
    
    # Convert to long format
    df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"), 
                            names_to = "preservation_level", values_to = "distance")
    df_long$preservation_level <- factor(df_long$preservation_level, levels = c("living", "low", "mid", "high", "all"))
    
    return(df_long)
  }))
}))

# Generate the combined plot
# Generate the combined plot
ggplot(combined_data, aes(x = preservation_level, y = distance, color = level, fill = level)) +
  stat_summary(
    aes(group = level),
    fun.data = function(y) {
      data.frame(
        y = median(y),
        ymin = quantile(y, 0.25),  # 25th percentile
        ymax = quantile(y, 0.75)  # 75th percentile
      )
    },
    geom = "ribbon",
    alpha = 0.2,  # Transparency for the shaded area
    color = NA    # No border for the ribbon
  ) +
  stat_summary(
    aes(group = level),
    fun = median,
    geom = "line",
    size = 1.2
  ) +
  labs(
    x = "Fossil Preservation Level",
    y = "Distance to True Value"
  ) +
  theme_minimal() +
  scale_y_reverse()   # Match fill colors to line colors



```