---
title: "ACE Project continuous"
author: "Caleb"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---


# Overall disparity JOB ID: 7904947
Treatment: 

* no rescaled
* no remove axes


```{R}
library(ggplot2)
library(tidyr)
library(patchwork)
library(scales)
# Read and rearrange in one step
rate_names <- c("slow", "med", "fast")


# r <- readRDS("../Data/cluster/discrete/raw/diff_sumvar_strict_001.rds")

methods <- c("strict", "sample", "point_postord", "sample_postord", "no_ace")
sample_sum_var <- list()
sum_var_change_sample_50 <- list()


diff_sum_var_50 <- setNames(lapply(methods, function(method) {
  setNames(lapply(rate_names, function(rate) {
    lapply(seq_len(100), function(rep_idx) {
      file_path <- file.path("..","Data","cluster","discrete","disparity","50t","raw",
                             paste0("8218707_diff_sumvar_", method, sprintf("_%03d.rds", rep_idx)))
      if(!file.exists(file_path)) return(NULL)         # avoid hard error if missing
      data <- readRDS(file_path)
      data[[rate]]
    })
  }), rate_names)
}), methods)


sum_var_data <- do.call(rbind, lapply(names(diff_sum_var_50), function(method){
  dataset <- diff_sum_var_50[[method]]
  do.call(rbind, lapply(names(dataset), function(rate){
    do.call(rbind, lapply(seq_along(dataset[[rate]]), function(replicate_index){
      data  <- dataset[[rate]][[replicate_index]]
      df <- data.frame(
        replicate = replicate_index,
        all = data$all,
        high = data$fossil_high,
        mid = data$fossil_med,
        low = data$fossil_low,
        living = data$living,
        rate = rate,
        method = method,
        metric = "Sum of variances"
      )
      pivot_longer(df, cols = c("living","low","mid","high","all"),
                   names_to = "preservation_level", values_to = "distance")
    }))
  }))
}))

# saveRDS(sum_var_diff_strict, "../Data/processed/discrete/dispDiffs/sum_var_diff_strict.rds")
# saveRDS(sum_var_diff_sample, "../Data/processed/discrete/dispDiffs/sum_var_diff_sample.rds")
# saveRDS(sum_var_diff_rel, "../Data/processed/discrete/dispDiffs/sum_var_diff_rel.rds")
# saveRDS(sum_var_diff_no_ace, "../Data/processed/discrete/dispDiffs/sum_var_diff_no_ace.rds")
# saveRDS(sum_var_diff_point_postord, "../Data/processed/discrete/dispDiffs/sum_var_diff_point_postord.rds")
# saveRDS(sum_var_diff_sample_postord, "../Data/processed/discrete/dispDiffs/sum_var_diff_sample_postord.rds")





########################################################################################################################
diff_neighbours_50 <- setNames(lapply(methods, function(method) {
  setNames(lapply(rate_names, function(rate) {
    lapply(seq_len(100), function(rep_idx) {
      file_path <- file.path("..","Data","cluster","discrete","disparity","50t","raw",
                             paste0("8218707_diff_neighbours_", method, sprintf("_%03d.rds", rep_idx)))
      if(!file.exists(file_path)) return(NULL)         # avoid hard error if missing
      data <- readRDS(file_path)
      data[[rate]]
    })
  }), rate_names)
}), methods)

neighbours_data <- do.call(rbind, lapply(names(diff_neighbours_50), function(method){
  dataset <- diff_neighbours_50[[method]]
  do.call(rbind, lapply(names(dataset), function(rate){
    do.call(rbind, lapply(seq_along(dataset[[rate]]), function(replicate_index){
      data  <- dataset[[rate]][[replicate_index]]
      df <- data.frame(
        replicate = replicate_index,
        all = data$all,
        high = data$fossil_high,
        mid = data$fossil_med,
        low = data$fossil_low,
        living = data$living,
        rate = rate,
        method = method,
        metric = "Average neighbour"
      )
      pivot_longer(df, cols = c("living","low","mid","high","all"),
                   names_to = "preservation_level", values_to = "distance")
    }))
  }))
}))




diff_quantiles_50 <- setNames(lapply(methods, function(method) {
  setNames(lapply(rate_names, function(rate) {
    lapply(seq_len(100), function(rep_idx) {
      file_path <- file.path("..","Data","cluster","discrete","disparity","50t","raw",
                             paste0("8218707_diff_quant_", method, sprintf("_%03d.rds", rep_idx)))
      if(!file.exists(file_path)) return(NULL)         # avoid hard error if missing
      data <- readRDS(file_path)
      data[[rate]]
    })
  }), rate_names)
}), methods)


quantiles_data <- do.call(rbind, lapply(names(diff_quantiles_50), function(method){
  dataset <- diff_quantiles_50[[method]]
  do.call(rbind, lapply(names(dataset), function(rate){
    do.call(rbind, lapply(seq_along(dataset[[rate]]), function(replicate_index){
      data  <- dataset[[rate]][[replicate_index]]
      df <- data.frame(
        replicate = replicate_index,
        all = data$all,
        high = data$fossil_high,
        mid = data$fossil_med,
        low = data$fossil_low,
        living = data$living,
        rate = rate,
        method = method,
        metric = "Sum of quantiles"
      )
      pivot_longer(df, cols = c("living","low","mid","high","all"),
                   names_to = "preservation_level", values_to = "distance")
    }))
  }))
}))

```



## Combined (neighbours and sum var) no rel

```{R}

# # Combine both datasets
# combined_data <- rbind(quantiles_data, sum_var_data)

# # Factor the variables
# combined_data$threshold_method <- factor(combined_data$threshold_method, 
#                                        levels = c("sum_var_diff_strict", "sum_var_diff_sample", "sum_var_diff_point_postord", "sum_var_diff_sample_postord", "sum_var_diff_no_ace"), 
#                                        labels = c("Strict majority rule", "No majority rule (sample)", "Post-ord ACE (point)", "Post-ord ACE (sample)", "No ACE"))

# combined_data$preservation_level <- factor(combined_data$preservation_level, 
#                                          levels = c("living", "low", "mid", "high", "all"))

# combined_data$metric <- factor(combined_data$metric, 
#                              levels = c( "Sum of quantiles", "Sum of Variances"))


# colours <- c("slow" = "#8805A8", "med" = "#00B945", "fast" = "#FFB600")

# combined_plot_no_rel <- ggplot(combined_data, aes(x = preservation_level, y = distance, color = level, fill = level)) +
#   geom_boxplot(
#     alpha = 0.7,
#     outlier.size = 0.5,
#     outlier.alpha = 0.6,
#     position = position_dodge(width = 0.8)
#   ) +
#   geom_hline(yintercept = 0, color = "black", linewidth = 0.6, linetype = "dashed") +
#   facet_grid(metric ~ threshold_method) +  # 2 rows (metrics) × 5 columns (methods)
#   labs(
#     x = "Fossil sampling",
#     y = "Relative disparity error",
#     color = "Transition Rate",
#     fill = "Transition Rate"
#   ) +
#   theme_minimal() +
#   scale_color_manual(values = colours) +
#   scale_fill_manual(values = colours) + 
#   theme(
#     axis.text = element_text(size = 10, face = "bold"),
#     axis.title = element_text(size = 12, face = "bold"),
#     axis.ticks = element_line(size = 1),
#     legend.position = "bottom",
#     panel.grid = element_blank(),
#     strip.text = element_text(size = 10, face = "bold"),
#     strip.background = element_rect(color = "black", fill = "white", linewidth = 0.5),
#     panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
#   ) +
#   coord_trans(y = "pseudo_log") +
#   scale_y_reverse()


# ggsave("../Manuscript/figures/overall_combined_7904947_no_rel.png", combined_plot_no_rel, width = 12, height = 8, dpi = 700, bg = "white")

```

## Ridge plot

```{R}

library(ggridges)
library(ggplot2)
library(tidyr)
library(scales)

# Your existing data preparation code stays the same...
combined_data <- rbind(quantiles_data, sum_var_data)

combined_data$method <-  factor(combined_data$method, 
                                       levels = c("strict", "sample", "point_postord", "sample_postord", "no_ace"), 
                                       labels = c("Pre-ord ACE (point)", "Pre-ord ACE (dist)", "Post-ord ACE (point)", "Post-ord ACE (dist)", "No ACE"))

combined_data$preservation_level <- factor(combined_data$preservation_level, 
                                         levels = c("living", "low", "mid", "high", "all"),
                                         labels = c("0%", "5%", "15%", "50%", "100%"))

colours <- c("slow" = "#8805A8", "med" = "#00B945", "fast" = "#FFB600")



## get bandwidth

library(MASS)
bw_global <- bw.nrd0(combined_data$distance)
# Create Sum of Quantiles plot
combined_ridge_plot <- ggplot(combined_data, aes(x = distance, y = preservation_level, fill = rate)) +
  stat_density_ridges(
    alpha = 0.7,           # Increase opacity for better visibility
    scale = 1,           # Better ridge separation
    rel_min_height = 0.01, # Show more detail
    # color = "white",       # Add white outlines
    # size = 0.3,
    bandwidth =   bw_global       # Thin white lines between ridges
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.8, linetype = "solid") +
  facet_grid(metric ~ method, 
             labeller = labeller(
               metric = c("Sum of quantiles" = "Sum of Quantiles", 
                         "Sum of variances" = "Sum of Variances"),
               method = c("Pre-ord ACE (point)" = "Pre-ordination ACE\n(Point)",
                                  "Pre-ord ACE (dist)" = "Pre-ordination ACE\n(Distributions)", 
                                  "Post-ord ACE (point)" = "Post-ordination ACE\n(Point)",
                                  "Post-ord ACE (dist)" = "Post-ordination ACE\n(Distributions)",
                                  "No ACE" = "No ACE")
             )) +
  labs(
    x = "Relative Disparity Error",
    y = "Fossil Sampling (%)",
    fill = "Transition Rate"
  ) +
  theme_minimal() +
  scale_fill_manual(
    values = colours,
    labels = c("slow" = "Slow", "med" = "Medium", "fast" = "Fast")
  ) + 
  theme(
    # Text sizes
    axis.text = element_text(size = 11, color = "black"),
    axis.text.x = element_text(angle = 0),
    axis.title = element_text(size = 13, face = "bold", color = "black"),
    
    # Title styling
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5, margin = margin(b = 10)),
    plot.subtitle = element_text(size = 12, hjust = 0.5, color = "gray30", margin = margin(b = 20)),
    
    # Legend
    legend.position = "bottom",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11),
    legend.key.size = unit(0.8, "cm"),
    legend.margin = margin(t = 15),
    
    # Facet labels
    strip.text = element_text(size = 10, face = "bold", color = "black"),
    strip.background = element_rect(fill = "gray95", color = "black", linewidth = 0.3),
    
    # Panel styling
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4),
    panel.grid = element_blank(),
    panel.spacing = unit(0.3, "cm"),
    
    # Overall plot margins
    plot.margin = margin(20, 20, 20, 20)
  ) +
  scale_x_continuous(
    breaks = c(-1, -0.5, 0, 0.5, 1),
    labels = c("-1.0", "-0.5", "0", "0.5", "1.0"),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  coord_cartesian(xlim = c(-1, 1))


ggsave("../Manuscript/figures/ridge_overall_combined_7904947.png", combined_ridge_plot, width = 12, height = 8, dpi = 700, bg = "white")

```


## Poster plot

```{r}


combined_ridge_plot <- ggplot(combined_data, aes(x = distance, y = preservation_level, fill = level)) +
  geom_density_ridges(
    alpha = 0.7,
    scale = 1,
    rel_min_height = 0.01,
    size = 0.1
  ) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1.2, linetype = "dashed") +
  facet_grid(metric ~ threshold_method) +
  labs(
    x = "Relative Disparity Error",
    y = NULL,
    fill = NULL,
    title = NULL,
    subtitle = NULL
  ) +
  theme_minimal() +
  scale_fill_manual(values = colours) +
  theme(
    axis.text = element_text(size = 14, color = "black"),
    axis.text.x = element_text(angle = 0),
    axis.title.y = element_blank(),
    axis.title.x = element_text(size = 18, face = "bold", color = "black"),
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    legend.position = "none",
    strip.text = element_blank(),
    strip.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4),
    panel.grid = element_blank(),
    panel.spacing = unit(0.7, "cm"),
    plot.margin = margin(20, 20, 20, 20),
    panel.background = element_rect(fill = NA, color = NA),  # Make panel transparent
    plot.background = element_rect(fill = NA, color = NA)    # Make plot background transparent
  ) +
  scale_x_continuous(
    breaks = c(-1, -0.5, 0, 0.5, 1),
    labels = c("-1.0", "-0.5", "0", "0.5", "1.0"),
    expand = expansion(mult = c(0.02, 0.02))
  ) +
  coord_cartesian(xlim = c(-1, 1)) +
  scale_y_discrete(expand = expansion(mult = c(0, 0.1)))

ggsave("../Manuscript/figures/ridge_overall_combined_no_labels.png",
       combined_ridge_plot,
       width = 11, height = 8, dpi = 700, bg = "transparent")

## Combined (neighbours and sum var) with rel



```


# Overall disparity JOB ID: 7908227 (rescaled)

## load data
```{R}


sum_var_diff_strict <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rescaled_7908227_12_09/diff_sumvar_strict_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_diff_sample <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rescaled_7908227_12_09/diff_sumvar_sample_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)


sum_var_diff_rel <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rescaled_7908227_12_09/diff_sumvar_rel_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_diff_no_ace <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rescaled_7908227_12_09/diff_sumvar_no_ace_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_diff_point_postord <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rescaled_7908227_12_09/diff_sumvar_point_postord_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)


sum_var_diff_sample_postord <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rescaled_7908227_12_09/diff_sumvar_sample_postord_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)


########################################################################################################################

neighbours_diff_strict <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rescaled_7908227_12_09/diff_neighbours_strict_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

neighbours_diff_sample <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rescaled_7908227_12_09/diff_neighbours_sample_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)


neighbours_diff_rel <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rescaled_7908227_12_09/diff_neighbours_rel_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

neighbours_diff_no_ace <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rescaled_7908227_12_09/diff_neighbours_no_ace_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

neighbours_diff_point_postord <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rescaled_7908227_12_09/diff_neighbours_point_postord_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)


neighbours_diff_sample_postord <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rescaled_7908227_12_09/diff_neighbours_sample_postord_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)







```
## Combined (neighbours and sum var) no rel

```{R}

library(scales)
sum_var_data <- do.call(rbind, lapply(c("sum_var_diff_strict", "sum_var_diff_sample", "sum_var_diff_point_postord", "sum_var_diff_sample_postord", "sum_var_diff_no_ace"), function(dataset_name) {
  dataset <- get(dataset_name)
  do.call(rbind, lapply(names(dataset), function(level) {
    do.call(rbind, lapply(seq_along(dataset[[level]]), function(replicate_index) {
      data <- dataset[[level]][[replicate_index]]
      df <- data.frame(
        replicate = replicate_index,
        all = data$all,
        high = data$fossil_high,
        mid = data$fossil_med,
        low = data$fossil_low,
        living = data$living,
        level = level,
        threshold_method = dataset_name,
        metric = "Sum of variances"  # Add metric identifier
      )
      df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"),
                              names_to = "preservation_level", values_to = "distance")
      return(df_long)
    }))
  }))
}))

# Create neighbours data
neighbours_data <- do.call(rbind, lapply(c("neighbours_diff_strict", "neighbours_diff_sample", "neighbours_diff_point_postord", "neighbours_diff_sample_postord", "neighbours_diff_no_ace"), function(dataset_name) {
  dataset <- get(dataset_name)
  do.call(rbind, lapply(names(dataset), function(level) {
    do.call(rbind, lapply(seq_along(dataset[[level]]), function(replicate_index) {
      data <- dataset[[level]][[replicate_index]]
      df <- data.frame(
        replicate = replicate_index,
        all = data$all,
        high = data$fossil_high,
        mid = data$fossil_med,
        low = data$fossil_low,
        living = data$living,
        level = level,
        threshold_method = gsub("neighbours_diff_", "sum_var_diff_", dataset_name),  # Match naming
        metric = "Neighbours"  # Add metric identifier
      )
      df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"),
                              names_to = "preservation_level", values_to = "distance")
      return(df_long)
    }))
  }))
}))

# Combine both datasets
combined_data <- rbind(sum_var_data, neighbours_data)

# Factor the variables
combined_data$threshold_method <- factor(combined_data$threshold_method, 
                                       levels = c("sum_var_diff_strict", "sum_var_diff_sample", "sum_var_diff_point_postord", "sum_var_diff_sample_postord", "sum_var_diff_no_ace"), 
                                       labels = c("Strict majority rule", "No majority rule (sample)", "Post-ord ACE (point)", "Post-ord ACE (sample)", "No ACE"))

combined_data$preservation_level <- factor(combined_data$preservation_level, 
                                         levels = c("living", "low", "mid", "high", "all"))

combined_data$metric <- factor(combined_data$metric, 
                             levels = c("Sum of variances", "Neighbours"))



combined_plot_no_rel <- ggplot(combined_data, aes(x = preservation_level, y = distance, color = level, fill = level)) +
  geom_boxplot(
    alpha = 0.7,
    outlier.size = 0.5,
    outlier.alpha = 0.6,
    position = position_dodge(width = 0.8)
  ) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.6, linetype = "dashed") +
  facet_grid(metric ~ threshold_method) +  # 2 rows (metrics) × 5 columns (methods)
  labs(
    x = "Fossil sampling",
    y = "Relative disparity error",
    color = "Transition Rate",
    fill = "Transition Rate"
  ) +
  theme_minimal() +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) + 
  theme(
    axis.text = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    axis.ticks = element_line(size = 1),
    legend.position = "bottom",
    panel.grid = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),
    strip.background = element_rect(color = "black", fill = "white", linewidth = 0.5),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  ) +
  coord_trans(y = "pseudo_log") +
  scale_y_reverse()


ggsave("../Manuscript/figures/overall_combined_7904947_no_rel_rescaled.png", combined_plot_no_rel, width = 12, height = 8, dpi = 700, bg = "white")

```

## Combined (neighbours and sum var) with rel

```{R}

library(scales)
sum_var_data <- do.call(rbind, lapply(c("sum_var_diff_strict","sum_var_diff_rel", "sum_var_diff_sample", "sum_var_diff_point_postord", "sum_var_diff_sample_postord", "sum_var_diff_no_ace"), function(dataset_name) {
  dataset <- get(dataset_name)
  do.call(rbind, lapply(names(dataset), function(level) {
    do.call(rbind, lapply(seq_along(dataset[[level]]), function(replicate_index) {
      data <- dataset[[level]][[replicate_index]]
      df <- data.frame(
        replicate = replicate_index,
        all = data$all,
        high = data$fossil_high,
        mid = data$fossil_med,
        low = data$fossil_low,
        living = data$living,
        level = level,
        threshold_method = dataset_name,
        metric = "Sum of variances"  # Add metric identifier
      )
      df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"),
                              names_to = "preservation_level", values_to = "distance")
      return(df_long)
    }))
  }))
}))

# Create neighbours data
neighbours_data <- do.call(rbind, lapply(c("neighbours_diff_strict", "neighbours_diff_rel", "neighbours_diff_sample", "neighbours_diff_point_postord", "neighbours_diff_sample_postord", "neighbours_diff_no_ace"), function(dataset_name) {
  dataset <- get(dataset_name)
  do.call(rbind, lapply(names(dataset), function(level) {
    do.call(rbind, lapply(seq_along(dataset[[level]]), function(replicate_index) {
      data <- dataset[[level]][[replicate_index]]
      df <- data.frame(
        replicate = replicate_index,
        all = data$all,
        high = data$fossil_high,
        mid = data$fossil_med,
        low = data$fossil_low,
        living = data$living,
        level = level,
        threshold_method = dataset_name,  # Match naming
        metric = "Neighbours"  # Add metric identifier
      )
      df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"),
                              names_to = "preservation_level", values_to = "distance")
      return(df_long)
    }))
  }))
}))

# Combine both datasets
combined_data <- rbind(sum_var_data, neighbours_data)

# Factor the variables
combined_data$threshold_method <- factor(combined_data$threshold_method, 
                                       levels = c("sum_var_diff_strict", "sum_var_diff_rel", "sum_var_diff_sample", "sum_var_diff_point_postord", "sum_var_diff_sample_postord", "sum_var_diff_no_ace"), 
                                       labels = c("Strict majority rule", "Relative majority rule", "No majority rule (sample)", "Post-ord ACE (point)", "Post-ord ACE (sample)", "No ACE"))

combined_data$preservation_level <- factor(combined_data$preservation_level, 
                                         levels = c("living", "low", "mid", "high", "all"))

combined_data$metric <- factor(combined_data$metric, 
                             levels = c("Sum of variances", "Neighbours"))



combined_plot_rel <- ggplot(combined_data, aes(x = preservation_level, y = distance, color = level, fill = level)) +
  geom_boxplot(
    alpha = 0.7,
    outlier.size = 0.5,
    outlier.alpha = 0.6,
    position = position_dodge(width = 0.8)
  ) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.6, linetype = "dashed") +
  facet_grid(metric ~ threshold_method) +  # 2 rows (metrics) × 5 columns (methods)
  labs(
    x = "Fossil sampling",
    y = "Relative disparity error",
    color = "Transition Rate",
    fill = "Transition Rate"
  ) +
  theme_minimal() +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) + 
  theme(
    axis.text = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    axis.ticks = element_line(size = 1),
    legend.position = "bottom",
    panel.grid = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),
    strip.background = element_rect(color = "black", fill = "white", linewidth = 0.5),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  ) +
  coord_trans(y = "pseudo_log") +
  scale_y_reverse()


ggsave("../Manuscript/figures/overall_combined_7904947_rel_rescaled.png", combined_plot_rel, width = 12, height = 8, dpi = 700, bg = "white")
```




# Overall disparity JOB ID: 7929071 (remove axes)


## Load data 

```{r}

sum_var_diff_strict <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rmaxes_7929071_16_09/diff_sumvar_strict_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_diff_sample <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rmaxes_7929071_16_09/diff_sumvar_sample_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)


sum_var_diff_rel <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rmaxes_7929071_16_09/diff_sumvar_rel_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_diff_no_ace <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rmaxes_7929071_16_09/diff_sumvar_no_ace_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_diff_point_postord <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rmaxes_7929071_16_09/diff_sumvar_point_postord_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)


sum_var_diff_sample_postord <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rmaxes_7929071_16_09/diff_sumvar_sample_postord_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)


########################################################################################################################

neighbours_diff_strict <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rmaxes_7929071_16_09/diff_neighbours_strict_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

neighbours_diff_sample <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rmaxes_7929071_16_09/diff_neighbours_sample_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)


neighbours_diff_rel <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rmaxes_7929071_16_09/diff_neighbours_rel_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

neighbours_diff_no_ace <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rmaxes_7929071_16_09/diff_neighbours_no_ace_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

neighbours_diff_point_postord <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rmaxes_7929071_16_09/diff_neighbours_point_postord_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)


neighbours_diff_sample_postord <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/clusterOut/rmaxes_7929071_16_09/diff_neighbours_sample_postord_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

```

## Combined (neighbours and sum var) no rel

```{R}

library(scales)
sum_var_data <- do.call(rbind, lapply(c("sum_var_diff_strict", "sum_var_diff_sample", "sum_var_diff_point_postord", "sum_var_diff_sample_postord", "sum_var_diff_no_ace"), function(dataset_name) {
  dataset <- get(dataset_name)
  do.call(rbind, lapply(names(dataset), function(level) {
    do.call(rbind, lapply(seq_along(dataset[[level]]), function(replicate_index) {
      data <- dataset[[level]][[replicate_index]]
      df <- data.frame(
        replicate = replicate_index,
        all = data$all,
        high = data$fossil_high,
        mid = data$fossil_med,
        low = data$fossil_low,
        living = data$living,
        level = level,
        threshold_method = dataset_name,
        metric = "Sum of variances"  # Add metric identifier
      )
      df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"),
                              names_to = "preservation_level", values_to = "distance")
      return(df_long)
    }))
  }))
}))

# Create neighbours data
neighbours_data <- do.call(rbind, lapply(c("neighbours_diff_strict", "neighbours_diff_sample", "neighbours_diff_point_postord", "neighbours_diff_sample_postord", "neighbours_diff_no_ace"), function(dataset_name) {
  dataset <- get(dataset_name)
  do.call(rbind, lapply(names(dataset), function(level) {
    do.call(rbind, lapply(seq_along(dataset[[level]]), function(replicate_index) {
      data <- dataset[[level]][[replicate_index]]
      df <- data.frame(
        replicate = replicate_index,
        all = data$all,
        high = data$fossil_high,
        mid = data$fossil_med,
        low = data$fossil_low,
        living = data$living,
        level = level,
        threshold_method = dataset_name,  # Match naming
        metric = "Neighbours"  # Add metric identifier
      )
      df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"),
                              names_to = "preservation_level", values_to = "distance")
      return(df_long)
    }))
  }))
}))

# Combine both datasets
combined_data <- rbind(sum_var_data, neighbours_data)

# Factor the variables
combined_data$threshold_method <- factor(combined_data$threshold_method, 
                                       levels = c("sum_var_diff_strict", "sum_var_diff_sample", "sum_var_diff_point_postord", "sum_var_diff_sample_postord", "sum_var_diff_no_ace"), 
                                       labels = c("Strict majority rule", "No majority rule (sample)", "Post-ord ACE (point)", "Post-ord ACE (sample)", "No ACE"))

combined_data$preservation_level <- factor(combined_data$preservation_level, 
                                         levels = c("living", "low", "mid", "high", "all"))

combined_data$metric <- factor(combined_data$metric, 
                             levels = c("Sum of variances", "Neighbours"))



combined_plot_no_rel <- ggplot(combined_data, aes(x = preservation_level, y = distance, color = level, fill = level)) +
  geom_boxplot(
    alpha = 0.7,
    outlier.size = 0.5,
    outlier.alpha = 0.6,
    position = position_dodge(width = 0.8)
  ) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.6, linetype = "dashed") +
  facet_grid(metric ~ threshold_method) +  # 2 rows (metrics) × 5 columns (methods)
  labs(
    x = "Fossil sampling",
    y = "Relative disparity error",
    color = "Transition Rate",
    fill = "Transition Rate"
  ) +
  theme_minimal() +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) + 
  theme(
    axis.text = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    axis.ticks = element_line(size = 1),
    legend.position = "bottom",
    panel.grid = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),
    strip.background = element_rect(color = "black", fill = "white", linewidth = 0.5),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  ) +
  coord_trans(y = "pseudo_log") +
  scale_y_reverse()


ggsave("../Manuscript/figures/overall_combined_7904947_no_rel_rmaxes.png", combined_plot_no_rel, width = 12, height = 8, dpi = 700, bg = "white")

```

## Combined (neighbours and sum var) with rel

```{R}

library(scales)
sum_var_data <- do.call(rbind, lapply(c("sum_var_diff_strict","sum_var_diff_rel", "sum_var_diff_sample", "sum_var_diff_point_postord", "sum_var_diff_sample_postord", "sum_var_diff_no_ace"), function(dataset_name) {
  dataset <- get(dataset_name)
  do.call(rbind, lapply(names(dataset), function(le-2.2772987vel) {
    do.call(rbind, lapply(seq_along(dataset[[level]]), function(replicate_index) {
      data <- dataset[[level]][[replicate_index]]
      df <- data.frame(
        replicate = replicate_index,
        all = data$all,
        high = data$fossil_high,
        mid = data$fossil_med,
        low = data$fossil_low,
        living = data$living,
        level = level,
        threshold_method = dataset_name,
        metric = "Sum of variances"  # Add metric identifier
      )
      df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"),
                              names_to = "preservation_level", values_to = "distance")
      return(df_long)
    }))
  }))
}))

# Create neighbours data
neighbours_data <- do.call(rbind, lapply(c("neighbours_diff_strict", "neighbours_diff_rel", "neighbours_diff_sample", "neighbours_diff_point_postord", "neighbours_diff_sample_postord", "neighbours_diff_no_ace"), function(dataset_name) {
  dataset <- get(dataset_name)
  do.call(rbind, lapply(names(dataset), function(level) {
    do.call(rbind, lapply(seq_along(dataset[[level]]), function(replicate_index) {
      data <- dataset[[level]][[replicate_index]]
      df <- data.frame(
        replicate = replicate_index,
        all = data$all,
        high = data$fossil_high,
        mid = data$fossil_med,
        low = data$fossil_low,
        living = data$living,
        level = level,
        threshold_method = dataset_name,  # Match naming
        metric = "Neighbours"  # Add metric identifier
      )
      df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"),
                              names_to = "preservation_level", values_to = "distance")
      return(df_long)
    }))
  }))
}))

# Combine both datasets
combined_data <- rbind(sum_var_data, neighbours_data)

# Factor the variables
combined_data$threshold_method <- factor(combined_data$threshold_method, 
                                       levels = c("sum_var_diff_strict", "sum_var_diff_rel", "sum_var_diff_sample", "sum_var_diff_point_postord", "sum_var_diff_sample_postord", "sum_var_diff_no_ace"), 
                                       labels = c("Strict majority rule", "Relative majority rule", "No majority rule (sample)", "Post-ord ACE (point)", "Post-ord ACE (sample)", "No ACE"))

combined_data$preservation_level <- factor(combined_data$preservation_level, 
                                         levels = c("living", "low", "mid", "high", "all"))

combined_data$metric <- factor(combined_data$metric, 
                             levels = c("Sum of variances", "Neighbours"))



combined_plot_rel <- ggplot(combined_data, aes(x = preservation_level, y = distance, color = level, fill = level)) +
  geom_boxplot(
    alpha = 0.7,
    outlier.size = 0.5,
    outlier.alpha = 0.6,
    position = position_dodge(width = 0.8)
  ) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.6, linetype = "dashed") +
  facet_grid(metric ~ threshold_method) +  # 2 rows (metrics) × 5 columns (methods)
  labs(
    x = "Fossil sampling",
    y = "Relative disparity error",
    color = "Transition Rate",
    fill = "Transition Rate"
  ) +
  theme_minimal() +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) + 
  theme(
    axis.text = element_text(size = 10, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    axis.ticks = element_line(size = 1),
    legend.position = "bottom",
    panel.grid = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),
    strip.background = element_rect(color = "black", fill = "white", linewidth = 0.5),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  ) +
  coord_trans(y = "pseudo_log") +
  scale_y_reverse()


ggsave("../Manuscript/figures/overall_combined_7904947_rel_rmaxes.png", combined_plot_rel, width = 12, height = 8, dpi = 700, bg = "white")
```


















## ANOVA stuff

```{R}
#################################################################################################################
## ANOVA
sum_var_diff_strict <-  readRDS("../Data/processed/discrete/dispDiffs/sum_var_diff_strict.rds")
sum_var_diff_sample <-  readRDS("../Data/processed/discrete/dispDiffs/sum_var_diff_sample.rds")
sum_var_diff_rel <-  readRDS("../Data/processed/discrete/dispDiffs/sum_var_diff_rel.rds")
sum_var_diff_no_ace <-  readRDS("../Data/processed/discrete/dispDiffs/sum_var_diff_no_ace.rds")
sum_var_diff_point_postord <-  readRDS("../Data/processed/discrete/dispDiffs/sum_var_diff_point_postord.rds")
sum_var_diff_sample_postord <-  readRDS("../Data/processed/discrete/dispDiffs/sum_var_diff_sample_postord.rds")


neighbours_diff_strict <-  readRDS("../Data/processed/discrete/dispDiffs/neighbours_diff_strict.rds")
neighbours_diff_sample <-  readRDS("../Data/processed/discrete/dispDiffs/neighbours_diff_sample.rds")
neighbours_diff_rel <-  readRDS("../Data/processed/discrete/dispDiffs/neighbours_diff_rel.rds")
neighbours_diff_no_ace <-  readRDS("../Data/processed/discrete/dispDiffs/neighbours_diff_no_ace.rds")
neighbours_diff_point_postord <-  readRDS("../Data/processed/discrete/dispDiffs/neighbours_diff_point_postord.rds")
neighbours_diff_sample_postord <-  readRDS("../Data/processed/discrete/dispDiffs/neighbours_diff_sample_postord.rds")


sum_var_data <- do.call(rbind, lapply(c("sum_var_diff_strict","sum_var_diff_rel", "sum_var_diff_sample", "sum_var_diff_point_postord", "sum_var_diff_sample_postord", "sum_var_diff_no_ace"), function(dataset_name) {
  dataset <- get(dataset_name)
  do.call(rbind, lapply(names(dataset), function(level) {
    do.call(rbind, lapply(seq_along(dataset[[level]]), function(replicate_index) {
      data <- dataset[[level]][[replicate_index]]
      df <- data.frame(
        replicate = replicate_index,
        all = data$all,
        high = data$fossil_high,
        mid = data$fossil_med,
        low = data$fossil_low,
        living = data$living,
        level = level,
        threshold_method = dataset_name,
        metric = "Sum of variances"  # Add metric identifier
      )
      df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"),
                              names_to = "preservation_level", values_to = "distance")
      return(df_long)
    }))
  }))
}))

# sum_var_data$abs_distance <- abs(sum_var_data$distance)


# sum_var_model <- aov(abs_distance ~ level * threshold_method * preservation_level, data = sum_var_data)
# summary(sum_var_model)

# par(mfrow = c(2,2))
# plot(model)  # Residual plots
# shapiro.test(residuals(model))



# ## model assumptions violated

# sum_var_data$log_abs_distance <- log(sum_var_data$abs_distance + 0.001)
# model_log <- aov(log_abs_distance ~ level * threshold_method * preservation_level, data = sum_var_data)
# plot(model_log)  # Residual plots



# Use aggregate function (base R equivalent of group_by + summarise)
sum_var_data_median <- aggregate(
  distance ~ replicate + level + threshold_method + preservation_level, 
  data = sum_var_data, 
  FUN = median, 
  na.rm = TRUE
)


sum_var_data_median$abs_distance <- abs(sum_var_data_median$distance)
sum_var_data_median$log_abs_distance <- log(sum_var_data_median$abs_distance + 0.001)


sum_var_median_model <- aov(log_abs_distance ~ level * threshold_method * preservation_level, 
                     data = sum_var_data_median)

# par(mfrow = c(2,2))
# plot(model_balanced)
# shapiro.test(residuals(model_balanced))


library(agricolae)

# Tukey HSD test for each factor
tukey_rate <- HSD.test(sum_var_median_model, "level", group = TRUE)
tukey_method <- HSD.test(sum_var_median_model, "threshold_method", group = TRUE)
tukey_preservation <- HSD.test(sum_var_median_model, "preservation_level", group = TRUE)

# View rankings with statistical groups
print("Rate rankings:")
tukey_rate$groups

print("Method rankings:")
tukey_method$groups

print("Preservation rankings:")
tukey_preservation$groups

library(emmeans)
combined_model_means <- emmeans(sum_var_median_model, ~ level * threshold_method * preservation_level)
rate_method_cld <- cld(combined_model_means, alpha = 0.05, Letters = letters)



neighbours_data_median <- aggregate(
  distance ~ replicate + level + threshold_method + preservation_level, 
  data = neighbours_data, 
  FUN = median, 
  na.rm = TRUE
)


neighbours_data_median$abs_distance <- abs(neighbours_data_median$distance)
neighbours_data_median$log_abs_distance <- log(neighbours_data_median$abs_distance + 0.001)


neighbours_median_model <- aov(log_abs_distance ~ level * threshold_method * preservation_level, 
                     data = neighbours_data_median)

# par(mfrow = c(2,2))
# plot(model_balanced)
# shapiro.test(residuals(model_balanced))


library(agricolae)

# Tukey HSD test for each factor
tukey_rate <- HSD.test(neighbours_median_model, "level", group = TRUE)
tukey_method <- HSD.test(neighbours_median_model, "threshold_method", group = TRUE)
tukey_preservation <- HSD.test(neighbours_median_model, "preservation_level", group = TRUE)

# View rankings with statistical groups
print("Rate rankings:")
tukey_rate$groups

print("Method rankings:")
tukey_method$groups

print("Preservation rankings:")
tukey_preservation$groups


combined_model_means <- emmeans(neighbours_median_model, ~ level * threshold_method * preservation_level)
rate_method_cld <- cld(combined_model_means, alpha = 0.05, Letters = letters)























rate_method_means <- emmeans(model_balanced, ~ level * threshold_method)
summary(rate_method_means)

# Rank by performance with statistical significance
pairs(rate_method_means, adjust = "tukey")
cld(rate_method_means, alpha = 0.05, Letters = letters)

# Same for other interactions
rate_preservation_means <- emmeans(model_balanced, ~ level * preservation_level)
method_preservation_means <- emmeans(model_balanced, ~ threshold_method * preservation_level)
summary(method_preservation_means)
cld(method_preservation_means, alpha = 0.05, Letters = letters)





```

##########################################################################################################################
## With less axes


sum_var_diff_strict <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/rmaxes/diff_sumvar_strict_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_diff_sample <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/rmaxes/diff_sumvar_sample_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_diff_sample_median <- lapply(sum_var_diff_sample, lapply, lapply, median)

sum_var_diff_rel <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/rmaxes/diff_sumvar_rel_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_diff_no_ace <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/rmaxes/diff_sumvar_no_ace_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_diff_point_postord <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/rmaxes/diff_sumvar_point_postord_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)


sum_var_diff_sample_postord <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/discrete/rmaxes/diff_sumvar_sample_postord_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)


library(ggplot2)
library(tidyr)

library(scales)
combined_data <- do.call(rbind, lapply(c("sum_var_diff_strict", "sum_var_diff_rel", "sum_var_diff_sample", "sum_var_diff_point_postord", "sum_var_diff_sample_postord", "sum_var_diff_no_ace"), function(dataset_name) {
  dataset <- get(dataset_name)  # Dynamically get the dataset by name
  do.call(rbind, lapply(names(dataset), function(level) {
    do.call(rbind, lapply(seq_along(dataset[[level]]), function(replicate_index) {
      data <- dataset[[level]][[replicate_index]]
      df <- data.frame(
        replicate = replicate_index,
        all = data$all,
        high = data$fossil_high,
        mid = data$fossil_med,
        low = data$fossil_low,
        living = data$living,
        level = level,
        threshold_method = dataset_name  # Add a column for treatment
      )
      df_long <- pivot_longer(df, cols = c("living", "low", "mid", "high", "all"),
                              names_to = "preservation_level", values_to = "distance")
      df_long$preservation_level <- factor(df_long$preservation_level, levels = c("living", "low", "mid", "high", "all"))
      return(df_long)
    }))
  }))
}))

combined_data$threshold_method <- factor(combined_data$threshold_method, levels = c("sum_var_diff_strict", "sum_var_diff_rel", "sum_var_diff_sample", "sum_var_diff_point_postord", "sum_var_diff_sample_postord", "sum_var_diff_no_ace"), labels = c("Strict majority rule", "Relative majority rule", "No majority rule (sample)", "Post-ord ACE (point)", "Post-ord ACE (sample)", "No ACE"))

# log10_reverse_trans <- trans_new(
#   name = "log10_reverse",
#   transform = function(x) -log10(x),  # Apply log10 and reverse
#   inverse = function(x) 10^(-x)      # Inverse of the transformation
# )

colors <- c("slow" = "#8805A8", "med" = "#00B945", "fast" = "#FFB600")
# Generate the combined plot

p <- ggplot(combined_data, aes(x = preservation_level, y = distance, color = level, fill = level)) +
  geom_boxplot(
    alpha = 0.7,
    outlier.size = 0.5,
    outlier.alpha = 0.6,
    position = position_dodge(width = 0.8)
  ) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.6, linetype = "dashed") +
  facet_wrap(~ threshold_method, ncol = 6, nrow = 1) +
  labs(
    x = "Fossil sampling",
    y = "Relative disparity error",
    color = "Transition Rate",
    fill = "Transition Rate"
  ) +
  theme_minimal() +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) + 
  theme(
    axis.text = element_text(size = 15, face = "bold"),
    axis.title = element_text(size = 20, face = "bold"),
    axis.ticks = element_line(size = 1),
    legend.position = "none",
    panel.grid = element_blank(),
    strip.text = element_text(size = 15, face = "bold", hjust = 0.5, vjust = 1),
    strip.background = element_rect(color = "black", fill = "white", linewidth = 0.5),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  ) +
  coord_trans(y = "pseudo_log") +
  scale_y_reverse()