---
title: "Random extinction plots"
author: "Caleb"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---



# Load data

## 50t, 100t 
```{R}
rate_names <- c("slow", "med", "fast")

rate_names <- c("slow", "med", "fast")

########################################################################################################################
# 50t data loading
########################################################################################################################

# Sum of variances
sum_var_change_sample_50 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_sample_sum_var_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_change_strict_50 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_strict_sum_var_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_change_point_postord_50 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_point_postord_sum_var_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_change_sample_postord_50 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_sample_postord_sum_var_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_change_no_ace_50 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_no_ace_sum_var_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_change_true_50 <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_true_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)

########################################################################################################################
# 100t data loading - Sum of variances
########################################################################################################################

sum_var_change_sample_100 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_sample_sum_var_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_change_strict_100 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_strict_sum_var_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_change_point_postord_100 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_point_postord_sum_var_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_change_sample_postord_100 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_sample_postord_sum_var_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_change_no_ace_100 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_no_ace_sum_var_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_var_change_true_100 <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_true_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)

########################################################################################################################
# 50t data loading - Sum of quantiles
########################################################################################################################

sum_quant_change_sample_50 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_sample_sum_quant_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_quant_change_strict_50 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_strict_sum_quant_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_quant_change_point_postord_50 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_point_postord_sum_quant_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_quant_change_sample_postord_50 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_sample_postord_sum_quant_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_quant_change_no_ace_50 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_no_ace_sum_quant_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_quant_change_true_50 <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_true_sum_quant_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)

########################################################################################################################
# 100t data loading - Sum of quantiles
########################################################################################################################

sum_quant_change_sample_100 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_sample_sum_quant_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_quant_change_strict_100 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_strict_sum_quant_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_quant_change_point_postord_100 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_point_postord_sum_quant_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_quant_change_sample_postord_100 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_sample_postord_sum_quant_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_quant_change_no_ace_100 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_no_ace_sum_quant_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

sum_quant_change_true_100 <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_true_sum_quant_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)

########################################################################################################################
# 50t data loading - Displacement
########################################################################################################################

displacement_change_sample_50 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_sample_displacement_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

displacement_change_strict_50 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_strict_displacement_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

displacement_change_point_postord_50 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_point_postord_displacement_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

displacement_change_sample_postord_50 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_sample_postord_displacement_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

displacement_change_no_ace_50 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_no_ace_displacement_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

displacement_change_true_50 <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_true_displacement_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)

########################################################################################################################
# 100t data loading - Displacement
########################################################################################################################

displacement_change_sample_100 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_sample_displacement_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

displacement_change_strict_100 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_strict_displacement_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

displacement_change_point_postord_100 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_point_postord_displacement_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

displacement_change_sample_postord_100 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_sample_postord_displacement_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

displacement_change_no_ace_100 <- setNames(lapply(rate_names, function(rate) {
    all_reps <- lapply(1:100, function(rep) {
        file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_no_ace_displacement_change_%03d.rds", rep)
        data <- readRDS(file_path)
        return(data[[rate]])
    })
}), rate_names)

displacement_change_true_100 <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_true_displacement_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)


```

# Clean and match data

This cleans the data of any NULL results (due to too low fossil sampling). eg in living. And then matches the true change to estimated change

```{R}
create.matched.data <- function(true, estimated) {
    result <- list()
    for(rate_name in names(estimated)) {
        true_matched <- c()
        est_matched <- c()
        fossil_levels <- c()
        for(rep in seq_along(estimated[[rate_name]])) {
            true_val <- true[[rate_name]][[rep]]
            est_rep_val <- estimated[[rate_name]][[rep]]
            fossil_names <- names(est_rep_val)
            # Loop over fossil levels
            for(i in seq_along(est_rep_val)) {
                est_vals <- est_rep_val[[i]]
                n_est <- length(est_vals)
                true_matched <- c(true_matched, rep(true_val, n_est))
                est_matched <- c(est_matched, est_vals)
                fossil_levels <- c(fossil_levels, rep(fossil_names[i], n_est))
            }
        }
        result[[rate_name]] <- list(
            true = true_matched,
            estimated = est_matched,
            fossil_level = fossil_levels
        )
    }
    return(result)
}

# sum_var_change_sample_unlist_50 <- lapply(sum_var_change_sample_50, function(rate_data) {
#     lapply(rate_data, function(replicate_data) {
#         # Apply median directly to each fossil level
#         result <- lapply(replicate_data, function(fossil_level_values) {
#             values <- unlist(fossil_level_values, use.names = FALSE)
#             return(values)
#         })
#         return(result)  # This ensures we return the fossil level list directly
#     })
# })

sum_quant_change_sample_unlist_50 <- lapply(sum_quant_change_sample_50, lapply, lapply, unlist, use.names = FALSE)
sum_quant_change_sample_unlist_100 <- lapply(sum_quant_change_sample_100, lapply, lapply, unlist, use.names = FALSE)
sum_var_change_sample_unlist_50 <- lapply(sum_var_change_sample_50, lapply, lapply, unlist, use.names = FALSE)
sum_var_change_sample_unlist_100 <- lapply(sum_var_change_sample_100, lapply, lapply, unlist, use.names = FALSE)
displacement_change_sample_unlist_50 <- lapply(displacement_change_sample_50, lapply, lapply, unlist, use.names = FALSE)
displacement_change_sample_unlist_100 <- lapply(displacement_change_sample_100, lapply, lapply, unlist, use.names = FALSE)
sum_var_change_sample_postord_unlist_50 <- lapply(sum_var_change_sample_postord_50, lapply, lapply, unlist, use.names = FALSE)
sum_var_change_sample_postord_unlist_100 <- lapply(sum_var_change_sample_postord_100, lapply, lapply, unlist, use.names = FALSE)
sum_quant_change_sample_postord_unlist_50 <- lapply(sum_quant_change_sample_postord_50, lapply, lapply, unlist, use.names = FALSE)
sum_quant_change_sample_postord_unlist_100 <- lapply(sum_quant_change_sample_postord_100, lapply, lapply, unlist, use.names = FALSE)
displacement_change_sample_postord_unlist_50 <- lapply(displacement_change_sample_postord_50, lapply, lapply, unlist, use.names = FALSE)
displacement_change_sample_postord_unlist_100 <- lapply(displacement_change_sample_postord_100, lapply, lapply, unlist, use.names = FALSE)


# Clean all datasets (remove NULLs)
# Sum of variances
sum_var_clean_sample_50 <- lapply(sum_var_change_sample_unlist_50, lapply, function(x) x[!sapply(x, is.null)])
sum_var_clean_sample_100 <- lapply(sum_var_change_sample_unlist_100, lapply, function(x) x[!sapply(x, is.null)])
sum_var_clean_strict_50 <- lapply(sum_var_change_strict_50, lapply, function(x) x[!sapply(x, is.null)])
sum_var_clean_strict_100 <- lapply(sum_var_change_strict_100, lapply, function(x) x[!sapply(x, is.null)])
sum_var_clean_point_postord_50 <- lapply(sum_var_change_point_postord_50, lapply, function(x) x[!sapply(x, is.null)])
sum_var_clean_point_postord_100 <- lapply(sum_var_change_point_postord_100, lapply, function(x) x[!sapply(x, is.null)])
sum_var_clean_sample_postord_50 <- lapply(sum_var_change_sample_postord_unlist_50, lapply, function(x) x[!sapply(x, is.null)])
sum_var_clean_sample_postord_100 <- lapply(sum_var_change_sample_postord_unlist_100, lapply, function(x) x[!sapply(x, is.null)])
sum_var_clean_no_ace_50 <- lapply(sum_var_change_no_ace_50, lapply, function(x) x[!sapply(x, is.null)])
sum_var_clean_no_ace_100 <- lapply(sum_var_change_no_ace_100, lapply, function(x) x[!sapply(x, is.null)])

# Sum of quantiles
sum_quant_clean_sample_50 <- lapply(sum_quant_change_sample_unlist_50, lapply, function(x) x[!sapply(x, is.null)])
sum_quant_clean_sample_100 <- lapply(sum_quant_change_sample_unlist_100, lapply, function(x) x[!sapply(x, is.null)])
sum_quant_clean_strict_50 <- lapply(sum_quant_change_strict_50, lapply, function(x) x[!sapply(x, is.null)])
sum_quant_clean_strict_100 <- lapply(sum_quant_change_strict_100, lapply, function(x) x[!sapply(x, is.null)])
sum_quant_clean_point_postord_50 <- lapply(sum_quant_change_point_postord_50, lapply, function(x) x[!sapply(x, is.null)])
sum_quant_clean_point_postord_100 <- lapply(sum_quant_change_point_postord_100, lapply, function(x) x[!sapply(x, is.null)])
sum_quant_clean_sample_postord_50 <- lapply(sum_quant_change_sample_postord_unlist_50, lapply, function(x) x[!sapply(x, is.null)])
sum_quant_clean_sample_postord_100 <- lapply(sum_quant_change_sample_postord_unlist_100, lapply, function(x) x[!sapply(x, is.null)])
sum_quant_clean_no_ace_50 <- lapply(sum_quant_change_no_ace_50, lapply, function(x) x[!sapply(x, is.null)])
sum_quant_clean_no_ace_100 <- lapply(sum_quant_change_no_ace_100, lapply, function(x) x[!sapply(x, is.null)])

# Displacement
displacement_clean_sample_50 <- lapply(displacement_change_sample_unlist_50, lapply, function(x) x[!sapply(x, is.null)])
displacement_clean_sample_100 <- lapply(displacement_change_sample_unlist_100, lapply, function(x) x[!sapply(x, is.null)])
displacement_clean_strict_50 <- lapply(displacement_change_strict_50, lapply, function(x) x[!sapply(x, is.null)])
displacement_clean_strict_100 <- lapply(displacement_change_strict_100, lapply, function(x) x[!sapply(x, is.null)])
displacement_clean_point_postord_50 <- lapply(displacement_change_point_postord_50, lapply, function(x) x[!sapply(x, is.null)])
displacement_clean_point_postord_100 <- lapply(displacement_change_point_postord_100, lapply, function(x) x[!sapply(x, is.null)])
displacement_clean_sample_postord_50 <- lapply(displacement_change_sample_postord_unlist_50, lapply, function(x) x[!sapply(x, is.null)])
displacement_clean_sample_postord_100 <- lapply(displacement_change_sample_postord_unlist_100, lapply, function(x) x[!sapply(x, is.null)])
displacement_clean_no_ace_50 <- lapply(displacement_change_no_ace_50, lapply, function(x) x[!sapply(x, is.null)])
displacement_clean_no_ace_100 <- lapply(displacement_change_no_ace_100, lapply, function(x) x[!sapply(x, is.null)])

# Create matched data for all methods
# Sum of variances
sum_var_matched_sample_50 <- create.matched.data(sum_var_change_true_50, sum_var_clean_sample_50)
sum_var_matched_sample_100 <- create.matched.data(sum_var_change_true_100, sum_var_clean_sample_100)
sum_var_matched_strict_50 <- create.matched.data(sum_var_change_true_50, sum_var_clean_strict_50)
sum_var_matched_strict_100 <- create.matched.data(sum_var_change_true_100, sum_var_clean_strict_100)
sum_var_matched_point_postord_50 <- create.matched.data(sum_var_change_true_50, sum_var_clean_point_postord_50)
sum_var_matched_point_postord_100 <- create.matched.data(sum_var_change_true_100, sum_var_clean_point_postord_100)
sum_var_matched_sample_postord_50 <- create.matched.data(sum_var_change_true_50, sum_var_clean_sample_postord_50)
sum_var_matched_sample_postord_100 <- create.matched.data(sum_var_change_true_100, sum_var_clean_sample_postord_100)
sum_var_matched_no_ace_50 <- create.matched.data(sum_var_change_true_50, sum_var_clean_no_ace_50)
sum_var_matched_no_ace_100 <- create.matched.data(sum_var_change_true_100, sum_var_clean_no_ace_100)

# Sum of quantiles
sum_quant_matched_sample_50 <- create.matched.data(sum_quant_change_true_50, sum_quant_clean_sample_50)
sum_quant_matched_sample_100 <- create.matched.data(sum_quant_change_true_100, sum_quant_clean_sample_100)
sum_quant_matched_strict_50 <- create.matched.data(sum_quant_change_true_50, sum_quant_clean_strict_50)
sum_quant_matched_strict_100 <- create.matched.data(sum_quant_change_true_100, sum_quant_clean_strict_100)
sum_quant_matched_point_postord_50 <- create.matched.data(sum_quant_change_true_50, sum_quant_clean_point_postord_50)
sum_quant_matched_point_postord_100 <- create.matched.data(sum_quant_change_true_100, sum_quant_clean_point_postord_100)
sum_quant_matched_sample_postord_50 <- create.matched.data(sum_quant_change_true_50, sum_quant_clean_sample_postord_50)
sum_quant_matched_sample_postord_100 <- create.matched.data(sum_quant_change_true_100, sum_quant_clean_sample_postord_100)
sum_quant_matched_no_ace_50 <- create.matched.data(sum_quant_change_true_50, sum_quant_clean_no_ace_50)
sum_quant_matched_no_ace_100 <- create.matched.data(sum_quant_change_true_100, sum_quant_clean_no_ace_100)

# Displacement
displacement_matched_sample_50 <- create.matched.data(displacement_change_true_50, displacement_clean_sample_50)
displacement_matched_sample_100 <- create.matched.data(displacement_change_true_100, displacement_clean_sample_100)
displacement_matched_strict_50 <- create.matched.data(displacement_change_true_50, displacement_clean_strict_50)
displacement_matched_strict_100 <- create.matched.data(displacement_change_true_100, displacement_clean_strict_100)
displacement_matched_point_postord_50 <- create.matched.data(displacement_change_true_50, displacement_clean_point_postord_50)
displacement_matched_point_postord_100 <- create.matched.data(displacement_change_true_100, displacement_clean_point_postord_100)
displacement_matched_sample_postord_50 <- create.matched.data(displacement_change_true_50, displacement_clean_sample_postord_50)
displacement_matched_sample_postord_100 <- create.matched.data(displacement_change_true_100, displacement_clean_sample_postord_100)
displacement_matched_no_ace_50 <- create.matched.data(displacement_change_true_50, displacement_clean_no_ace_50)
displacement_matched_no_ace_100 <- create.matched.data(displacement_change_true_100, displacement_clean_no_ace_100)



combine_datasets <- function(data_50, data_100) {
  Map(function(rate_50, rate_100) {
    Map(function(treatment_50, treatment_100) {
      c(treatment_50, treatment_100)
    }, rate_50, rate_100)
  }, data_50, data_100)
}

# Sum of variances
sum_var_matched_sample <- combine_datasets(sum_var_matched_sample_50, sum_var_matched_sample_100)
sum_var_matched_strict <- combine_datasets(sum_var_matched_strict_50, sum_var_matched_strict_100)
sum_var_matched_point_postord <- combine_datasets(sum_var_matched_point_postord_50, sum_var_matched_point_postord_100)
sum_var_matched_sample_postord <- combine_datasets(sum_var_matched_sample_postord_50, sum_var_matched_sample_postord_100)
sum_var_matched_no_ace <- combine_datasets(sum_var_matched_no_ace_50, sum_var_matched_no_ace_100)

# Sum of quantiles
sum_quant_matched_sample <- combine_datasets(sum_quant_matched_sample_50, sum_quant_matched_sample_100)
sum_quant_matched_strict <- combine_datasets(sum_quant_matched_strict_50, sum_quant_matched_strict_100)
sum_quant_matched_point_postord <- combine_datasets(sum_quant_matched_point_postord_50, sum_quant_matched_point_postord_100)
sum_quant_matched_sample_postord <- combine_datasets(sum_quant_matched_sample_postord_50, sum_quant_matched_sample_postord_100)
sum_quant_matched_no_ace <- combine_datasets(sum_quant_matched_no_ace_50, sum_quant_matched_no_ace_100)

# Displacement
displacement_matched_sample <- combine_datasets(displacement_matched_sample_50, displacement_matched_sample_100)
displacement_matched_strict <- combine_datasets(displacement_matched_strict_50, displacement_matched_strict_100)
displacement_matched_point_postord <- combine_datasets(displacement_matched_point_postord_50, displacement_matched_point_postord_100)
displacement_matched_sample_postord <- combine_datasets(displacement_matched_sample_postord_50, displacement_matched_sample_postord_100)
displacement_matched_no_ace <- combine_datasets(displacement_matched_no_ace_50, displacement_matched_no_ace_100)

calculate.diffs <- function(matched_data) {
  lapply(matched_data, function(rate_data) {
    rate_data$diffs <- rate_data$estimated - rate_data$true
    return(rate_data)
  })
}

# calculate.diffs.sign <- function(matched_data) {
#   lapply(matched_data, function(rate_data) {
#     # Assuming rate_data is a list of "estimated" and "true" vectors
#     rate_data$diffs <- rate_data$estimated - rate_data$true
#     rate_data$wrong_dir <- sign(rate_data$estimated) != sign(rate_data$true)
    
#     # Keep only entries where direction is wrong
#     rate_data$estimated <- rate_data$estimated[rate_data$wrong_dir]
#     rate_data$true <- rate_data$true[rate_data$wrong_dir]
#     rate_data$diffs <- rate_data$diffs[rate_data$wrong_dir]
    
#     return(rate_data)
#   })
# }

# Apply to all datasets
all_matched_datasets <- list(
  sum_var_matched_sample, sum_var_matched_strict, sum_var_matched_point_postord, 
  sum_var_matched_sample_postord, sum_var_matched_no_ace,
  sum_quant_matched_sample, sum_quant_matched_strict, sum_quant_matched_point_postord, 
  sum_quant_matched_sample_postord, sum_quant_matched_no_ace,
  displacement_matched_sample, displacement_matched_strict, displacement_matched_point_postord, 
  displacement_matched_sample_postord, displacement_matched_no_ace
)

# Calculate diffs for all
all_matched_datasets <- lapply(all_matched_datasets, calculate.diffs)
list2env(setNames(all_matched_datasets, c(
  "sum_var_matched_sample", "sum_var_matched_strict", "sum_var_matched_point_postord", 
  "sum_var_matched_sample_postord", "sum_var_matched_no_ace",
  "sum_quant_matched_sample", "sum_quant_matched_strict", "sum_quant_matched_point_postord", 
  "sum_quant_matched_sample_postord", "sum_quant_matched_no_ace",
  "displacement_matched_sample", "displacement_matched_strict", "displacement_matched_point_postord", 
  "displacement_matched_sample_postord", "displacement_matched_no_ace"
)), envir = .GlobalEnv)


library(ggridges)
library(ggplot2)
library(tidyr)

# Create data for all three metrics and all methods
create_ridge_data <- function(matched_data, metric_name) {
  data.frame(
    diffs = c(matched_data$slow$diffs, matched_data$med$diffs, matched_data$fast$diffs),
    fossil_level = c(matched_data$slow$fossil_level, matched_data$med$fossil_level, matched_data$fast$fossil_level),
    rate = rep(c("slow", "med", "fast"), 
               c(length(matched_data$slow$diffs), length(matched_data$med$diffs), length(matched_data$fast$diffs))),
    metric = metric_name
  )
}

# Combine all data for all methods
sample_data_all <- rbind(
  create_ridge_data(sum_var_matched_sample, "Sum of Variances"),
  create_ridge_data(sum_quant_matched_sample, "Sum of Quantiles"),
  create_ridge_data(displacement_matched_sample, "Displacement")
)
sample_data_all$method <- "Pre-ord ACE (dist)"

strict_data_all <- rbind(
  create_ridge_data(sum_var_matched_strict, "Sum of Variances"),
  create_ridge_data(sum_quant_matched_strict, "Sum of Quantiles"), 
  create_ridge_data(displacement_matched_strict, "Displacement")
)
strict_data_all$method <- "Pre-ord ACE (point)"

point_postord_data_all <- rbind(
  create_ridge_data(sum_var_matched_point_postord, "Sum of Variances"),
  create_ridge_data(sum_quant_matched_point_postord, "Sum of Quantiles"), 
  create_ridge_data(displacement_matched_point_postord, "Displacement")
)
point_postord_data_all$method <- "Post-ord ACE (point)"

sample_postord_data_all <- rbind(
  create_ridge_data(sum_var_matched_sample_postord, "Sum of Variances"),
  create_ridge_data(sum_quant_matched_sample_postord, "Sum of Quantiles"), 
  create_ridge_data(displacement_matched_sample_postord, "Displacement")
)
sample_postord_data_all$method <- "Post-ord ACE (dist)"

no_ace_data_all <- rbind(
  create_ridge_data(sum_var_matched_no_ace, "Sum of Variances"),
  create_ridge_data(sum_quant_matched_no_ace, "Sum of Quantiles"), 
  create_ridge_data(displacement_matched_no_ace, "Displacement")
)
no_ace_data_all$method <- "No ACE"

# Combine all methods
combined_ridge_data <- rbind(sample_data_all, strict_data_all, point_postord_data_all, 
                            sample_postord_data_all, no_ace_data_all)

# Filter out living data and set factor levels
level_order <- c("fossil_low", "fossil_med", "fossil_high", "all")
combined_ridge_data_filtered <- combined_ridge_data[combined_ridge_data$fossil_level != "living", ]
combined_ridge_data_filtered$fossil_level <- factor(combined_ridge_data_filtered$fossil_level, 
                                                   levels = level_order,
                                                   labels = c("5%", "15%", "50%", "100%"))

combined_ridge_data_filtered$metric <- factor(combined_ridge_data_filtered$metric, 
                                             levels = c("Sum of Variances", "Sum of Quantiles", "Displacement"))

combined_ridge_data_filtered$method <- factor(combined_ridge_data_filtered$method,
                                             levels = c("Pre-ord ACE (point)", "Pre-ord ACE (dist)", 
                                                       "Post-ord ACE (point)", "Post-ord ACE (dist)", "No ACE"))

# Color scheme
colours <- c("slow" = "#8805A8", "med" = "#00B945", "fast" = "#FFB600")

# combined_ridge_data_filtered <- combined_ridge_data_filtered %>%
#   group_by(method) %>%
#   mutate(weight_method = 1 / n()) %>%  # weight = 1 / total points in method
# #   ungroup()

# randext_ridge_plot <- ggplot(combined_ridge_data_filtered, aes(x = diffs, y = fossil_level, fill = rate)) +
#   geom_density_ridges(
#   alpha = 0.7,
#   scale = 1,          # controls how tall ridges are relative to each other
#   rel_min_height = 0.001
# ) +
#   geom_vline(xintercept = 0, color = "black", linewidth = 0.6, linetype = "dashed") +
#   facet_grid(metric ~ method) +
#   labs(
#     x = "Estimation Error",
#     y = NULL,
#     fill = NULL,
#     title = NULL,
#     subtitle = NULL
#   ) +
#   theme_minimal() +
#   scale_fill_manual(values = colours) + 
#   theme(
#     axis.text = element_text(size = 14, color = "black"),
#     axis.text.x = element_text(angle = 0),
#     axis.title.y = element_blank(),
#     axis.title.x = element_text(size = 18, face = "bold", color = "black"),
#     plot.title = element_blank(),
#     plot.subtitle = element_blank(),
#     legend.position = "none",
#     strip.text = element_text(size = 12, face = "bold", color = "black"),  # Add this back
#     strip.background = element_rect(fill = "white", color = "black"),      # Optional: add background
#     panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4),
#     panel.grid = element_blank(),
#     panel.spacing = unit(0.7, "cm"),
#     plot.margin = margin(20, 20, 20, 20),
#     panel.background = element_rect(fill = NA, color = NA),
#     plot.background = element_rect(fill = NA, color = NA)
#   ) +
#   scale_x_continuous(
#     breaks = c(-1.0, -0.5, 0, 0.5, 1.0),
#     labels = c("-1.0", "-0.5", "0", "0.5", "1.0"),
#     expand = expansion(mult = c(0.02, 0.02))
#   ) +
#   coord_cartesian(xlim = c(-1, 1.0)) +
#   scale_y_discrete(expand = expansion(mult = c(0, 0.1)))






















# plot_data <- data.frame(
#   diffs = c(matched_sample$slow$diffs, matched_sample$med$diffs, matched_sample$fast$diffs),
#   fossil_level = c(matched_sample$slow$fossil_level, matched_sample$med$fossil_level, matched_sample$fast$fossil_level),
#   rate = rep(c("slow", "med", "fast"), 
#              c(length(matched_sample$slow$diffs), length(matched_sample$med$diffs), length(matched_sample$fast$diffs))),
#   method = "Sample ACE"
# )








# Create data for all three metrics - Sample ACE only
sample_data_all <- rbind(
    create_ridge_data(sum_quant_matched_sample, "Sum of quantiles"),
    create_ridge_data(sum_var_matched_sample, "Sum of variances"),
    create_ridge_data(displacement_matched_sample, "Average displacement")
)
sample_data_all$method <- "Pre-ord ACE (dist)"

# Create data for all three metrics - No ACE only
no_ace_data_all <- rbind(
      create_ridge_data(sum_quant_matched_no_ace, "Sum of quantiles"), 
  create_ridge_data(sum_var_matched_no_ace, "Sum of variances"),
  create_ridge_data(displacement_matched_no_ace, "Average displacement")
)
no_ace_data_all$method <- "No ACE"

# Combine just these two methods
combined_ridge_data <- rbind(sample_data_all, no_ace_data_all)

# Filter out living data and set factor levels
level_order <- c("fossil_low", "fossil_med", "fossil_high", "all")
combined_ridge_data_filtered <- combined_ridge_data[combined_ridge_data$fossil_level != "living", ]
combined_ridge_data_filtered$fossil_level <- factor(combined_ridge_data_filtered$fossil_level, 
                                                   levels = level_order,
                                                   labels = c("5%", "15%", "50%", "100%"))

combined_ridge_data_filtered$metric <- factor(combined_ridge_data_filtered$metric, 
                                             levels = c("Sum of quantiles", "Sum of variances", "Average displacement"))

combined_ridge_data_filtered$method <- factor(combined_ridge_data_filtered$method,
                                             levels = c("Pre-ord ACE (dist)", "No ACE"))


bw_table <- combined_ridge_data_filtered %>%  ### need to calculate bandwidth per treatment, due to different sample sizes
  group_by(method, metric, rate) %>%
  summarise(
    n_points = n(),
    sd_diffs = sd(diffs, na.rm = TRUE),
    bw = 1.06 * sd_diffs * n_points^(-1/5),
    .groups = "drop"
  )


# Color scheme
colours <- c("slow" = "#8805A8", "med" = "#00B945", "fast" = "#FFB600")

# Create the ridge plot for just sample vs no ace
ridge_data <- combined_ridge_data_filtered %>%
  left_join(bw_table, by = c("method", "metric", "rate"))

ridge_data_grouped <- ridge_data %>%
  group_by(method, metric, rate) %>%
  mutate(bw_group = first(bw)) %>%   # pick one bw per group
  ungroup()


# Corrected ggplot call
rand_ext_ridge <- ggplot(ridge_data, aes(x = diffs, y = fossil_level, fill = rate)) +
  geom_density_ridges(
    alpha = 0.7,
    scale = 1,
    rel_min_height = 0.001,
    size = 0.1
    # The line 'stat = "density"' has been removed
  ) +
  facet_grid(metric ~ method + rate) +
  labs(
    x = "Relative change in disparity error",
    y = "Fossil sampling",
    fill = NULL
  ) +
  scale_fill_manual(values = c("slow" = "#8805A8", "med" = "#00B945", "fast" = "#FFB600")) +
  theme_minimal(base_size = 14) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.6, linetype = "dashed") +
  theme(
    axis.text = element_text(size = 14, color = "black"),
    axis.text.x = element_text(angle = 0),
    axis.title.y = element_blank(),
    axis.title.x = element_text(size = 18, face = "bold", color = "black"),
    plot.title = element_blank(),
    strip.text.y = element_text(face = "bold", size = 14),
    plot.subtitle = element_blank(),
    legend.position = "none",
    strip.text = element_blank(),
    strip.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.4),
    panel.grid = element_blank(),
    panel.spacing = unit(0.7, "cm"),
    plot.margin = margin(20, 20, 20, 20),
    panel.background = element_rect(fill = NA, color = NA),  # Make panel transparent
    plot.background = element_rect(fill = NA, color = NA)
  ) +
  scale_x_continuous(
    breaks = c(-0.5, 0, 0.5),
    labels = c("-0.5", "0", "0.5")
  ) +
  coord_cartesian(xlim = c(-0.5, 0.5)) 

# This should now run without error
print(rand_ext_ridge)


# Save the plot
ggsave("../Manuscript/figures/rand_ext_ridge_pre_ord_no_ace.png", 
       rand_ext_ridge, 
       width = 11, height = 8, dpi = 700, bg = "transparent")



























sample_data_50 <- data.frame(
  diffs = c(matched_sample_50$slow$diffs, matched_sample_50$med$diffs, matched_sample_50$fast$diffs),
  fossil_level = c(matched_sample_50$slow$fossil_level, matched_sample_50$med$fossil_level, matched_sample_50$fast$fossil_level),
  rate = rep(c("slow", "med", "fast"), 
             c(length(matched_sample_50$slow$diffs), length(matched_sample_50$med$diffs), length(matched_sample_50$fast$diffs))),
  method = "Sample ACE"
)

# Add no_ace data
no_ace_data_50 <- data.frame(
  diffs = c(matched_no_ace_50$slow$diffs, matched_no_ace_50$med$diffs, matched_no_ace_50$fast$diffs),
  fossil_level = c(matched_no_ace_50$slow$fossil_level, matched_no_ace_50$med$fossil_level, matched_no_ace_50$fast$fossil_level),
  rate = rep(c("slow", "med", "fast"), 
             c(length(matched_no_ace_50$slow$diffs), length(matched_no_ace_50$med$diffs), length(matched_no_ace_50$fast$diffs))),
  method = "No ACE"
)

combined_data_50 <- rbind(sample_data_50, no_ace_data_50)


# Set the order for fossil levels (low to high preservation, then all)
level_order <- c("living", "fossil_low", "fossil_med", "fossil_high", "all")

# Convert fossil_level to factor with specified order in your combined_data
combined_data_50$fossil_level <- factor(combined_data_50$fossil_level, levels = level_order)

# Now your ridge plot will display in the correct order
ggplot(combined_data_50, aes(x = diffs, y = interaction(fossil_level, method), fill = method)) +
  geom_density_ridges(alpha = 0.7, scale = 1.2) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 1) +
  scale_fill_viridis_d(option = "plasma", end = 0.8) +
  facet_wrap(~rate, scales = "free_x") +
  theme_ridges() +
  theme(legend.position = "bottom") +
  labs(title = "Ridge Plot: Distribution of Estimation Errors",
       subtitle = "Higher peaks = more concentrated errors",
       x = "Estimation Error (Estimated - True)",
       y = "Fossil Level × Method",
       fill = "Method")

# library(viridis)
# fossil_colors <- viridis(5, option = "plasma", direction = -1)
# names(fossil_colors) <- c("all", "fossil_high", "fossil_med", "fossil_low", "living")

# plot(
#   c(matched_sample$slow$true, matched_sample$med$true, matched_sample$fast$true),
#   c(matched_sample$slow$estimated, matched_sample$med$estimated, matched_sample$fast$estimated),
#   xlab = "True Disparity Change",
#   ylab = "Est Disparity Change",
#   main = "Disparity Change: No majority Rule vs True",
#   pch = 16,
#   col = fossil_colors[
#     c(matched_sample$slow$fossil_level, matched_sample$med$fossil_level, matched_sample$fast$fossil_level)
#   ],
#   cex = 1.7,
#   ylim = c(-1, 1)
# )

# # Add 1:1 reference line
# abline(0, 1, col = "red", lty = 2, lwd = 2)

# plot(c(matched_no_ace$slow$true, matched_no_ace$med$true, matched_no_ace$fast$true),
#   c(matched_no_ace$slow$estimated, matched_no_ace$med$estimated, matched_no_ace$fast$estimated),
#     xlab = "True Disparity Change",
#     ylab = "Estimated Disparity Change",
#     main = "Disparity Change: No ace vs True",
#     pch = 16, 
#     col = fossil_colors[
#     c(matched_no_ace$slow$fossil_level, matched_no_ace$med$fossil_level, matched_no_ace$fast$fossil_level)
#   ],
#     cex = 1.7,
#     ylim = c(-1, 1))

# abline(0, 1, col = "red", lty = 2, lwd = 2)


```

## test of how many are within 10% of true


```{R}

# sum_var_change_true_ranges <- lapply(sum_var_change_true, lapply, function (x){
#     range <- x * 0.1
#     return(c((x- range), (x+range)))
# })

# coverage_no_majority <- Map(function(sample_rate, true_rate){
#     Map(function(sample_rep, true_rep){
#         true_vals <- true_rep
#         lapply(sample_rep, function(fossil){
#             sapply(fossil, function(v){
#                 is_within <- v >= min(true_vals) && v <= max(true_vals)
#                 return(is_within)
#             })
#         })
#     }, sample_rate, true_rate)
# }, clean_sample, sum_var_change_true_ranges)

# coverage_no_ace <- Map(function(sample_rate, true_rate){
#     Map(function(sample_rep, true_rep){
#         true_vals <- true_rep
#         lapply(sample_rep, function(fossil){
#             is_within <- fossil >= min(true_vals) && fossil <= max(true_vals)
#             return(is_within)
#         })
#     }, sample_rate, true_rate)
# }, clean_no_ace, sum_var_change_true_ranges)


```




# 100t 
## sum var

```{R}


rate_names <- c("slow", "med", "fast")

# Load sample data (keep as is)
sum_var_change_sample_100 <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_sample_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)

sum_var_change_no_ace_100 <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_no_ace_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)



# Load true data (keep as is)
sum_var_change_true_100 <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_true_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)


# Calculate median on filtered sample data
# sum_var_change_sample_median <- lapply(sum_var_change_sample, function(rate_data) {
#     lapply(rate_data, function(replicate_data) {
#         # Apply median directly to each fossil level
#         result <- lapply(replicate_data, function(fossil_level_values) {
#             values <- unlist(fossil_level_values, use.names = FALSE)
#             return(median(values, na.rm = TRUE))
#         })
#         return(result)  
#     })
# })
sum_var_change_sample_unlist_100 <- lapply(sum_var_change_sample, function(rate_data) {
    lapply(rate_data, function(replicate_data) {
        # Apply median directly to each fossil level
        result <- lapply(replicate_data, function(fossil_level_values) {
            values <- unlist(fossil_level_values, use.names = FALSE)
            return(values)
        })
        return(result)  
    })
})

clean_sample_100 <- lapply(sum_var_change_sample_unlist_100, lapply, function(x){
    x[!sapply(x, is.null)]  
})

clean_no_ace_100 <- lapply(sum_var_change_no_ace_100, lapply, function(x){
    x[!sapply(x, is.null)]  
})




# Test the match
matched_sample_100 <- create.matched.data(sum_var_change_true_100, clean_sample_100)
matched_no_ace_100 <- create.matched.data(sum_var_change_true_100, clean_no_ace_100)




# Calculate differences for sample method
matched_sample_100$slow$diffs  <- matched_sample_100$slow$estimated - matched_sample_100$slow$true
matched_sample_100$med$diffs   <- matched_sample_100$med$estimated - matched_sample_100$med$true
matched_sample_100$fast$diffs  <- matched_sample_100$fast$estimated - matched_sample_100$fast$true

# Calculate differences for no_ace method
matched_no_ace_100$slow$diffs  <- matched_no_ace_100$slow$estimated - matched_no_ace_100$slow$true
matched_no_ace_100$med$diffs   <- matched_no_ace_100$med$estimated - matched_no_ace_100$med$true
matched_no_ace_100$fast$diffs  <- matched_no_ace_100$fast$estimated - matched_no_ace_100$fast$true



# matched_sample_100$slow$diffs  <- matched_sample$slow$estimated - matched_sample$slow$true
# matched_sample$med$diffs  <- matched_sample$med$estimated - matched_sample$med$true
# matched_sample$fast$diffs  <- matched_sample$fast$estimated - matched_sample$fast$true


# matched_no_ace <- create.matched.data(sum_var_change_true, clean_no_ace)
# matched_no_ace$slow$diffs  <- matched_no_ace$slow$estimated - matched_no_ace$slow$true
# matched_no_ace$med$diffs  <- matched_no_ace$med$estimated - matched_no_ace$med$true
# matched_no_ace$fast$diffs  <- matched_no_ace$fast$estimated - matched_no_ace$fast$true
# all_ace <- c(matched_sample$estimated, matched_no_ace$estimated)


sample_data_100 <- data.frame(
  diffs = c(matched_sample_100$slow$diffs, matched_sample_100$med$diffs, matched_sample_100$fast$diffs),
  fossil_level = c(matched_sample_100$slow$fossil_level, matched_sample_100$med$fossil_level, matched_sample_100$fast$fossil_level),
  rate = rep(c("slow", "med", "fast"), 
             c(length(matched_sample_100$slow$diffs), length(matched_sample_100$med$diffs), length(matched_sample_100$fast$diffs))),
  method = "Sample ACE"
)

# Add no_ace data
no_ace_data_100 <- data.frame(
  diffs = c(matched_no_ace_100$slow$diffs, matched_no_ace_100$med$diffs, matched_no_ace_100$fast$diffs),
  fossil_level = c(matched_no_ace_100$slow$fossil_level, matched_no_ace_100$med$fossil_level, matched_no_ace_100$fast$fossil_level),
  rate = rep(c("slow", "med", "fast"), 
             c(length(matched_no_ace_100$slow$diffs), length(matched_no_ace_100$med$diffs), length(matched_no_ace_100$fast$diffs))),
  method = "No ACE"
)

combined_data_100 <- rbind(sample_data_100, no_ace_data_100)


# Set the order for fossil levels (low to high preservation, then all)
level_order <- c("living", "fossil_low", "fossil_med", "fossil_high", "all")

# Convert fossil_level to factor with specified order in your combined_data
combined_data_100$fossil_level <- factor(combined_data_100$fossil_level, levels = level_order)

# Now your ridge plot will display in the correct order
ggplot(combined_data_100, aes(x = diffs, y = interaction(fossil_level, method), fill = method)) +
  geom_density_ridges(alpha = 0.7, scale = 1.2) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 1) +
  scale_fill_viridis_d(option = "plasma", end = 0.8) +
  facet_wrap(~rate, scales = "free_x") +
  theme_ridges() +
  theme(legend.position = "bottom") +
  labs(title = "Ridge Plot: Distribution of Estimation Errors",
       subtitle = "Higher peaks = more concentrated errors",
       x = "Estimation Error (Estimated - True)",
       y = "Fossil Level × Method",
       fill = "Method")


library(viridis)
fossil_colors <- viridis(5, option = "plasma", direction = -1)
names(fossil_colors) <- c("all", "fossil_high", "fossil_med", "fossil_low", "living")

level_order <- c("living", "fossil_low", "fossil_med", "fossil_high", "all")

# # Combine all data
# all_true <- c(matched_sample$slow$true, matched_sample$med$true, matched_sample$fast$true)
# all_est <- c(matched_sample$slow$estimated, matched_sample$med$estimated, matched_sample$fast$estimated)
# all_fossils <- c(matched_sample$slow$fossil_level, matched_sample$med$fossil_level, matched_sample$fast$fossil_level)

# plot(all_true, all_est,
#      xlab = "True Disparity Change",
#      ylab = "Est Disparity Change", 
#      main = "Disparity Change: No majority Rule vs True",
#      type = "n", 
#      ylim = c(-0.6, 0.6))

# for(level in level_order) {
#   indices <- all_fossils == level
#   points(all_true[indices], all_est[indices],
#          pch = 20, col = fossil_colors[level], cex = 1.7)
# }

# abline(0, 1, col = "red", lty = 2, lwd = 2)
# legend("topleft", 
#        legend = level_order,  # Legend in same order as plotting
#        col = fossil_colors[level_order], 
#        pch = 20, cex = 1.2, title = "Fossil Level")


# plot(c(matched_no_ace$slow$true, matched_no_ace$med$true, matched_no_ace$fast$true),
#   c(matched_no_ace$slow$estimated, matched_no_ace$med$estimated, matched_no_ace$fast$estimated),
#     xlab = "True Disparity Change",
#     ylab = "Estimated Disparity Change",
#     main = "Disparity Change: No ace vs True",
#     pch = 20, 
#     col = fossil_colors[
#     c(matched_no_ace$slow$fossil_level, matched_no_ace$med$fossil_level, matched_no_ace$fast$fossil_level)
#   ],
#     cex = 1.7,
#     ylim = c(-0.6, 0.6))
# abline(0, 1, col = "red", lty = 2, lwd = 2)

```

## Combined 50-100 t

```{R}


matched_sample <- Map(function(rate_50, rate_100){
    Map(function(treatment_50, treatment_100){
        c(treatment_50, treatment_100)
    }, rate_50, rate_100)
}, matched_sample_50, matched_sample_100)

matched_no_ace <- Map(function(rate_50, rate_100){
    Map(function(treatment_50, treatment_100){
        c(treatment_50, treatment_100)
    }, rate_50, rate_100)
}, matched_no_ace_50, matched_no_ace_100)

# Calculate differences for sample method
matched_sample$slow$diffs  <- matched_sample$slow$estimated - matched_sample$slow$true
matched_sample$med$diffs   <- matched_sample$med$estimated - matched_sample$med$true
matched_sample$fast$diffs  <- matched_sample$fast$estimated - matched_sample$fast$true

# Calculate differences for no_ace method
matched_no_ace$slow$diffs  <- matched_no_ace$slow$estimated - matched_no_ace$slow$true
matched_no_ace$med$diffs   <- matched_no_ace$med$estimated - matched_no_ace$med$true
matched_no_ace$fast$diffs  <- matched_no_ace$fast$estimated - matched_no_ace$fast$true
plot_data <- data.frame(
  diffs = c(matched_sample$slow$diffs, matched_sample$med$diffs, matched_sample$fast$diffs),
  fossil_level = c(matched_sample$slow$fossil_level, matched_sample$med$fossil_level, matched_sample$fast$fossil_level),
  rate = rep(c("slow", "med", "fast"), 
             c(length(matched_sample$slow$diffs), length(matched_sample$med$diffs), length(matched_sample$fast$diffs))),
  method = "Sample ACE"
)

# Add no_ace data
no_ace_data <- data.frame(
  diffs = c(matched_no_ace$slow$diffs, matched_no_ace$med$diffs, matched_no_ace$fast$diffs),
  fossil_level = c(matched_no_ace$slow$fossil_level, matched_no_ace$med$fossil_level, matched_no_ace$fast$fossil_level),
  rate = rep(c("slow", "med", "fast"), 
             c(length(matched_no_ace$slow$diffs), length(matched_no_ace$med$diffs), length(matched_no_ace$fast$diffs))),
  method = "No ACE"
)

combined_data <- rbind(plot_data, no_ace_data)


# Set the order for fossil levels (low to high preservation, then all)
level_order <- c("living", "fossil_low", "fossil_med", "fossil_high", "all")

# Convert fossil_level to factor with specified order in your combined_data
combined_data$fossil_level <- factor(combined_data$fossil_level, levels = level_order)

# Now your ridge plot will display in the correct order
ggplot(combined_data, aes(x = diffs, y = interaction(fossil_level, method), fill = method)) +
  geom_density_ridges(alpha = 0.7, scale = 1.2) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 1) +
  scale_fill_viridis_d(option = "plasma", end = 0.8) +
  facet_wrap(~rate, scales = "free_x") +
  theme_ridges() +
  theme(legend.position = "bottom") +
  labs(title = "Distribution of error in change in disparity",
       x = "Estimation Error (Estimated - True)",
       fill = "Method")





# create.lm.model <- function(matched, fossil) {
#     indices <- matched$fossil_level == fossil
#     estimated <- matched$estimated[indices]
#     true <- matched$true[indices]
#     model <- lm(estimated ~ true)
#     # Return a named list so you can use $all, $fossil_high, etc.
#     result <- list(estimated = estimated, true = true, model = model)
#     names(result) <- c("estimated", "true", "model")
#     return(result)
# }


# matched_sample_fossil_high <- create.lm.model(matched_sample$med, "fossil_low")
# matched_no_ace_fossil_high <- create.lm.model(matched_no_ace$med, "fossil_low")


# summary(matched_sample_fossil_high$model)
# summary(matched_no_ace_fossil_high$model)



# all_indices <- matched_sample$fossil_level == "all"
# matched_sample$all$estimated <- matched_sample$estimated[all_indices]
# matched_sample$all$true <- matched_sample$true[all_indices]

# all_indices <- matched_no_ace$fossil_level == "all"
# matched_no_ace$all$estimated <- matched_no_ace$estimated[all_indices]
# matched_no_ace$all$true <- matched_no_ace$true[all_indices]





# all_model_sample <- lm(matched_sample$all$estimated ~ matched_sample$all$true)
# all_model_no_ace <- lm(matched_no_ace$all$estimated ~ matched_no_ace$all$true)




```

## sum quant
```{r}


# 100t sum var

rate_names <- c("slow", "med", "fast")

# Load sample data (keep as is)
sum_quant_change_sample <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8100947_sample_sum_quant_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)

sum_quant_change_no_ace <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8100947_no_ace_sum_quant_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)



# Load true data (keep as is)
sum_quant_change_true <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8100947_true_sum_quant_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)


# Calculate median on filtered sample data
sum_quant_change_sample_median <- lapply(sum_quant_change_sample, function(rate_data) {
    lapply(rate_data, function(replicate_data) {
        # Apply median directly to each fossil level
        result <- lapply(replicate_data, function(fossil_level_values) {
            values <- unlist(fossil_level_values, use.names = FALSE)
            return(median(values, na.rm = TRUE))
        })
        return(result)  # This ensures we return the fossil level list directly
    })
})

clean_sample <- lapply(sum_quant_change_sample_median, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})

clean_no_ace <- lapply(sum_quant_change_no_ace, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})




# Test the match
matched_sample <- create.matched.data(sum_quant_change_true, clean_sample)
matched_no_ace <- create.matched.data(sum_quant_change_true, clean_no_ace)

all_ace <- c(matched_sample$estimated, matched_no_ace$estimated)



library(viridis)
fossil_colors <- viridis(5, option = "plasma", direction = -1)
names(fossil_colors) <- c("all", "fossil_high", "fossil_med", "fossil_low", "living")

plot(c(matched_sample$slow$true), c(matched_sample$slow$estimated),
    xlab = "True Disparity Change",
    ylab = "Est Disparity Change",
    main = "Disparity Change: No majority Rule vs True",
    pch = 16, 
    col = fossil_colors[matched_sample$fast$fossil_level],
    cex = 1.7
)

# Add 1:1 reference line
abline(0, 1, col = "red", lty = 2, lwd = 2)

plot(matched_no_ace$slow$true, matched_no_ace$slow$estimated,
    xlab = "True Disparity Change",
    ylab = "Estimated Disparity Change",
    main = "Disparity Change: No ace vs True",
    pch = 16, 
    col = fossil_colors[matched_no_ace$slow$fossil_level],
    cex = 1.7
)

abline(0, 1, col = "red", lty = 2, lwd = 2)




```




## Displacement
```{r}


# 100t sum var

rate_names <- c("slow", "med", "fast")

# Load sample data (keep as is)
displacement_change_sample <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8100947_sample_displacement_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)

displacement_change_no_ace <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8100947_no_ace_displacement_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)



# Load true data (keep as is)
displacement_change_true <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8100947_true_displacement_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)


# Calculate median on filtered sample data
displacement_change_sample_median <- lapply(displacement_change_sample, function(rate_data) {
    lapply(rate_data, function(replicate_data) {
        # Apply median directly to each fossil level
        result <- lapply(replicate_data, function(fossil_level_values) {
            values <- unlist(fossil_level_values, use.names = FALSE)
            return(median(values, na.rm = TRUE))
        })
        return(result)  # This ensures we return the fossil level list directly
    })
})

clean_sample <- lapply(displacement_change_sample_median, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})

clean_no_ace <- lapply(displacement_change_no_ace, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})




# Test the match
matched_sample <- create.matched.data(displacement_change_true, clean_sample)
matched_no_ace <- create.matched.data(displacement_change_true, clean_no_ace)

all_ace <- c(matched_sample$estimated, matched_no_ace$estimated)



library(viridis)
fossil_colors <- viridis(5, option = "plasma", direction = -1)
names(fossil_colors) <- c("all", "fossil_high", "fossil_med", "fossil_low", "living")

plot(c(matched_sample$slow$true, matched_sample$med$true, matched_sample$fast$true), c(matched_sample$slow$estimated, matched_sample$med$estimated, matched_sample$fast$estimated),
    xlab = "True Disparity Change",
    ylab = "Est Disparity Change",
    main = "Disparity Change: No majority Rule vs True",
    pch = 16, 
    col = fossil_colors[matched_sample$fast$fossil_level],
    cex = 1.7
)

# Add 1:1 reference line
abline(0, 1, col = "red", lty = 2, lwd = 2)

plot(c(matched_no_ace$slow$true, matched_no_ace$med$true, matched_no_ace$fast$true), c(matched_no_ace$slow$estimated, matched_no_ace$med$estimated, matched_no_ace$fast$estimated),
    xlab = "True Disparity Change",
    ylab = "Estimated Disparity Change",
    main = "Disparity Change: No ace vs True",
    pch = 16, 
    col = fossil_colors[c(matched_no_ace$slow$fossil_level, matched_no_ace$med$fossil_level, matched_no_ace$fast$fossil_level)],
    cex = 1.7
)

abline(0, 1, col = "red", lty = 2, lwd = 2)
legend("topleft", 
       legend = names(fossil_colors),
       col = fossil_colors,
       pch = 16,
       cex = 1.2,
       title = "Fossil Level")



```

# 50t with higher turnover and 

```{R}


# 50t
```{R}
rate_names <- c("slow", "med", "fast")

# Load sample data (keep as is)
sum_var_change_sample <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8115692_sample_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)

sum_var_change_strict <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8115692_strict_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)

sum_var_change_no_ace <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8115692_no_ace_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)



# Load true data (keep as is)
sum_var_change_true <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8115692_true_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)




# Calculate median on filtered sample data
sum_var_change_sample_median <- lapply(sum_var_change_sample, function(rate_data) {
    lapply(rate_data, function(replicate_data) {
        # Apply median directly to each fossil level
        result <- lapply(replicate_data, function(fossil_level_values) {
            values <- unlist(fossil_level_values, use.names = FALSE)
            return(median(values, na.rm = TRUE))
        })
        return(result)  # This ensures we return the fossil level list directly
    })
})

sum_var_change_sample_unlist <- lapply(sum_var_change_sample, function(rate_data) {
    lapply(rate_data, function(replicate_data) {
        # Apply median directly to each fossil level
        result <- lapply(replicate_data, function(fossil_level_values) {
            values <- unlist(fossil_level_values, use.names = FALSE)
            return(values)
        })
        return(result)  # This ensures we return the fossil level list directly
    })
})



clean_sample <- lapply(sum_var_change_sample_unlist, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})

clean_no_ace <- lapply(sum_var_change_no_ace, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})



clean_strict <- lapply(sum_var_change_strict, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})
# 
#Use filtered true data
# sum_var_change_true <- sum_var_change_true_filtered

# Your function should now work perfectly
create.matched.data <- function(true, estimated) {
    result <- list()
    for(rate_name in names(estimated)) {
        true_matched <- c()
        est_matched <- c()
        fossil_levels <- c()
        for(rep in seq_along(estimated[[rate_name]])) {
            true_val <- true[[rate_name]][[rep]]
            est_rep_val <- estimated[[rate_name]][[rep]]
            fossil_names <- names(est_rep_val)
            # Loop over fossil levels
            for(i in seq_along(est_rep_val)) {
                est_vals <- est_rep_val[[i]]
                n_est <- length(est_vals)
                true_matched <- c(true_matched, rep(true_val, n_est))
                est_matched <- c(est_matched, est_vals)
                fossil_levels <- c(fossil_levels, rep(fossil_names[i], n_est))
            }
        }
        result[[rate_name]] <- list(
            true = true_matched,
            estimated = est_matched,
            fossil_level = fossil_levels
        )
    }
    return(result)
}
matched_sample <- create.matched.data(sum_var_change_true, clean_sample)
matched_no_ace <- create.matched.data(sum_var_change_true, clean_no_ace)
matched_strict <- create.matched.data(sum_var_change_true, clean_strict)

all_ace <- c(matched_sample$estimated, matched_no_ace$estimated)



library(viridis)
fossil_colors <- viridis(5, option = "plasma", direction = -1)
names(fossil_colors) <- c("all", "fossil_high", "fossil_med", "fossil_low", "living")

plot(
  c(matched_sample$slow$true, matched_sample$med$true, matched_sample$fast$true),
  c(matched_sample$slow$estimated, matched_sample$med$estimated, matched_sample$fast$estimated),
  xlab = "True Disparity Change",
  ylab = "Est Disparity Change",
  main = "Disparity Change: No majority Rule vs True",
  pch = 20,
  col = fossil_colors[
    c(matched_sample$slow$fossil_level, matched_sample$med$fossil_level, matched_sample$fast$fossil_level)
  ],
  cex = 1.7,
  ylim = c(-0.6, 0.6)
)

# Add 1:1 reference line
abline(0, 1, col = "red", lty = 2, lwd = 2)

plot(c(matched_no_ace$slow$true, matched_no_ace$med$true, matched_no_ace$fast$true),
  c(matched_no_ace$slow$estimated, matched_no_ace$med$estimated, matched_no_ace$fast$estimated),
    xlab = "True Disparity Change",
    ylab = "Estimated Disparity Change",
    main = "Disparity Change: No ace vs True",
    pch = 20, 
    col = fossil_colors[
    c(matched_no_ace$slow$fossil_level, matched_no_ace$med$fossil_level, matched_no_ace$fast$fossil_level)
  ],
    cex = 1.7,
      ylim = c(-0.6, 0.6)
)

abline(0, 1, col = "red", lty = 2, lwd = 2)



plot(c(matched_strict$slow$true, matched_strict$med$true, matched_strict$fast$true),
  c(matched_strict$slow$estimated, matched_strict$med$estimated, matched_strict$fast$estimated),
    xlab = "True Disparity Change",
    ylab = "Estimated Disparity Change",
    main = "Disparity Change: Strict majority vs True",
    pch = 20, 
    col = fossil_colors[
    c(matched_strict$slow$fossil_level)
  ],
    cex = 1.7,
      ylim = c(-0.6, 0.6)
)

abline(0, 1, col = "red", lty = 2, lwd = 2)


```



# 50t with higher turnover and rm axes and 10my subsets

```{R}


# 50t
```{R}
rate_names <- c("slow", "med", "fast")

# Load sample data (keep as is)
sum_var_change_sample <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8120770_sample_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)

sum_var_change_strict <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8120770_strict_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)

sum_var_change_no_ace <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8120770_no_ace_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)



# Load true data (keep as is)
sum_var_change_true <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8120770_true_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)




# Calculate median on filtered sample data
sum_var_change_sample_median <- lapply(sum_var_change_sample, function(rate_data) {
    lapply(rate_data, function(replicate_data) {
        # Apply median directly to each fossil level
        result <- lapply(replicate_data, function(fossil_level_values) {
            values <- unlist(fossil_level_values, use.names = FALSE)
            return(median(values, na.rm = TRUE))
        })
        return(result)  # This ensures we return the fossil level list directly
    })
})

sum_var_change_sample_unlist <- lapply(sample_sum_var_change, function(rate_data) {
    lapply(rate_data, function(replicate_data) {
        # Apply median directly to each fossil level
        result <- lapply(replicate_data, function(fossil_level_values) {
            values <- unlist(fossil_level_values, use.names = FALSE)
            return(values)
        })
        return(result)  # This ensures we return the fossil level list directly
    })
})



clean_sample <- lapply(sum_var_change_sample_unlist, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})

clean_no_ace <- lapply(no_ace_sum_var_change, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})



clean_strict <- lapply(strict_sum_var_change, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})
# 
#Use filtered true data
# sum_var_change_true <- sum_var_change_true_filtered

# Your function should now work perfectly
create.matched.data <- function(true, estimated) {
    result <- list()
    for(rate_name in names(estimated)) {
        true_matched <- c()
        est_matched <- c()
        fossil_levels <- c()
        for(rep in seq_along(estimated[[rate_name]])) {
            true_val <- true[[rate_name]][[rep]]
            est_rep_val <- estimated[[rate_name]][[rep]]
            fossil_names <- names(est_rep_val)
            # Loop over fossil levels
            for(i in seq_along(est_rep_val)) {
                est_vals <- est_rep_val[[i]]
                n_est <- length(est_vals)
                true_matched <- c(true_matched, rep(true_val, n_est))
                est_matched <- c(est_matched, est_vals)
                fossil_levels <- c(fossil_levels, rep(fossil_names[i], n_est))
            }
        }
        result[[rate_name]] <- list(
            true = true_matched,
            estimated = est_matched,
            fossil_level = fossil_levels
        )
    }
    return(result)
}
matched_sample <- create.matched.data(true_sum_var_change, clean_sample)
matched_no_ace <- create.matched.data(true_sum_var_change, clean_no_ace)
matched_strict <- create.matched.data(true_sum_var_change, clean_strict)

all_ace <- c(matched_sample$estimated, matched_no_ace$estimated)



library(viridis)
fossil_colors <- viridis(5, option = "plasma", direction = -1)
names(fossil_colors) <- c("all", "fossil_high", "fossil_med", "fossil_low", "living")

plot(
  c(matched_sample$slow$true, matched_sample$med$true, matched_sample$fast$true),
  c(matched_sample$slow$estimated, matched_sample$med$estimated, matched_sample$fast$estimated),
  xlab = "True Disparity Change",
  ylab = "Est Disparity Change",
  main = "Disparity Change: No majority Rule vs True",
  pch = 20,
  col = fossil_colors[
    c(matched_sample$slow$fossil_level, matched_sample$med$fossil_level, matched_sample$fast$fossil_level)
  ],
  cex = 1.7,
  ylim = c(-0.6, 0.6)
)

# Add 1:1 reference line
abline(0, 1, col = "red", lty = 2, lwd = 2)

plot(c(matched_no_ace$slow$true, matched_no_ace$med$true, matched_no_ace$fast$true),
  c(matched_no_ace$slow$estimated, matched_no_ace$med$estimated, matched_no_ace$fast$estimated),
    xlab = "True Disparity Change",
    ylab = "Estimated Disparity Change",
    main = "Disparity Change: No ace vs True",
    pch = 20, 
    col = fossil_colors[
    c(matched_no_ace$slow$fossil_level, matched_no_ace$med$fossil_level, matched_no_ace$fast$fossil_level)
  ],
    cex = 1.7,
      ylim = c(-0.6, 0.6)
)

abline(0, 1, col = "red", lty = 2, lwd = 2)



plot(c(matched_strict$slow$true, matched_strict$med$true, matched_strict$fast$true),
  c(matched_strict$slow$estimated, matched_strict$med$estimated, matched_strict$fast$estimated),
    xlab = "True Disparity Change",
    ylab = "Estimated Disparity Change",
    main = "Disparity Change: Strict majority vs True",
    pch = 20, 
    col = fossil_colors[
    c(matched_strict$slow$fossil_level)
  ],
    cex = 1.7,
      ylim = c(-0.6, 0.6)
)

abline(0, 1, col = "red", lty = 2, lwd = 2)


```