---
title: "Random extinction plots"
author: "Caleb"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---


# 50t
```{R}
rate_names <- c("slow", "med", "fast")

# Load sample data (keep as is)
sum_var_change_sample_50 <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_sample_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)

sum_var_change_no_ace_50 <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_no_ace_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)



# Load true data (keep as is)
sum_var_change_true_50 <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8122585_true_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)




sum_var_change_sample_unlist_50 <- lapply(sum_var_change_sample_50, function(rate_data) {
    lapply(rate_data, function(replicate_data) {
        # Apply median directly to each fossil level
        result <- lapply(replicate_data, function(fossil_level_values) {
            values <- unlist(fossil_level_values, use.names = FALSE)
            return(values)
        })
        return(result)  # This ensures we return the fossil level list directly
    })
})



clean_sample_50 <- lapply(sum_var_change_sample_unlist_50, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})

clean_no_ace_50 <- lapply(sum_var_change_no_ace_50, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})

# Use filtered true data
# sum_var_change_true <- sum_var_change_true_filtered

# Your function should now work perfectly
create.matched.data <- function(true, estimated) {
    result <- list()
    for(rate_name in names(estimated)) {
        true_matched <- c()
        est_matched <- c()
        fossil_levels <- c()
        for(rep in seq_along(estimated[[rate_name]])) {
            true_val <- true[[rate_name]][[rep]]
            est_rep_val <- estimated[[rate_name]][[rep]]
            fossil_names <- names(est_rep_val)
            # Loop over fossil levels
            for(i in seq_along(est_rep_val)) {
                est_vals <- est_rep_val[[i]]
                n_est <- length(est_vals)
                true_matched <- c(true_matched, rep(true_val, n_est))
                est_matched <- c(est_matched, est_vals)
                fossil_levels <- c(fossil_levels, rep(fossil_names[i], n_est))
            }
        }
        result[[rate_name]] <- list(
            true = true_matched,
            estimated = est_matched,
            fossil_level = fossil_levels
        )
    }
    return(result)
}


matched_sample_50 <- create.matched.data(sum_var_change_true_50, clean_sample_50)
matched_no_ace_50 <- create.matched.data(sum_var_change_true_50, clean_no_ace_50)




# Calculate differences for sample method
matched_sample_50$slow$diffs  <- matched_sample_50$slow$estimated - matched_sample_50$slow$true
matched_sample_50$med$diffs   <- matched_sample_50$med$estimated - matched_sample_50$med$true
matched_sample_50$fast$diffs  <- matched_sample_50$fast$estimated - matched_sample_50$fast$true

# Calculate differences for no_ace method
matched_no_ace_50$slow$diffs  <- matched_no_ace_50$slow$estimated - matched_no_ace_50$slow$true
matched_no_ace_50$med$diffs   <- matched_no_ace_50$med$estimated - matched_no_ace_50$med$true
matched_no_ace_50$fast$diffs  <- matched_no_ace_50$fast$estimated - matched_no_ace_50$fast$true




sample_data_50 <- data.frame(
  diffs = c(matched_sample_50$slow$diffs, matched_sample_50$med$diffs, matched_sample_50$fast$diffs),
  fossil_level = c(matched_sample_50$slow$fossil_level, matched_sample_50$med$fossil_level, matched_sample_50$fast$fossil_level),
  rate = rep(c("slow", "med", "fast"), 
             c(length(matched_sample_50$slow$diffs), length(matched_sample_50$med$diffs), length(matched_sample_50$fast$diffs))),
  method = "Sample ACE"
)

# Add no_ace data
no_ace_data_50 <- data.frame(
  diffs = c(matched_no_ace_50$slow$diffs, matched_no_ace_50$med$diffs, matched_no_ace_50$fast$diffs),
  fossil_level = c(matched_no_ace_50$slow$fossil_level, matched_no_ace_50$med$fossil_level, matched_no_ace_50$fast$fossil_level),
  rate = rep(c("slow", "med", "fast"), 
             c(length(matched_no_ace_50$slow$diffs), length(matched_no_ace_50$med$diffs), length(matched_no_ace_50$fast$diffs))),
  method = "No ACE"
)

combined_data_50 <- rbind(sample_data_50, no_ace_data_50)


# Set the order for fossil levels (low to high preservation, then all)
level_order <- c("living", "fossil_low", "fossil_med", "fossil_high", "all")

# Convert fossil_level to factor with specified order in your combined_data
combined_data_50$fossil_level <- factor(combined_data_50$fossil_level, levels = level_order)

# Now your ridge plot will display in the correct order
ggplot(combined_data_50, aes(x = diffs, y = interaction(fossil_level, method), fill = method)) +
  geom_density_ridges(alpha = 0.7, scale = 1.2) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 1) +
  scale_fill_viridis_d(option = "plasma", end = 0.8) +
  facet_wrap(~rate, scales = "free_x") +
  theme_ridges() +
  theme(legend.position = "bottom") +
  labs(title = "Ridge Plot: Distribution of Estimation Errors",
       subtitle = "Higher peaks = more concentrated errors",
       x = "Estimation Error (Estimated - True)",
       y = "Fossil Level × Method",
       fill = "Method")

# library(viridis)
# fossil_colors <- viridis(5, option = "plasma", direction = -1)
# names(fossil_colors) <- c("all", "fossil_high", "fossil_med", "fossil_low", "living")

# plot(
#   c(matched_sample$slow$true, matched_sample$med$true, matched_sample$fast$true),
#   c(matched_sample$slow$estimated, matched_sample$med$estimated, matched_sample$fast$estimated),
#   xlab = "True Disparity Change",
#   ylab = "Est Disparity Change",
#   main = "Disparity Change: No majority Rule vs True",
#   pch = 16,
#   col = fossil_colors[
#     c(matched_sample$slow$fossil_level, matched_sample$med$fossil_level, matched_sample$fast$fossil_level)
#   ],
#   cex = 1.7,
#   ylim = c(-1, 1)
# )

# # Add 1:1 reference line
# abline(0, 1, col = "red", lty = 2, lwd = 2)

# plot(c(matched_no_ace$slow$true, matched_no_ace$med$true, matched_no_ace$fast$true),
#   c(matched_no_ace$slow$estimated, matched_no_ace$med$estimated, matched_no_ace$fast$estimated),
#     xlab = "True Disparity Change",
#     ylab = "Estimated Disparity Change",
#     main = "Disparity Change: No ace vs True",
#     pch = 16, 
#     col = fossil_colors[
#     c(matched_no_ace$slow$fossil_level, matched_no_ace$med$fossil_level, matched_no_ace$fast$fossil_level)
#   ],
#     cex = 1.7,
#     ylim = c(-1, 1))

# abline(0, 1, col = "red", lty = 2, lwd = 2)


```

## test of how many are within 10% of true


```{R}

# sum_var_change_true_ranges <- lapply(sum_var_change_true, lapply, function (x){
#     range <- x * 0.1
#     return(c((x- range), (x+range)))
# })

# coverage_no_majority <- Map(function(sample_rate, true_rate){
#     Map(function(sample_rep, true_rep){
#         true_vals <- true_rep
#         lapply(sample_rep, function(fossil){
#             sapply(fossil, function(v){
#                 is_within <- v >= min(true_vals) && v <= max(true_vals)
#                 return(is_within)
#             })
#         })
#     }, sample_rate, true_rate)
# }, clean_sample, sum_var_change_true_ranges)

# coverage_no_ace <- Map(function(sample_rate, true_rate){
#     Map(function(sample_rep, true_rep){
#         true_vals <- true_rep
#         lapply(sample_rep, function(fossil){
#             is_within <- fossil >= min(true_vals) && fossil <= max(true_vals)
#             return(is_within)
#         })
#     }, sample_rate, true_rate)
# }, clean_no_ace, sum_var_change_true_ranges)


```




# 100t 
## sum var

```{R}


rate_names <- c("slow", "med", "fast")

# Load sample data (keep as is)
sum_var_change_sample_100 <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_sample_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)

sum_var_change_no_ace_100 <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_no_ace_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)



# Load true data (keep as is)
sum_var_change_true_100 <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8140859_true_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)


# Calculate median on filtered sample data
# sum_var_change_sample_median <- lapply(sum_var_change_sample, function(rate_data) {
#     lapply(rate_data, function(replicate_data) {
#         # Apply median directly to each fossil level
#         result <- lapply(replicate_data, function(fossil_level_values) {
#             values <- unlist(fossil_level_values, use.names = FALSE)
#             return(median(values, na.rm = TRUE))
#         })
#         return(result)  
#     })
# })
sum_var_change_sample_unlist_100 <- lapply(sum_var_change_sample_100, function(rate_data) {
    lapply(rate_data, function(replicate_data) {
        # Apply median directly to each fossil level
        result <- lapply(replicate_data, function(fossil_level_values) {
            values <- unlist(fossil_level_values, use.names = FALSE)
            return(values)
        })
        return(result)  
    })
})

clean_sample_100 <- lapply(sum_var_change_sample_unlist_100, lapply, function(x){
    x[!sapply(x, is.null)]  
})

clean_no_ace_100 <- lapply(sum_var_change_no_ace_100, lapply, function(x){
    x[!sapply(x, is.null)]  
})




# Test the match
matched_sample_100 <- create.matched.data(sum_var_change_true_100, clean_sample_100)
matched_no_ace_100 <- create.matched.data(sum_var_change_true_100, clean_no_ace_100)




# Calculate differences for sample method
matched_sample_100$slow$diffs  <- matched_sample_100$slow$estimated - matched_sample_100$slow$true
matched_sample_100$med$diffs   <- matched_sample_100$med$estimated - matched_sample_100$med$true
matched_sample_100$fast$diffs  <- matched_sample_100$fast$estimated - matched_sample_100$fast$true

# Calculate differences for no_ace method
matched_no_ace_100$slow$diffs  <- matched_no_ace_100$slow$estimated - matched_no_ace_100$slow$true
matched_no_ace_100$med$diffs   <- matched_no_ace_100$med$estimated - matched_no_ace_100$med$true
matched_no_ace_100$fast$diffs  <- matched_no_ace_100$fast$estimated - matched_no_ace_100$fast$true



# matched_sample_100$slow$diffs  <- matched_sample$slow$estimated - matched_sample$slow$true
# matched_sample$med$diffs  <- matched_sample$med$estimated - matched_sample$med$true
# matched_sample$fast$diffs  <- matched_sample$fast$estimated - matched_sample$fast$true


# matched_no_ace <- create.matched.data(sum_var_change_true, clean_no_ace)
# matched_no_ace$slow$diffs  <- matched_no_ace$slow$estimated - matched_no_ace$slow$true
# matched_no_ace$med$diffs  <- matched_no_ace$med$estimated - matched_no_ace$med$true
# matched_no_ace$fast$diffs  <- matched_no_ace$fast$estimated - matched_no_ace$fast$true
# all_ace <- c(matched_sample$estimated, matched_no_ace$estimated)


sample_data_100 <- data.frame(
  diffs = c(matched_sample_100$slow$diffs, matched_sample_100$med$diffs, matched_sample_100$fast$diffs),
  fossil_level = c(matched_sample_100$slow$fossil_level, matched_sample_100$med$fossil_level, matched_sample_100$fast$fossil_level),
  rate = rep(c("slow", "med", "fast"), 
             c(length(matched_sample_100$slow$diffs), length(matched_sample_100$med$diffs), length(matched_sample_100$fast$diffs))),
  method = "Sample ACE"
)

# Add no_ace data
no_ace_data_100 <- data.frame(
  diffs = c(matched_no_ace_100$slow$diffs, matched_no_ace_100$med$diffs, matched_no_ace_100$fast$diffs),
  fossil_level = c(matched_no_ace_100$slow$fossil_level, matched_no_ace_100$med$fossil_level, matched_no_ace_100$fast$fossil_level),
  rate = rep(c("slow", "med", "fast"), 
             c(length(matched_no_ace_100$slow$diffs), length(matched_no_ace_100$med$diffs), length(matched_no_ace_100$fast$diffs))),
  method = "No ACE"
)

combined_data_100 <- rbind(sample_data_100, no_ace_data_100)


# Set the order for fossil levels (low to high preservation, then all)
level_order <- c("living", "fossil_low", "fossil_med", "fossil_high", "all")

# Convert fossil_level to factor with specified order in your combined_data
combined_data_100$fossil_level <- factor(combined_data_100$fossil_level, levels = level_order)

# Now your ridge plot will display in the correct order
ggplot(combined_data_100, aes(x = diffs, y = interaction(fossil_level, method), fill = method)) +
  geom_density_ridges(alpha = 0.7, scale = 1.2) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 1) +
  scale_fill_viridis_d(option = "plasma", end = 0.8) +
  facet_wrap(~rate, scales = "free_x") +
  theme_ridges() +
  theme(legend.position = "bottom") +
  labs(title = "Ridge Plot: Distribution of Estimation Errors",
       subtitle = "Higher peaks = more concentrated errors",
       x = "Estimation Error (Estimated - True)",
       y = "Fossil Level × Method",
       fill = "Method")


library(viridis)
fossil_colors <- viridis(5, option = "plasma", direction = -1)
names(fossil_colors) <- c("all", "fossil_high", "fossil_med", "fossil_low", "living")

level_order <- c("living", "fossil_low", "fossil_med", "fossil_high", "all")

# # Combine all data
# all_true <- c(matched_sample$slow$true, matched_sample$med$true, matched_sample$fast$true)
# all_est <- c(matched_sample$slow$estimated, matched_sample$med$estimated, matched_sample$fast$estimated)
# all_fossils <- c(matched_sample$slow$fossil_level, matched_sample$med$fossil_level, matched_sample$fast$fossil_level)

# plot(all_true, all_est,
#      xlab = "True Disparity Change",
#      ylab = "Est Disparity Change", 
#      main = "Disparity Change: No majority Rule vs True",
#      type = "n", 
#      ylim = c(-0.6, 0.6))

# for(level in level_order) {
#   indices <- all_fossils == level
#   points(all_true[indices], all_est[indices],
#          pch = 20, col = fossil_colors[level], cex = 1.7)
# }

# abline(0, 1, col = "red", lty = 2, lwd = 2)
# legend("topleft", 
#        legend = level_order,  # Legend in same order as plotting
#        col = fossil_colors[level_order], 
#        pch = 20, cex = 1.2, title = "Fossil Level")


# plot(c(matched_no_ace$slow$true, matched_no_ace$med$true, matched_no_ace$fast$true),
#   c(matched_no_ace$slow$estimated, matched_no_ace$med$estimated, matched_no_ace$fast$estimated),
#     xlab = "True Disparity Change",
#     ylab = "Estimated Disparity Change",
#     main = "Disparity Change: No ace vs True",
#     pch = 20, 
#     col = fossil_colors[
#     c(matched_no_ace$slow$fossil_level, matched_no_ace$med$fossil_level, matched_no_ace$fast$fossil_level)
#   ],
#     cex = 1.7,
#     ylim = c(-0.6, 0.6))
# abline(0, 1, col = "red", lty = 2, lwd = 2)

```

## Combined 50-100 t

```{R}


matched_sample <- Map(function(rate_50, rate_100){
    Map(function(treatment_50, treatment_100){
        c(treatment_50, treatment_100)
    }, rate_50, rate_100)
}, matched_sample_50, matched_sample_100)

matched_no_ace <- Map(function(rate_50, rate_100){
    Map(function(treatment_50, treatment_100){
        c(treatment_50, treatment_100)
    }, rate_50, rate_100)
}, matched_no_ace_50, matched_no_ace_100)

# Calculate differences for sample method
matched_sample$slow$diffs  <- matched_sample$slow$estimated - matched_sample$slow$true
matched_sample$med$diffs   <- matched_sample$med$estimated - matched_sample$med$true
matched_sample$fast$diffs  <- matched_sample$fast$estimated - matched_sample$fast$true

# Calculate differences for no_ace method
matched_no_ace$slow$diffs  <- matched_no_ace$slow$estimated - matched_no_ace$slow$true
matched_no_ace$med$diffs   <- matched_no_ace$med$estimated - matched_no_ace$med$true
matched_no_ace$fast$diffs  <- matched_no_ace$fast$estimated - matched_no_ace$fast$true
plot_data <- data.frame(
  diffs = c(matched_sample$slow$diffs, matched_sample$med$diffs, matched_sample$fast$diffs),
  fossil_level = c(matched_sample$slow$fossil_level, matched_sample$med$fossil_level, matched_sample$fast$fossil_level),
  rate = rep(c("slow", "med", "fast"), 
             c(length(matched_sample$slow$diffs), length(matched_sample$med$diffs), length(matched_sample$fast$diffs))),
  method = "Sample ACE"
)

# Add no_ace data
no_ace_data <- data.frame(
  diffs = c(matched_no_ace$slow$diffs, matched_no_ace$med$diffs, matched_no_ace$fast$diffs),
  fossil_level = c(matched_no_ace$slow$fossil_level, matched_no_ace$med$fossil_level, matched_no_ace$fast$fossil_level),
  rate = rep(c("slow", "med", "fast"), 
             c(length(matched_no_ace$slow$diffs), length(matched_no_ace$med$diffs), length(matched_no_ace$fast$diffs))),
  method = "No ACE"
)

combined_data <- rbind(plot_data, no_ace_data)


# Set the order for fossil levels (low to high preservation, then all)
level_order <- c("living", "fossil_low", "fossil_med", "fossil_high", "all")

# Convert fossil_level to factor with specified order in your combined_data
combined_data$fossil_level <- factor(combined_data$fossil_level, levels = level_order)

# Now your ridge plot will display in the correct order
ggplot(combined_data, aes(x = diffs, y = interaction(fossil_level, method), fill = method)) +
  geom_density_ridges(alpha = 0.7, scale = 1.2) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 1) +
  scale_fill_viridis_d(option = "plasma", end = 0.8) +
  facet_wrap(~rate, scales = "free_x") +
  theme_ridges() +
  theme(legend.position = "bottom") +
  labs(title = "Distribution of error in change in disparity",
       x = "Estimation Error (Estimated - True)",
       fill = "Method")





# create.lm.model <- function(matched, fossil) {
#     indices <- matched$fossil_level == fossil
#     estimated <- matched$estimated[indices]
#     true <- matched$true[indices]
#     model <- lm(estimated ~ true)
#     # Return a named list so you can use $all, $fossil_high, etc.
#     result <- list(estimated = estimated, true = true, model = model)
#     names(result) <- c("estimated", "true", "model")
#     return(result)
# }


# matched_sample_fossil_high <- create.lm.model(matched_sample$med, "fossil_low")
# matched_no_ace_fossil_high <- create.lm.model(matched_no_ace$med, "fossil_low")


# summary(matched_sample_fossil_high$model)
# summary(matched_no_ace_fossil_high$model)



# all_indices <- matched_sample$fossil_level == "all"
# matched_sample$all$estimated <- matched_sample$estimated[all_indices]
# matched_sample$all$true <- matched_sample$true[all_indices]

# all_indices <- matched_no_ace$fossil_level == "all"
# matched_no_ace$all$estimated <- matched_no_ace$estimated[all_indices]
# matched_no_ace$all$true <- matched_no_ace$true[all_indices]





# all_model_sample <- lm(matched_sample$all$estimated ~ matched_sample$all$true)
# all_model_no_ace <- lm(matched_no_ace$all$estimated ~ matched_no_ace$all$true)




```

## sum quant
```{r}


# 100t sum var

rate_names <- c("slow", "med", "fast")

# Load sample data (keep as is)
sum_quant_change_sample <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8100947_sample_sum_quant_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)

sum_quant_change_no_ace <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8100947_no_ace_sum_quant_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)



# Load true data (keep as is)
sum_quant_change_true <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8100947_true_sum_quant_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)


# Calculate median on filtered sample data
sum_quant_change_sample_median <- lapply(sum_quant_change_sample, function(rate_data) {
    lapply(rate_data, function(replicate_data) {
        # Apply median directly to each fossil level
        result <- lapply(replicate_data, function(fossil_level_values) {
            values <- unlist(fossil_level_values, use.names = FALSE)
            return(median(values, na.rm = TRUE))
        })
        return(result)  # This ensures we return the fossil level list directly
    })
})

clean_sample <- lapply(sum_quant_change_sample_median, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})

clean_no_ace <- lapply(sum_quant_change_no_ace, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})




# Test the match
matched_sample <- create.matched.data(sum_quant_change_true, clean_sample)
matched_no_ace <- create.matched.data(sum_quant_change_true, clean_no_ace)

all_ace <- c(matched_sample$estimated, matched_no_ace$estimated)



library(viridis)
fossil_colors <- viridis(5, option = "plasma", direction = -1)
names(fossil_colors) <- c("all", "fossil_high", "fossil_med", "fossil_low", "living")

plot(c(matched_sample$slow$true), c(matched_sample$slow$estimated),
    xlab = "True Disparity Change",
    ylab = "Est Disparity Change",
    main = "Disparity Change: No majority Rule vs True",
    pch = 16, 
    col = fossil_colors[matched_sample$fast$fossil_level],
    cex = 1.7
)

# Add 1:1 reference line
abline(0, 1, col = "red", lty = 2, lwd = 2)

plot(matched_no_ace$slow$true, matched_no_ace$slow$estimated,
    xlab = "True Disparity Change",
    ylab = "Estimated Disparity Change",
    main = "Disparity Change: No ace vs True",
    pch = 16, 
    col = fossil_colors[matched_no_ace$slow$fossil_level],
    cex = 1.7
)

abline(0, 1, col = "red", lty = 2, lwd = 2)




```




## Displacement
```{r}


# 100t sum var

rate_names <- c("slow", "med", "fast")

# Load sample data (keep as is)
displacement_change_sample <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8100947_sample_displacement_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)

displacement_change_no_ace <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8100947_no_ace_displacement_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)



# Load true data (keep as is)
displacement_change_true <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8100947_true_displacement_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)


# Calculate median on filtered sample data
displacement_change_sample_median <- lapply(displacement_change_sample, function(rate_data) {
    lapply(rate_data, function(replicate_data) {
        # Apply median directly to each fossil level
        result <- lapply(replicate_data, function(fossil_level_values) {
            values <- unlist(fossil_level_values, use.names = FALSE)
            return(median(values, na.rm = TRUE))
        })
        return(result)  # This ensures we return the fossil level list directly
    })
})

clean_sample <- lapply(displacement_change_sample_median, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})

clean_no_ace <- lapply(displacement_change_no_ace, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})




# Test the match
matched_sample <- create.matched.data(displacement_change_true, clean_sample)
matched_no_ace <- create.matched.data(displacement_change_true, clean_no_ace)

all_ace <- c(matched_sample$estimated, matched_no_ace$estimated)



library(viridis)
fossil_colors <- viridis(5, option = "plasma", direction = -1)
names(fossil_colors) <- c("all", "fossil_high", "fossil_med", "fossil_low", "living")

plot(c(matched_sample$slow$true, matched_sample$med$true, matched_sample$fast$true), c(matched_sample$slow$estimated, matched_sample$med$estimated, matched_sample$fast$estimated),
    xlab = "True Disparity Change",
    ylab = "Est Disparity Change",
    main = "Disparity Change: No majority Rule vs True",
    pch = 16, 
    col = fossil_colors[matched_sample$fast$fossil_level],
    cex = 1.7
)

# Add 1:1 reference line
abline(0, 1, col = "red", lty = 2, lwd = 2)

plot(c(matched_no_ace$slow$true, matched_no_ace$med$true, matched_no_ace$fast$true), c(matched_no_ace$slow$estimated, matched_no_ace$med$estimated, matched_no_ace$fast$estimated),
    xlab = "True Disparity Change",
    ylab = "Estimated Disparity Change",
    main = "Disparity Change: No ace vs True",
    pch = 16, 
    col = fossil_colors[c(matched_no_ace$slow$fossil_level, matched_no_ace$med$fossil_level, matched_no_ace$fast$fossil_level)],
    cex = 1.7
)

abline(0, 1, col = "red", lty = 2, lwd = 2)
legend("topleft", 
       legend = names(fossil_colors),
       col = fossil_colors,
       pch = 16,
       cex = 1.2,
       title = "Fossil Level")



```

# 50t with higher turnover and 

```{R}


# 50t
```{R}
rate_names <- c("slow", "med", "fast")

# Load sample data (keep as is)
sum_var_change_sample <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8115692_sample_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)

sum_var_change_strict <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8115692_strict_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)

sum_var_change_no_ace <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8115692_no_ace_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)



# Load true data (keep as is)
sum_var_change_true <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8115692_true_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)




# Calculate median on filtered sample data
sum_var_change_sample_median <- lapply(sum_var_change_sample, function(rate_data) {
    lapply(rate_data, function(replicate_data) {
        # Apply median directly to each fossil level
        result <- lapply(replicate_data, function(fossil_level_values) {
            values <- unlist(fossil_level_values, use.names = FALSE)
            return(median(values, na.rm = TRUE))
        })
        return(result)  # This ensures we return the fossil level list directly
    })
})

sum_var_change_sample_unlist <- lapply(sum_var_change_sample, function(rate_data) {
    lapply(rate_data, function(replicate_data) {
        # Apply median directly to each fossil level
        result <- lapply(replicate_data, function(fossil_level_values) {
            values <- unlist(fossil_level_values, use.names = FALSE)
            return(values)
        })
        return(result)  # This ensures we return the fossil level list directly
    })
})



clean_sample <- lapply(sum_var_change_sample_unlist, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})

clean_no_ace <- lapply(sum_var_change_no_ace, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})



clean_strict <- lapply(sum_var_change_strict, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})
# 
#Use filtered true data
# sum_var_change_true <- sum_var_change_true_filtered

# Your function should now work perfectly
create.matched.data <- function(true, estimated) {
    result <- list()
    for(rate_name in names(estimated)) {
        true_matched <- c()
        est_matched <- c()
        fossil_levels <- c()
        for(rep in seq_along(estimated[[rate_name]])) {
            true_val <- true[[rate_name]][[rep]]
            est_rep_val <- estimated[[rate_name]][[rep]]
            fossil_names <- names(est_rep_val)
            # Loop over fossil levels
            for(i in seq_along(est_rep_val)) {
                est_vals <- est_rep_val[[i]]
                n_est <- length(est_vals)
                true_matched <- c(true_matched, rep(true_val, n_est))
                est_matched <- c(est_matched, est_vals)
                fossil_levels <- c(fossil_levels, rep(fossil_names[i], n_est))
            }
        }
        result[[rate_name]] <- list(
            true = true_matched,
            estimated = est_matched,
            fossil_level = fossil_levels
        )
    }
    return(result)
}
matched_sample <- create.matched.data(sum_var_change_true, clean_sample)
matched_no_ace <- create.matched.data(sum_var_change_true, clean_no_ace)
matched_strict <- create.matched.data(sum_var_change_true, clean_strict)

all_ace <- c(matched_sample$estimated, matched_no_ace$estimated)



library(viridis)
fossil_colors <- viridis(5, option = "plasma", direction = -1)
names(fossil_colors) <- c("all", "fossil_high", "fossil_med", "fossil_low", "living")

plot(
  c(matched_sample$slow$true, matched_sample$med$true, matched_sample$fast$true),
  c(matched_sample$slow$estimated, matched_sample$med$estimated, matched_sample$fast$estimated),
  xlab = "True Disparity Change",
  ylab = "Est Disparity Change",
  main = "Disparity Change: No majority Rule vs True",
  pch = 20,
  col = fossil_colors[
    c(matched_sample$slow$fossil_level, matched_sample$med$fossil_level, matched_sample$fast$fossil_level)
  ],
  cex = 1.7,
  ylim = c(-0.6, 0.6)
)

# Add 1:1 reference line
abline(0, 1, col = "red", lty = 2, lwd = 2)

plot(c(matched_no_ace$slow$true, matched_no_ace$med$true, matched_no_ace$fast$true),
  c(matched_no_ace$slow$estimated, matched_no_ace$med$estimated, matched_no_ace$fast$estimated),
    xlab = "True Disparity Change",
    ylab = "Estimated Disparity Change",
    main = "Disparity Change: No ace vs True",
    pch = 20, 
    col = fossil_colors[
    c(matched_no_ace$slow$fossil_level, matched_no_ace$med$fossil_level, matched_no_ace$fast$fossil_level)
  ],
    cex = 1.7,
      ylim = c(-0.6, 0.6)
)

abline(0, 1, col = "red", lty = 2, lwd = 2)



plot(c(matched_strict$slow$true, matched_strict$med$true, matched_strict$fast$true),
  c(matched_strict$slow$estimated, matched_strict$med$estimated, matched_strict$fast$estimated),
    xlab = "True Disparity Change",
    ylab = "Estimated Disparity Change",
    main = "Disparity Change: Strict majority vs True",
    pch = 20, 
    col = fossil_colors[
    c(matched_strict$slow$fossil_level)
  ],
    cex = 1.7,
      ylim = c(-0.6, 0.6)
)

abline(0, 1, col = "red", lty = 2, lwd = 2)


```



# 50t with higher turnover and rm axes and 10my subsets

```{R}


# 50t
```{R}
rate_names <- c("slow", "med", "fast")

# Load sample data (keep as is)
sum_var_change_sample <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8120770_sample_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)

sum_var_change_strict <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8120770_strict_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)

sum_var_change_no_ace <- setNames(lapply(rate_names, function(rate) {
    # Load all replicates first
    all_reps <- lapply(1:100, function(rep) {
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8120770_no_ace_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
    })
}), rate_names
)



# Load true data (keep as is)
sum_var_change_true <- setNames(lapply(rate_names, function(rate) {
    lapply(1:100, function(rep) {
        tryCatch({
            file_path <- sprintf("../Data/cluster/randomExtinction/disp/8120770_true_sum_var_change_%03d.rds", rep)
            data <- readRDS(file_path)
            return(data[[rate]])
        }, warning = function(w) {
            NULL
        })
    })
}), rate_names)




# Calculate median on filtered sample data
sum_var_change_sample_median <- lapply(sum_var_change_sample, function(rate_data) {
    lapply(rate_data, function(replicate_data) {
        # Apply median directly to each fossil level
        result <- lapply(replicate_data, function(fossil_level_values) {
            values <- unlist(fossil_level_values, use.names = FALSE)
            return(median(values, na.rm = TRUE))
        })
        return(result)  # This ensures we return the fossil level list directly
    })
})

sum_var_change_sample_unlist <- lapply(sample_sum_var_change, function(rate_data) {
    lapply(rate_data, function(replicate_data) {
        # Apply median directly to each fossil level
        result <- lapply(replicate_data, function(fossil_level_values) {
            values <- unlist(fossil_level_values, use.names = FALSE)
            return(values)
        })
        return(result)  # This ensures we return the fossil level list directly
    })
})



clean_sample <- lapply(sum_var_change_sample_unlist, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})

clean_no_ace <- lapply(no_ace_sum_var_change, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})



clean_strict <- lapply(strict_sum_var_change, lapply, function(x){
    x[!sapply(x, is.null)]  # Keep only non-NULL elements
})
# 
#Use filtered true data
# sum_var_change_true <- sum_var_change_true_filtered

# Your function should now work perfectly
create.matched.data <- function(true, estimated) {
    result <- list()
    for(rate_name in names(estimated)) {
        true_matched <- c()
        est_matched <- c()
        fossil_levels <- c()
        for(rep in seq_along(estimated[[rate_name]])) {
            true_val <- true[[rate_name]][[rep]]
            est_rep_val <- estimated[[rate_name]][[rep]]
            fossil_names <- names(est_rep_val)
            # Loop over fossil levels
            for(i in seq_along(est_rep_val)) {
                est_vals <- est_rep_val[[i]]
                n_est <- length(est_vals)
                true_matched <- c(true_matched, rep(true_val, n_est))
                est_matched <- c(est_matched, est_vals)
                fossil_levels <- c(fossil_levels, rep(fossil_names[i], n_est))
            }
        }
        result[[rate_name]] <- list(
            true = true_matched,
            estimated = est_matched,
            fossil_level = fossil_levels
        )
    }
    return(result)
}
matched_sample <- create.matched.data(true_sum_var_change, clean_sample)
matched_no_ace <- create.matched.data(true_sum_var_change, clean_no_ace)
matched_strict <- create.matched.data(true_sum_var_change, clean_strict)

all_ace <- c(matched_sample$estimated, matched_no_ace$estimated)



library(viridis)
fossil_colors <- viridis(5, option = "plasma", direction = -1)
names(fossil_colors) <- c("all", "fossil_high", "fossil_med", "fossil_low", "living")

plot(
  c(matched_sample$slow$true, matched_sample$med$true, matched_sample$fast$true),
  c(matched_sample$slow$estimated, matched_sample$med$estimated, matched_sample$fast$estimated),
  xlab = "True Disparity Change",
  ylab = "Est Disparity Change",
  main = "Disparity Change: No majority Rule vs True",
  pch = 20,
  col = fossil_colors[
    c(matched_sample$slow$fossil_level, matched_sample$med$fossil_level, matched_sample$fast$fossil_level)
  ],
  cex = 1.7,
  ylim = c(-0.6, 0.6)
)

# Add 1:1 reference line
abline(0, 1, col = "red", lty = 2, lwd = 2)

plot(c(matched_no_ace$slow$true, matched_no_ace$med$true, matched_no_ace$fast$true),
  c(matched_no_ace$slow$estimated, matched_no_ace$med$estimated, matched_no_ace$fast$estimated),
    xlab = "True Disparity Change",
    ylab = "Estimated Disparity Change",
    main = "Disparity Change: No ace vs True",
    pch = 20, 
    col = fossil_colors[
    c(matched_no_ace$slow$fossil_level, matched_no_ace$med$fossil_level, matched_no_ace$fast$fossil_level)
  ],
    cex = 1.7,
      ylim = c(-0.6, 0.6)
)

abline(0, 1, col = "red", lty = 2, lwd = 2)



plot(c(matched_strict$slow$true, matched_strict$med$true, matched_strict$fast$true),
  c(matched_strict$slow$estimated, matched_strict$med$estimated, matched_strict$fast$estimated),
    xlab = "True Disparity Change",
    ylab = "Estimated Disparity Change",
    main = "Disparity Change: Strict majority vs True",
    pch = 20, 
    col = fossil_colors[
    c(matched_strict$slow$fossil_level)
  ],
    cex = 1.7,
      ylim = c(-0.6, 0.6)
)

abline(0, 1, col = "red", lty = 2, lwd = 2)


```